---
title: "Beta-diversity benchmark — Brito et al. (IL 2016; body site)"
subtitle: "PERMDISP (`vegan::betadisper`) distances; stool vs oral cavity"
author: "Mangiola Laboratory"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    fig-width: 10
    fig-height: 8
    theme: cosmo
execute:
  warning: false
  message: false
  fig.path: "britoil_2016_paper_figures/"
---

> **Memory requirement:** Rendering this report can require about **40 GB of RAM**.

## Overview

This report generates **paper-ready comparison figures** comparing multiple distance constructions for PERMDISP / `betadisper`:

- **PERMDISP distance inputs**: Bray–Curtis, Aitchison, robust Aitchison, Jaccard, Hellinger

**Simulation settings**: This report uses **Setting 5 from damping_optimization.qmd** (Moderate OD — Treatment Coding — Equal Baseline): σ ≈ 0.011, effect size = 2 for asymmetric DA, effect size = 1 for symmetric DA. The sweep plots (Figure 3) use equal baseline and treatment coding to match this setting.

We intentionally keep the analysis narrow and figure-focused.

## Method comparison

```{r method-comparison-table}
#| echo: false
#| tbl-cap: "Key properties of distance constructions for PERMDISP / betadisper"

library(knitr)
library(kableExtra)

# Define method categories
methods <- c(
  "PERMDISP distance inputs (Bray–Curtis, Aitchison, etc.)"
)

comparison_df <- data.frame(
  Method = methods,
  `Sensitive to dispersion` = c("Yes"),
  `Sensitive to composition` = c("Yes"),
  `Aware of mean–variance association` = c("No"),
  `Residual-based` = c("No"),
  `Handles zeros` = c("Pseudocount (Aitchison)"),
  `π-aware variance` = c("No"),
  check.names = FALSE
)

# Create table with Nature-style formatting (thick top and bottom borders)
cat('<div style="border-top: 3px solid black; border-bottom: 3px solid black; padding: 10px 0; margin: 20px 0;">')
kable(comparison_df, 
      format = "html",
      align = c("l", "c", "c", "c", "c", "c", "c"),
      caption = "Key properties of distance constructions for PERMDISP / betadisper") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center",
    htmltable_class = "nature-style-table"
  ) %>%
  row_spec(0, bold = TRUE, align = "c", extra_css = "border-bottom: 1px solid black;") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2:7, width = "12em")
cat('</div>')
```

**Key properties:**

- **Sensitive to dispersion**: Detects true differential stochastic overdispersion when present.
- **Sensitive to composition**: Produces false positives when groups differ only in mean composition (undesirable).
- **Aware of mean–variance association**: Accounts for the relationship between mean abundance and variance (log(σ) = -k×μ + c).
- **Residual-based**: Uses residuals (observed - expected) rather than raw counts/proportions.
- **Handles zeros**: Strategy for zero-inflated data (pseudocount, filtering, etc.).
- **π-aware variance**: Uses second-order delta method that explicitly incorporates expected proportions in variance calculations.

## Setup

```{r setup}
#| include: false

# Global filtering threshold will be loaded from targets pipeline

library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(vegan)
library(parallelly)
library(patchwork)
library(forcats)
library(RColorBrewer)
library(tidybulk)
library(stringr)
library(ggside)
library(tidyHeatmap)
library(sccomp)
library(ggrepel)
library(lme4)

source("functions.R")

knitr::opts_chunk$set(
  fig.path = "britoil_2016_paper_figures/",
  dpi = 220
)

# --- Theme and formatting helpers ---
# Theme for multipanel figures (from https://gist.github.com/stemangiola/fc67b08101df7d550683a5100106561c)
theme_multipanel <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(linewidth = 0.25),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 0, b = 0, l = 0), size = 7),
    axis.title.x = element_text(margin = margin(t = 0, r = 0, b = 0, l = 0), size = 7),
    panel.spacing.x = unit(0.1, "lines"),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.key.size = unit(5, 'mm'),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    strip.clip = "off",  # Avoid text clipping for facets
    plot.title = element_text(size = 7),
    axis.line.x = element_line(linewidth = 0.25),
    axis.line.y = element_line(linewidth = 0.25),
    axis.ticks.x = element_line(linewidth = 0.25),
    axis.ticks.y = element_line(linewidth = 0.25)
  )

# Function to drop leading zero from proportions (0.xxx -> .xxx)
dropLeadingZero <- function(l) {
  stringr::str_replace(l, '0(?=\\.)', '')
}

# Helper function to format axis labels: exactly 2 significant digits
# Examples: 25, 2.1, 0.2, 0.05
format_axis_labels <- function(x) {
  # Format to exactly 2 significant digits
  formatted <- signif(x, 2)
  
  # Convert to character with appropriate precision
  result <- character(length(formatted))
  for (i in seq_along(formatted)) {
    val <- formatted[i]
    if (is.na(val)) {
      result[i] <- "NA"
    } else if (val == 0) {
      result[i] <- "0"
    } else {
      val_abs <- abs(val)
      if (val_abs >= 1) {
        # For values >= 1: if integer, show as integer; otherwise 1 decimal
      if (abs(val - round(val)) < 1e-6) {
        result[i] <- as.character(round(val))
      } else {
        result[i] <- sprintf("%.1f", val)
      }
    } else {
        # For values < 1: determine decimal places needed to show exactly 2 significant digits
        # After signif(x, 2), we have a value with 2 significant digits
        # We need to determine how many decimal places are needed to display both digits
        
        # Find the position of the first non-zero digit (most significant)
        first_digit_pos <- floor(log10(val_abs))
        # For 2 significant digits, we need to show digits at positions:
        # first_digit_pos and (first_digit_pos - 1)
        # Number of decimal places needed = -(first_digit_pos - 1) = -first_digit_pos + 1
        # But we need at least enough to show the second digit
        
        # More precise: count how many digits after decimal are needed
        # For 0.25: first digit at -1, second at -2, need 2 decimals
        # For 0.05: first digit at -2, second at -3, need 2 decimals  
        # For 0.2: first digit at -1, but 0.2 rounded to 2 sig digits is 0.20, need 2 decimals
        # Actually, signif(0.2, 2) = 0.2, which has 1 sig digit, so we pad to show 2
        
        # The safest approach: use enough decimals to show 2 significant digits
        # Maximum decimals needed = 2 (for values like 0.05)
        # But we want minimum needed, so:
        # If first digit is at position p, we need to show positions p and p-1
        # Decimal places = max(1, -(p-1)) = max(1, -p+1)
        
        # Actually simpler: for values < 1, we always need at least 1 decimal
        # and at most 2 decimals to show 2 significant digits
        # Let's use a fixed approach: try 2 decimals, then remove trailing zeros
        
        # Use 2 decimals max, then clean up trailing zeros
        result[i] <- sprintf("%.2f", val)
      }
    }
  }
  
  # Clean up trailing zeros after decimal point, but preserve standalone zeros
  # Remove trailing zeros: "0.250" -> "0.25", "0.200" -> "0.2", but "0" stays "0"
  result <- stringr::str_replace(result, '(\\.\\d*?)0+$', '\\1')
  # Remove trailing decimal point: "1." -> "1"
  result <- stringr::str_replace(result, '\\.$', '')
  result
}

# --- Paper figure helpers (ggplot + patchwork) ---
plot_betadisper_pcoa_and_boxplot <- function(
  bd,
  title_prefix = "",
  hull = TRUE,
  segments = TRUE,
  label = TRUE,
  cols = c(
    stool = RColorBrewer::brewer.pal(11, "PRGn")[10],
    oralcavity = RColorBrewer::brewer.pal(11, "PRGn")[2]
  ),
  p_value = NULL,
  p_se = NULL,
  n_perm = NULL,
  show_group_legend = TRUE,
  horizontal_box = FALSE
) {
  # Match vegan::plot.betadisper defaults:
  # - asp = 1
  # - hull = TRUE (convex hull)
  # - segments = TRUE (spider to centroid)
  # - label = TRUE (centroid labels)
  sites <- as.data.frame(vegan::scores(bd, display = "sites"))
  if (ncol(sites) < 2) stop("betadisper site scores have < 2 dimensions.")
  colnames(sites)[1:2] <- c("Axis1", "Axis2")
  sites$group <- factor(bd$group, levels = levels(bd$group))
  # Draw second factor level on top in layered geoms.
  sites <- sites %>% dplyr::arrange(.data$group)

  centroids <- as.data.frame(vegan::scores(bd, display = "centroids"))
  if (ncol(centroids) < 2) stop("betadisper centroid scores have < 2 dimensions.")
  colnames(centroids)[1:2] <- c("Axis1", "Axis2")
  centroids$group <- factor(rownames(centroids), levels = levels(bd$group))
  centroids <- centroids %>% dplyr::arrange(.data$group)

  gl <- levels(sites$group)
  shape_vals <- stats::setNames(rep_len(c(16L, 17L, 15L, 18L), length(gl)), gl)

  seg_df <- NULL
  if (segments) {
    seg_df <- sites %>%
      dplyr::left_join(centroids %>% dplyr::select(group, cx = Axis1, cy = Axis2), by = "group")
  }

  hull_df <- NULL
  if (hull) {
    hull_df <- sites %>%
      dplyr::group_by(group) %>%
      dplyr::group_modify(\(d, ...) {
        if (nrow(d) < 3) return(d[0, , drop = FALSE])
        idx <- grDevices::chull(d$Axis1, d$Axis2)
        d[idx, , drop = FALSE] %>% dplyr::slice(c(seq_len(n()), 1))
      }) %>%
      dplyr::ungroup() %>%
      dplyr::arrange(.data$group)
  }

  p_pcoa <- ggplot() +
    {if (segments) geom_segment(
      data = seg_df,
      aes(x = cx, y = cy, xend = Axis1, yend = Axis2, color = group),
      linewidth = 0.1,
      alpha = 0.5
    )} +
    {if (hull) geom_polygon(
      data = hull_df,
      aes(x = Axis1, y = Axis2, fill = group, group = group, color = group),
      alpha = 0.15,
      linewidth = 0.1
    )} +
    geom_point(
      data = sites,
      aes(x = Axis1, y = Axis2, color = group, shape = group),
      size = 0.1,
      alpha = 0.8,
      stroke = 0
    ) +
    geom_point(
      data = centroids,
      aes(x = Axis1, y = Axis2, color = group),
      shape = 16,
      size = 0.1,
      stroke = 0
    ) +
    scale_color_manual(
      values = cols,
      name = NULL,
      guide = if (show_group_legend) "legend" else "none"
    ) +
    scale_shape_manual(
      values = shape_vals,
      name = NULL,
      guide = if (show_group_legend) "legend" else "none"
    ) +
    scale_fill_manual(
      values = cols,
      name = NULL,
      guide = "none"  # Fill is only for polygon shading, not in legend
    ) +
    labs(
      x = "PCoA 1",
      y = "PCoA 2"
    ) +
    scale_x_continuous(labels = format_axis_labels) +
    scale_y_continuous(labels = format_axis_labels) +
    theme_multipanel +
    theme(aspect.ratio = NULL)  # Allow PCoA plot to resize based on patchwork width

  dist_df <- tibble::tibble(
    group = factor(bd$group, levels = gl),
    distance = unname(bd$distances)
  )

  fmt_p <- function(p) {
    if (!is.finite(p)) return("NA")
    if (p < 0.001) return("<0.001")
    sprintf("%.3f", p)
  }
  ann_label <- NULL
  if (!is.null(p_value)) {
    ann_label <- paste0("p = ", fmt_p(p_value))
    # MC SE and permutation count removed - described in figure caption
  }
  y_max <- max(dist_df$distance, na.rm = TRUE)
  y_annotate <- if (is.finite(y_max)) y_max else NA_real_
  x_min <- min(dist_df$distance, na.rm = TRUE)
  x_rng <- diff(range(dist_df$distance, na.rm = TRUE))
  x_annotate_left <- if (is.finite(x_min) && is.finite(x_rng)) x_min + 0.02 * x_rng else x_min

  if (horizontal_box) {
    # Vertical boxplot: group on x-axis, distance on y-axis (even when horizontal_box=TRUE, use vertical for consistency)
    p_box <- ggplot(dist_df, aes(x = group, y = distance, fill = group)) +
      geom_boxplot(width = 0.5, outlier.size = 0.1, alpha = 0.85, linewidth = 0.25, outlier.stroke = 0) +
      geom_jitter(width = 0.12, size = 1.0, alpha = 0.25, stroke = 0) +
      scale_fill_manual(values = cols, guide = "none") +
      labs(x = NULL, y = "Distance to centroid") +
      {if (!is.null(ann_label)) annotate(
        geom = "text",
        x = 1.5,
        y = Inf,
        label = ann_label,
        size = 3.0,
        fontface = "plain",
        vjust = 1.1
      )} +
      coord_cartesian(clip = "off") +
      theme_multipanel +
      theme(
        aspect.ratio = NULL  # Allow boxplot to resize based on patchwork width
      )
  } else {
    # Vertical boxplot: group on x-axis, distance on y-axis
    p_box <- ggplot(dist_df, aes(x = group, y = distance, fill = group)) +
      geom_boxplot(width = 0.5, outlier.size = 0.1, alpha = 0.85, linewidth = 0.25, outlier.stroke = 0) +
      geom_jitter(width = 0.12, size = 1.0, alpha = 0.25, stroke = 0) +
      scale_fill_manual(values = cols, guide = "none") +
      labs(x = NULL, y = "Distance to centroid") +
      {if (!is.null(ann_label)) annotate(
        geom = "text",
        x = 1.5,
        y = Inf,
        label = ann_label,
        size = 3.0,
        fontface = "plain",
        vjust = 1.1
      )} +
      coord_cartesian(clip = "off") +
      theme_multipanel +
      theme(
        aspect.ratio = NULL  # Allow boxplot to resize based on patchwork width
      )
  }

  list(pcoa = p_pcoa, box = p_box)
}

# Express centroid separation as a percentage of average group hull radius (PCoA 1-2).
# 100% means a centroid shift equal to the average hull radius.
compute_centroid_distance_pct_hull_radius <- function(bd) {
  sites <- as.data.frame(vegan::scores(bd, display = "sites"))
  if (ncol(sites) < 2) {
    return(list(
      centroid_dist_pcoa12 = NA_real_,
      mean_hull_radius = NA_real_,
      pct_of_mean_hull_radius = NA_real_
    ))
  }
  colnames(sites)[1:2] <- c("Axis1", "Axis2")
  sites$group <- bd$group

  centroids <- as.data.frame(vegan::scores(bd, display = "centroids"))
  if (nrow(centroids) != 2 || ncol(centroids) < 2) {
    return(list(
      centroid_dist_pcoa12 = NA_real_,
      mean_hull_radius = NA_real_,
      pct_of_mean_hull_radius = NA_real_
    ))
  }
  colnames(centroids)[1:2] <- c("Axis1", "Axis2")
  centroid_dist <- sqrt(sum((centroids[1, 1:2] - centroids[2, 1:2])^2))

  hull_radii <- sites %>%
    dplyr::group_by(group) %>%
    dplyr::group_map(\(d, ...) {
      if (nrow(d) < 3) return(NA_real_)
      cx <- mean(d$Axis1, na.rm = TRUE)
      cy <- mean(d$Axis2, na.rm = TRUE)
      radial_dist <- sqrt((d$Axis1 - cx)^2 + (d$Axis2 - cy)^2)
      max(radial_dist, na.rm = TRUE)
    }) %>%
    unlist(use.names = FALSE)

  mean_hull_radius <- mean(hull_radii, na.rm = TRUE)
  pct <- if (is.finite(mean_hull_radius) && mean_hull_radius > 0) {
    100 * centroid_dist / mean_hull_radius
  } else {
    NA_real_
  }

  list(
    centroid_dist_pcoa12 = centroid_dist,
    mean_hull_radius = mean_hull_radius,
    pct_of_mean_hull_radius = pct
  )
}

# Format analysis results for plotting - takes output from run_analysis_job(return_full = TRUE)
# This function formats analysis results into structures needed for plotting (stats, betadisper objects, distances)
format_analysis_results_for_plotting <- function(
  analysis_results,
  case_label,
  standard_distance_methods = c("bray", "aitchison", "robust.aitchison", "jaccard", "hellinger"),
  group_levels = c("stool", "oralcavity")
) {

  if (!is.list(analysis_results) || !"method_results" %in% names(analysis_results)) {
    stop("analysis_results must be output from run_analysis_job(return_full = TRUE)")
  }

  method_results <- analysis_results$method_results
  method_names <- names(method_results)
  
  # Validate that we have the expected methods
  if (!is.character(standard_distance_methods) || length(standard_distance_methods) < 1) {
    stop("standard_distance_methods must be a non-empty character vector.")
  }
  if (anyDuplicated(standard_distance_methods)) {
    stop("standard_distance_methods contains duplicates.")
  }

  # Format statistics
  stats <- purrr::imap_dfr(method_results, \(mr, mn) {
    tibble::tibble(
      case = case_label,
      method = mn,
      group1 = mr$group1,
      group2 = mr$group2,
      delta = mr$delta,
      p_value = mr$p_value,
      p_se = mr$p_se,
      n_perm = mr$n_perm
    )
  }) %>%
    dplyr::mutate(method = factor(.data$method, levels = method_names))

  # Format distances for boxplots
  distances_long <- purrr::imap_dfr(method_results, \(mr, mn) {
    tibble::tibble(
      sample_id = names(mr$distances),
      group = mr$group,
      distance = mr$distances,
      method = mn
    )
  }) %>%
    dplyr::mutate(
      method = factor(.data$method, levels = method_names),
      group = factor(.data$group, levels = group_levels)
    )

  # Extract betadisper objects
  bd_by_method <- purrr::imap(method_results, \(mr, mn) mr$bd)
  
  # Return formatted results
  list(
    stats = stats,
    bd_by_method = bd_by_method,
    bd = list(
      standard = method_results[[paste0("permdisp_", standard_distance_methods[[1]])]]$bd
    ),
    distances_long = distances_long
  )
}

plot_alpha_diversity_scatter <- function(sim_result, case_label = "Case") {
  # This is the same style used in the discovery report:
  # `print(alpha_1$plot)` from `run_alpha_diversity_analysis()`.
  # BritoIL: two body sites (hardcoded for this QMD only).
  alpha <- run_alpha_diversity_analysis(
    sim_result,
    case_label = case_label,
    group_levels = c("stool", "oralcavity"),
    group_colors = c(
      stool = RColorBrewer::brewer.pal(11, "PRGn")[10],
      oralcavity = RColorBrewer::brewer.pal(11, "PRGn")[2]
    )
  )
  color_vals <- c(
    stool = RColorBrewer::brewer.pal(11, "PRGn")[10],
    oralcavity = RColorBrewer::brewer.pal(11, "PRGn")[2]
  )
  # Update plot to use theme_multipanel and remove titles/subtitles
  # Note: stat_ellipse linewidth is set in functions.R
  # Add marginal densities using ggside (fill will inherit from color scale if overridden)
  # Override point size to match boxplots (run_alpha_diversity_analysis uses size = 0.1)
  alpha$plot +
    labs(title = NULL, subtitle = NULL) +
    theme_multipanel +
    ggplot2::geom_point(ggplot2::aes(x = richness, y = pielou_evenness, color = group), 
                        size = 1.0, alpha = 0.6, stroke = 0,
                        position = ggplot2::position_jitter(width = 0.25, height = 0.003),
                        inherit.aes = FALSE, data = alpha$data) +
    ggside::geom_xsidedensity(aes(x = richness, fill = group), alpha = 0.5, linewidth = 0.25) +
    ggside::geom_ysidedensity(aes(y = pielou_evenness, fill = group), alpha = 0.5, linewidth = 0.25) +
    ggplot2::scale_fill_manual(values = color_vals, guide = "none") +
    ggplot2::theme(
      ggside.axis.ticks.x.bottom = ggplot2::element_line(linewidth = 0.25),
      ggside.axis.ticks.y.left = ggplot2::element_line(linewidth = 0.25),
      ggside.axis.line.x.bottom = ggplot2::element_line(linewidth = 0.25),
      ggside.axis.line.y.left = ggplot2::element_line(linewidth = 0.25),
      ggside.panel.scale.x = 0.15,
      ggside.panel.scale.y = 0.15
    )
}

# Whisker plot: 95% intervals of proportions per taxon and group,
# with taxa ordered by overall mean abundance (descending).
plot_mean_dispersion_scatter <- function(sim_result, case_label = "Case", cols = NULL) {
  if (is.null(cols)) {
    cols <- c(
      stool = RColorBrewer::brewer.pal(11, "PRGn")[10],
      oralcavity = RColorBrewer::brewer.pal(11, "PRGn")[2]
    )
  }

  d <- sim_result$count_long %>%
    dplyr::mutate(proportion = count / library_size) %>%
    dplyr::group_by(group, taxon_id) %>%
    dplyr::summarise(
      mean_prop = mean(proportion, na.rm = TRUE),
      p_low = stats::quantile(proportion, 0.025, na.rm = TRUE),
      p_high = stats::quantile(proportion, 0.975, na.rm = TRUE),
      .groups = "drop"
    )

  taxon_order <- d %>%
    dplyr::group_by(taxon_id) %>%
    dplyr::summarise(overall_mean = mean(mean_prop, na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(overall_mean)) %>%
    dplyr::pull(taxon_id)

  d <- d %>%
    dplyr::mutate(taxon_id = factor(taxon_id, levels = taxon_order))

  ggplot(d, aes(x = taxon_id, y = mean_prop, ymin = p_low, ymax = p_high, color = group)) +
    ggplot2::geom_linerange(position = ggplot2::position_dodge(width = 0.6), linewidth = 0.25, alpha = 0.8) +
    ggplot2::geom_point(position = ggplot2::position_dodge(width = 0.6), size = 0.8, stroke = 0) +
    ggplot2::scale_color_manual(values = cols, name = NULL) +
    ggplot2::scale_y_continuous(labels = function(x) dropLeadingZero(scales::percent_format(accuracy = 1)(x))) +
    ggplot2::labs(
      x = "Taxon (ordered by mean proportion)",
      y = "Proportion (95% interval)"
    ) +
    theme_multipanel +
    ggplot2::theme(
      axis.text.x = ggplot2::element_blank(),
      axis.ticks.x = ggplot2::element_blank()
    )
}

# Distance input helpers live in `_targets_BritoIL_2016.R` (analysis only).

# Method label functions
pretty_method_label <- function(method_key) {
  if (startsWith(method_key, "permdisp_")) {
    dm <- sub("^permdisp_", "", method_key)
    if (dm == "bray") return("Bray–Curtis")
    if (dm == "aitchison") return("Aitchison")
    if (dm == "robust.aitchison") return("robust Aitchison")
    if (dm == "jaccard") return("Jaccard")
    if (dm == "hellinger") return("Hellinger")
    return(dm)
  }
  method_key
}

pretty_method_label_roc <- function(method_key) {
  # Short labels for ROC plots
  if (startsWith(method_key, "permdisp_")) {
    dm <- sub("^permdisp_", "", method_key)
    if (dm == "bray") return("Bray–Curtis")
    if (dm == "aitchison") return("Aitchison")
    if (dm == "robust.aitchison") return("robust Aitchison")
    if (dm == "jaccard") return("Jaccard")
    if (dm == "hellinger") return("Hellinger")
    return(dm)
  }
  method_key
}

pretty_method_label_box <- function(method_key) {
  # Labels for boxplots (same as pretty_method_label)
  pretty_method_label(method_key)
}

# Define consistent color scheme for all methods (used across all plots)
roc_colors_all <- c(
  "Bray–Curtis" = RColorBrewer::brewer.pal(9, "Set1")[1],
  "Aitchison" = RColorBrewer::brewer.pal(9, "Set1")[2],
  "robust Aitchison" = RColorBrewer::brewer.pal(9, "Set1")[3],
  "Jaccard" = RColorBrewer::brewer.pal(9, "Set1")[4],
  "Hellinger" = RColorBrewer::brewer.pal(9, "Set1")[5]
)
method_levels <- names(roc_colors_all)

# Single source of truth for method levels/colors used in all sweeps and summaries.
coerce_method_levels <- function(x) {
  x_chr <- trimws(as.character(x))
  unknown <- setdiff(unique(x_chr), method_levels)
  if (length(unknown) > 0) {
    stop("Unknown method labels (no color mapping defined): ", paste(unknown, collapse = ", "))
  }
  factor(x_chr, levels = method_levels)
}

# format_analysis_results_for_plotting is defined in this QMD (plotting helper function)
```

```{r brito-sccomp-input}
#| cache: false

# BritoIL_2016 sccomp fit derived from curatedMetagenomicData.
# Factor of interest for this benchmark: body_site (stool vs oralcavity).
# Following mia/OMA-style preprocessing, we keep target body sites and apply
# prevalence+detection filtering before model fitting.
sccomp_file_path <- "BritoIL_2016_sccomp_results.rds"

suppressPackageStartupMessages({
  library(curatedMetagenomicData)
  library(TreeSummarizedExperiment)
  library(dplyr)
  library(tidyr)
  library(sccomp)
  library(tidySummarizedExperiment)
  library(magrittr)
})

tse <- curatedMetagenomicData(
  "BritoIL_2016.relative_abundance",
  dryrun = FALSE,
  counts = TRUE
) %$%
 `2021-03-31.BritoIL_2016.relative_abundance` |>
  dplyr::filter(body_site %in% c("stool", "oralcavity")) |>
  dplyr::mutate(body_site = factor(body_site, levels = c("stool", "oralcavity")))

# MIA-style feature filtering:
# - detection threshold: at least 10 counts to call a taxon "present"
# - prevalence threshold: present in at least 10% of samples

n_taxa_before <- nrow(tse)
tse <- mia::subsetByPrevalent(
  tse,
  detection = 10L,
  prevalence = 0.10,
  assay.type = "relative_abundance"
)
n_taxa_after <- nrow(tse)

cat("BritoIL_2016 sccomp input prepared.\n")
cat("Samples:", n_distinct(tse$.sample), "\n")
cat("Taxa:", n_distinct(tse$.feature), "\n")
cat("Taxa before/after prevalence filter:", n_taxa_before, "/", n_taxa_after, "\n")
print(
  tse |>
    dplyr::distinct(.sample, body_site) |>
    dplyr::count(body_site)
)
```

```{r fit-sccomp-brito-if-needed}
#| cache: false

if (!file.exists(sccomp_file_path)) {
  message("Running sccomp_estimate() because file is missing: ", sccomp_file_path)
  fit <- mia::meltSE(
    tse,
    assay.type = "relative_abundance",
    add.row = FALSE,
    add.col = TRUE
  ) |>

  sccomp::sccomp_estimate(
    formula_composition = ~ body_site + subject_id + age_category + gender + DNA_extraction_kit,
    formula_variability = ~ body_site,
    sample = "SampleID",
    cell_group = "FeatureID",
    abundance = "relative_abundance",
    inference_method = "hmc",
    cores = max(1, parallelly::availableCores() - 1), 
    bimodal_mean_variability_association = TRUE,
    verbose = TRUE, 
    max_sampling_iterations = 3000,
    sig_figs = 6, 
    portable = FALSE,

    # TEST
    exclude_priors = TRUE
  ) |>
    sccomp_test(contrasts = c("Intercept" = "`(Intercept)`", "body_siteoralcavity"))
  
  fit = fit |> 
    sccomp:::incorporate_parameters_into_sccomp_object(
    parameters_to_load = c( 
      "prec_intercept_1",
      "prec_slope_1",
      "prec_intercept_2",
      "prec_slope_2",
      "prec_sd",
      "mix_p",
      "beta",
      "alpha"
    ))
  
  saveRDS(fit, sccomp_file_path)
  message("Saved sccomp fit to: ", sccomp_file_path)
} else {
  message("Using existing sccomp fit: ", sccomp_file_path)
}

# Always load the fit object for downstream QC/targets chunks.
fit <- readRDS(sccomp_file_path)
```

## QC plots
```{r qc-plot}
fit |>
  sccomp_test(contrasts = "body_siteoralcavity") |>
  sccomp_plot_intervals_2D(factor = "body_site")
```


```{r targets-run}
#| eval: true

library(targets)
library(VGAM)

targets::tar_script({

library(targets)
library(tarchetypes)
library(crew)
library(crew.cluster)

# Local controller for Mac development
local_controller <- crew_controller_local(
  name = "local",
  workers = 8, #parallelly::availableCores(),
  seconds_idle = 30
)

# SLURM controller for HPC/Linux
# Helper (optional) to avoid repetition
new_elastic <- function(name, mem_gb, time_min, workers, crashes_max, backup = NULL) {
  crew_controller_slurm(
    name = name,
    workers = workers,
    crashes_max = crashes_max,
    seconds_idle = 30,
    options_cluster = crew_options_slurm(
      memory_gigabytes_required = mem_gb,
      cpus_per_task = 2,
      time_minutes = time_min
    ),
    backup = backup
  )
}

# Small → large, with fallbacks to the next size up
elastic_160 <- new_elastic("elastic_160", 160, 60 * 24, workers = 8,  crashes_max = 2)
elastic_80  <- new_elastic("elastic_80",   80,  60 * 4,  workers = 16, crashes_max = 1, backup = elastic_160)
elastic_40  <- new_elastic("elastic_40",   40,  60 * 4,  workers = 24, crashes_max = 1, backup = elastic_80)
elastic_20  <- new_elastic("elastic_20",   20,  60 * 4,  workers = 32, crashes_max = 1, backup = elastic_40)
elastic_10  <- new_elastic("elastic_10",   10,  60 * 4,  workers = 48, crashes_max = 1, backup = elastic_20)
elastic_5   <- new_elastic("elastic_5",     5, 60 * 4,  workers = 200, crashes_max = 6, backup = elastic_10)

# Group for targets (small → large)
slurm_controller <- crew_controller_group(
  elastic_5, elastic_10, elastic_20, elastic_40, elastic_80, elastic_160
)

# Select controller based on OS
controller <- if (Sys.info()["sysname"] == "Darwin") {
  local_controller  # macOS
} else {
  slurm_controller   # Linux/HPC
}

tar_option_set(
  memory = "transient",
  garbage_collection = 100,
  storage = "worker",
  retrieval = "worker",
  # error = "continue",
  workspace_on_error = TRUE,
  format = "rds",
  controller = controller,
  packages = c(
    "dplyr", "tidyr", "purrr", "tibble", "stringr",
    "ggplot2", "vegan", "parallelly",
    "sccomp", "patchwork"
  )
)

#!/usr/bin/env R

# Helper functions for {targets} pipeline.
# These mirror the parameter extraction logic in the Quarto report.

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(tibble)
  library(vegan)
})

source("functions.R")

utils::globalVariables(c(
  "parameter", "c_effect", "c_lower", "c_upper", "v_effect",
  "sample_id", "taxon_id", "count", "library_size", "mean_prop",
  "group", "n_da_taxa", "method", "p_value", "n_fp", "n_sims",
  "simulate_compositional_bb"
))

load_sccomp_result <- function(path) {
  readRDS(path)
}

list(
  tar_target(
    sccomp_result_file,
    "BritoIL_2016_sccomp_results.rds",
    format = "file"
  ),
  # Simulation/analysis helpers live in functions.R; track as a file target so changing the
  # sim algorithm invalidates downstream targets (source() alone is invisible to {targets}).
  tar_target(
    functions_r,
    "functions.R",
    format = "file"
  ),
  tar_target(
    sccomp_result,
    load_sccomp_result(sccomp_result_file)
  ),
  tar_target(
    sccomp_params,
    {
      functions_r
      # Dispersion for Brito sims: taxon intercepts from v_effect (log-sigma mapping), not prec_intercept.
      # DA/DA_sd sweeps use mean_dispersion_assoc_slope = 0; keep k_real = 0 so DO paths match.
      extract_sccomp_params_brito(sccomp_result)
    }
  ),
  # Number of simulation replications (set to small number for development)
  tar_target(
    n_reps_auc,
    ifelse(Sys.info()["sysname"] == "Darwin", 1L, 100L)  # Change this to 20 or higher for final runs
  ),
  tar_target(
    abundance_thresholds,
    c(0.0, 0.005, 0.001)  # 0.0 = unfiltered, 0.005 = 0.5%, 0.001 = 0.1% 
  ),
  tar_target(
    sim_params,
    build_simulation_params_brito(
      sccomp_params,
      n_reps_auc = n_reps_auc,
      group_levels = c("stool", "oralcavity")
    )
  ),
  # Setting DA_no_sd: Simulation targets (differential abundance, no sd on overdispersion)
  tar_target(
    sweep_setting_DA_no_sd_jobs,
    {
      jobs <- tidyr::crossing(seed = sim_params$rep_seeds_auc, n_da_taxa = 0:40)
      jobs$branch_index <- seq_len(nrow(jobs))
      jobs$target_name <- "sweep_setting_DA_no_sd_sim"
      jobs
    }, packages = c("VGAM")
  ),
  tar_target(
    sweep_setting_DA_no_sd_sim,
    run_simulation_job_brito(
      sweep_setting_DA_no_sd_jobs,
      dispersion_intercept_by_taxon = sim_params$intercept_disp_realistic_taxon,
      composition_intercept_by_taxon = sim_params$mu_inv_softmax_base_realistic,
      mean_dispersion_assoc_slope = 0,
      slope_effects_distribution = sccomp_params$slope_effects_significant,
      sd_log_overdispersion = 0,
      n_taxa = sim_params$n_taxa,
      n_samples = sim_params$n_samples,
      n_samples_per_group = sim_params$n_samples_per_group,
      library_size_mean = sim_params$library_size_mean,
      library_size_sd = sim_params$library_size_sd
    ) |> synthetic_sim_add_two_group_factor(c("stool", "oralcavity")),
    pattern = map(sweep_setting_DA_no_sd_jobs),
    iteration = "list"
  ),
  # Setting DA_no_sd: Analysis targets (with thresholds including unfiltered)
  # Create jobs grid: sim results × thresholds (0.0 = unfiltered, 0.005, 0.001)
  tar_target(
    sweep_setting_DA_no_sd_jobs_with_threshold,
    {
      sim_jobs <- sweep_setting_DA_no_sd_jobs
      jobs <- tidyr::crossing(
        seed = sim_jobs$seed,
        n_da_taxa = sim_jobs$n_da_taxa,
        threshold = abundance_thresholds
      )
      jobs$branch_index <- seq_len(nrow(jobs))
      jobs
    }
  ),
  tar_target(
    sweep_setting_DA_no_sd_sim_with_threshold,
    run_simulation_job_brito(
      sweep_setting_DA_no_sd_jobs_with_threshold,
      dispersion_intercept_by_taxon = sim_params$intercept_disp_realistic_taxon,
      composition_intercept_by_taxon = sim_params$mu_inv_softmax_base_realistic,
      mean_dispersion_assoc_slope = 0,
      slope_effects_distribution = sccomp_params$slope_effects_significant,
      sd_log_overdispersion = 0,
      n_taxa = sim_params$n_taxa,
      n_samples = sim_params$n_samples,
      n_samples_per_group = sim_params$n_samples_per_group,
      library_size_mean = sim_params$library_size_mean,
      library_size_sd = sim_params$library_size_sd
    ) |> synthetic_sim_add_two_group_factor(c("stool", "oralcavity")),
    pattern = map(sweep_setting_DA_no_sd_jobs_with_threshold),
    iteration = "list"
  ),
  tar_target(
    sweep_setting_DA_no_sd_res,
    {
      threshold <- sweep_setting_DA_no_sd_jobs_with_threshold$threshold
      
      # Apply filtering (threshold = 0.0 keeps all taxa, effectively unfiltered)
      sim_filtered <- sweep_setting_DA_no_sd_sim_with_threshold
      sim_filtered$count_long <- filter_by_abundance_cases(
        sweep_setting_DA_no_sd_sim_with_threshold$count_long,
        threshold = threshold
      )
      result <- run_analysis_job(sim_r = sim_filtered, return_full = TRUE)
      
      # Add threshold to the result for later filtering/plotting
      result$threshold <- threshold
      result$seed <- sweep_setting_DA_no_sd_sim_with_threshold$seed
      result$n_da_taxa <- sweep_setting_DA_no_sd_jobs_with_threshold$n_da_taxa
      result
    },
    pattern = map(sweep_setting_DA_no_sd_jobs_with_threshold, sweep_setting_DA_no_sd_sim_with_threshold),
    iteration = "list"
  ),
  tar_target(
    sweep_setting_DA_no_sd_fpr,
    summarise_fpr(sweep_setting_DA_no_sd_res)
  ),
  tar_target(
    sweep_setting_DA_no_sd_permanova_res,
    {
      threshold <- sweep_setting_DA_no_sd_jobs_with_threshold$threshold
      seed <- sweep_setting_DA_no_sd_sim_with_threshold$seed
      n_da_taxa <- sweep_setting_DA_no_sd_jobs_with_threshold$n_da_taxa

      run_permanova_job(
        sim_r = sweep_setting_DA_no_sd_sim_with_threshold,
        threshold = threshold,
        seed = seed,
        n_da_taxa = n_da_taxa,
        permutations = 199
      )
    },
    pattern = map(sweep_setting_DA_no_sd_jobs_with_threshold, sweep_setting_DA_no_sd_sim_with_threshold),
    iteration = "list"
  ),
  tar_target(
    sweep_setting_DA_no_sd_permanova_tpr,
    summarise_permanova_tpr(dplyr::bind_rows(sweep_setting_DA_no_sd_permanova_res))
  ),
  # Setting DA_sd: Simulation targets (differential abundance, with sd on overdispersion)
  tar_target(
    sweep_setting_DA_sd_jobs,
    {
      jobs <- tidyr::crossing(seed = sim_params$rep_seeds_auc, n_da_taxa = 0:40)
      jobs$branch_index <- seq_len(nrow(jobs))
      jobs$target_name <- "sweep_setting_DA_sd_sim"
      jobs
    }, packages = c("VGAM")
  ),
  tar_target(
    sweep_setting_DA_sd_sim,
    run_simulation_job_brito(
      sweep_setting_DA_sd_jobs,
      dispersion_intercept_by_taxon = sim_params$intercept_disp_realistic_taxon,
      composition_intercept_by_taxon = sim_params$mu_inv_softmax_base_realistic,
      mean_dispersion_assoc_slope = 0,
      slope_effects_distribution = sccomp_params$slope_effects_significant,
      sd_log_overdispersion = sim_params$sd_log_overdispersion_realistic,
      n_taxa = sim_params$n_taxa,
      n_samples = sim_params$n_samples,
      n_samples_per_group = sim_params$n_samples_per_group,
      library_size_mean = sim_params$library_size_mean,
      library_size_sd = sim_params$library_size_sd
    ) |> synthetic_sim_add_two_group_factor(c("stool", "oralcavity")),
    pattern = map(sweep_setting_DA_sd_jobs),
    iteration = "list"
  ),
  # Setting DA_sd: Analysis targets (with thresholds including unfiltered)
  tar_target(
    sweep_setting_DA_sd_jobs_with_threshold,
    {
      sim_jobs <- sweep_setting_DA_sd_jobs
      jobs <- tidyr::crossing(
        seed = sim_jobs$seed,
        n_da_taxa = sim_jobs$n_da_taxa,
        threshold = abundance_thresholds
      )
      jobs$branch_index <- seq_len(nrow(jobs))
      jobs
    }
  ),
  tar_target(
    sweep_setting_DA_sd_sim_with_threshold,
    run_simulation_job_brito(
      sweep_setting_DA_sd_jobs_with_threshold,
      dispersion_intercept_by_taxon = sim_params$intercept_disp_realistic_taxon,
      composition_intercept_by_taxon = sim_params$mu_inv_softmax_base_realistic,
      mean_dispersion_assoc_slope = 0,
      slope_effects_distribution = sccomp_params$slope_effects_significant,
      sd_log_overdispersion = sim_params$sd_log_overdispersion_realistic,
      n_taxa = sim_params$n_taxa,
      n_samples = sim_params$n_samples,
      n_samples_per_group = sim_params$n_samples_per_group,
      library_size_mean = sim_params$library_size_mean,
      library_size_sd = sim_params$library_size_sd
    ) |> synthetic_sim_add_two_group_factor(c("stool", "oralcavity")),
    pattern = map(sweep_setting_DA_sd_jobs_with_threshold),
    iteration = "list"
  ),
  tar_target(
    sweep_setting_DA_sd_res,
    {
      threshold <- sweep_setting_DA_sd_jobs_with_threshold$threshold

      # Apply filtering (threshold = 0.0 keeps all taxa, effectively unfiltered)
      sim_filtered <- sweep_setting_DA_sd_sim_with_threshold
      sim_filtered$count_long <- filter_by_abundance_cases(
        sweep_setting_DA_sd_sim_with_threshold$count_long,
        threshold = threshold
      )
      result <- run_analysis_job(sim_r = sim_filtered, return_full = TRUE)
      
      result$threshold <- threshold
      result$seed <- sweep_setting_DA_sd_sim_with_threshold$seed
      result$n_da_taxa <- sweep_setting_DA_sd_jobs_with_threshold$n_da_taxa
      result
    },
    pattern = map(sweep_setting_DA_sd_jobs_with_threshold, sweep_setting_DA_sd_sim_with_threshold),
    iteration = "list"
  ),
  tar_target(
    sweep_setting_DA_sd_fpr,
    summarise_fpr(sweep_setting_DA_sd_res)
  ),

  # Setting DO_no_sd: Define jobs (differential overdispersion, no sd, no DA)
  tar_target(
    sweep_setting_DO_no_sd_jobs,
    {
      set.seed(105)
      n_reps_roc <- sim_params$n_reps_auc
      rep_seeds <- 21000 + seq_len(n_reps_roc)
      delta_grid <- c(0, 0.05, 0.10, 0.15, 0.20, 0.25, 0.375, 0.5, 0.625, 0.75, 0.875, 1.0)
      jobs <- tidyr::crossing(seed = rep_seeds, delta = delta_grid, threshold = abundance_thresholds)
      jobs$branch_index <- seq_len(nrow(jobs))
      jobs
    }
  ),
  
  # Setting DO_no_sd: Run each job (simulation + analysis)
  tar_target(
    sweep_setting_DO_no_sd_res,
    {
      intercept_base <- -6
      perm_reps_roc <- 199
      methods <- c("bray", "aitchison", "robust.aitchison", "jaccard", "hellinger")
      
      seed <- sweep_setting_DO_no_sd_jobs$seed
      delta <- sweep_setting_DO_no_sd_jobs$delta
      threshold <- sweep_setting_DO_no_sd_jobs$threshold
      intercept_disp_cohort <- c(intercept_base, intercept_base + delta)
      intercept_disp_matrix <- outer(intercept_disp_cohort, rep(1, sim_params$n_taxa))

      sim_r <- simulate_case_realistic_taxa(
        slope_vector = rep(0, sim_params$n_taxa),
        mu_inv_softmax = sim_params$mu_inv_softmax_base_realistic,
        log_dispersion_assoc = 0,
        intercept_dispersion_matrix = intercept_disp_matrix,
        sd_log_overdispersion = 0,
        seed = seed,
        n_taxa = sim_params$n_taxa,
        n_samples_per_group = sim_params$n_samples_per_group,
        library_size_mean = sim_params$library_size_mean,
        library_size_sd = sim_params$library_size_sd
      ) |> synthetic_sim_add_two_group_factor(c("stool", "oralcavity"))

      # Apply filtering (threshold = 0.0 keeps all taxa, effectively unfiltered)
      sim_r$count_long <- filter_by_abundance_cases(sim_r$count_long, threshold = threshold)
      n_taxa_after_filter <- length(unique(sim_r$count_long$taxon_id))
      if (n_taxa_after_filter < 3) {
        return(tibble::tibble(
          seed = integer(0),
          delta = numeric(0),
          truth = logical(0),
          method = character(0),
          p_value = numeric(0),
          centroid_dist = numeric(0),
          centroid_sep_pcoa12 = numeric(0),
          mean_hull_radius = numeric(0),
          centroid_sep_pct_hull_radius = numeric(0)
        ))
      }

      inputs_std <- purrr::set_names(
        lapply(methods, \(dm) {
          build_standard_betadisper_inputs_paper(
            count_long = sim_r$count_long,
            sample_metadata = sim_r$sample_metadata,
            distance_method = dm,
            group_levels = c("stool", "oralcavity")
          )
        }),
        nm = paste0("permdisp_", methods)
      )

      purrr::imap_dfr(inputs_std, \(inp, method_key) {
        bd <- betadisper(inp$dist, inp$group)
        p <- permutest_betadisper_fixed(bd, permutations = perm_reps_roc)$p_value

        dist_df <- data.frame(distance = bd$distances, group = bd$group)
        group_means <- tapply(dist_df$distance, dist_df$group, mean)
        centroid_distance <- abs(group_means["oralcavity"] - group_means["stool"])

        # Centroid separation in PCoA 1-2, scaled by average group hull radius.
        sites <- as.data.frame(vegan::scores(bd, display = "sites"))
        centroids <- as.data.frame(vegan::scores(bd, display = "centroids"))
        centroid_sep_pcoa12 <- NA_real_
        mean_hull_radius <- NA_real_
        centroid_sep_pct_hull_radius <- NA_real_
        if (ncol(sites) >= 2 && nrow(centroids) == 2 && ncol(centroids) >= 2) {
          colnames(sites)[1:2] <- c("Axis1", "Axis2")
          sites$group <- bd$group
          colnames(centroids)[1:2] <- c("Axis1", "Axis2")

          centroid_sep_pcoa12 <- sqrt(sum((centroids[1, 1:2] - centroids[2, 1:2])^2))
          hull_radii <- sites %>%
            dplyr::group_by(group) %>%
            dplyr::summarise(
              hull_radius = {
                cx <- mean(Axis1, na.rm = TRUE)
                cy <- mean(Axis2, na.rm = TRUE)
                max(sqrt((Axis1 - cx)^2 + (Axis2 - cy)^2), na.rm = TRUE)
              },
              .groups = "drop"
            )
          mean_hull_radius <- mean(hull_radii$hull_radius, na.rm = TRUE)
          if (is.finite(mean_hull_radius) && mean_hull_radius > 0) {
            centroid_sep_pct_hull_radius <- 100 * centroid_sep_pcoa12 / mean_hull_radius
          }
        }

        tibble::tibble(
          seed = seed,
          delta = delta,
          threshold = threshold,
          truth = delta > 0,
          method = pretty_method_label_roc(method_key),
          p_value = p,
          centroid_dist = as.numeric(centroid_distance),
          centroid_sep_pcoa12 = as.numeric(centroid_sep_pcoa12),
          mean_hull_radius = as.numeric(mean_hull_radius),
          centroid_sep_pct_hull_radius = as.numeric(centroid_sep_pct_hull_radius)
        )
      })
    },
    pattern = map(sweep_setting_DO_no_sd_jobs)
  ),
  
  # Setting DO_no_sd: Aggregate results
  tar_target(
    sweep_setting_DO_no_sd_fpr,
    {
      dplyr::bind_rows(sweep_setting_DO_no_sd_res) %>%
        dplyr::mutate(score = -log10(pmax(p_value, 1e-300))) %>%
        dplyr::filter(method != "Arcsin resid.")
    }
  )
)


}, script = "_targets_BritoIL_2016.R", ask=FALSE)

# Run the targets pipeline
targets::tar_make(script = "_targets_BritoIL_2016.R", store = "_targets_BritoIL_2016")
```

```{r targets-load}
# Set up targets configuration and load basic parameters used throughout
# Individual chunks will load their specific targets as needed
targets_script <- file.path(getwd(), "_targets_BritoIL_2016.R")
if (!file.exists(targets_script)) {
  qmd_dir <- dirname(knitr::current_input(dir = TRUE))
  targets_script <- file.path(qmd_dir, "_targets_BritoIL_2016.R")
}
targets::tar_config_set(script = targets_script)
# Source helper functions from _targets_BritoIL_2016.R AFTER tar_script/tar_make.
source(targets_script)

# Load basic parameters that are used in many places
sccomp_params <- targets::tar_read("sccomp_params", store = "_targets_BritoIL_2016")
sim_params <- targets::tar_read("sim_params", store = "_targets_BritoIL_2016")
abundance_thresholds <- targets::tar_read("abundance_thresholds", store = "_targets_BritoIL_2016")
# Default threshold for regular plots (use 0.005)
abundance_threshold <- 0.005

# Extract commonly used parameters
n_samples_real <- sccomp_params$n_samples_real
n_taxa_real <- sccomp_params$n_taxa_real
k_real <- sccomp_params$k_real
k_sd_real <- sccomp_params$k_sd_real
c_real <- sccomp_params$c_real
prec_sd_real <- sccomp_params$prec_sd_real
intercept_effects <- sccomp_params$intercept_effects
slope_effects <- sccomp_params$slope_effects
slope_effects_significant <- sccomp_params$slope_effects_significant
v_intercept <- sccomp_params$v_intercept
v_slope <- sccomp_params$v_slope

n_taxa <- sim_params$n_taxa
n_samples_per_group <- sim_params$n_samples_per_group
n_groups <- sim_params$n_groups
n_samples <- sim_params$n_samples
group_levels <- sim_params$group_levels
library_size_mean <- sim_params$library_size_mean
library_size_sd <- sim_params$library_size_sd
mu_inv_softmax_base_realistic <- sim_params$mu_inv_softmax_base_realistic
intercept_disp_realistic <- sim_params$intercept_disp_realistic
sigma_realistic <- sim_params$sigma_realistic
slope_realistic <- sim_params$slope_realistic
n_reps_auc <- sim_params$n_reps_auc
perm_reps_auc <- sim_params$perm_reps_auc
rep_seeds_auc <- sim_params$rep_seeds_auc
```


## Shared simulation settings

Simulation parameters and helper functions are now defined in the targets pipeline
(`_targets_BritoIL_2016.R`) and loaded via the `targets-load` chunk.



# Data Properties

## Goal

Use **realistic simulation parameters extracted from BritoIL_2016 microbiome data** (sccomp fit) to evaluate the performance of various beta-diversity methods under conditions that mirror real microbiome datasets.

This approach ensures our simulation sweeps reflect:

1. **Real sample and taxa counts** from BritoIL_2016
2. **Real mean-overdispersion association (k)** estimated from the data
3. **Real variability around the association (σ)**
4. **Real baseline abundance distribution** (intercept effects)
5. **Real slope distribution** (group effects)

## Extract Parameters from BritoIL_2016 sccomp Fit

```{r load-sccomp-results, cache: false}
# Load the sccomp results
result <- readRDS(sccomp_file_path)

# Identify the non-intercept coefficient name dynamically
# (e.g. groupIBD, body_siteoralcavity, etc.).
group_parameter <- result |>
  dplyr::filter(.data$parameter != "Intercept") |>
  dplyr::distinct(.data$parameter) |>
  dplyr::pull(.data$parameter) |>
  (\(x) x[[1]])()

if (is.na(group_parameter) || is.null(group_parameter)) {
  stop("Could not identify non-intercept group parameter in sccomp result.")
}

# Extract fit object
fit <- attr(result, "fit")

# Get basic dataset dimensions from the result
# The sccomp result stores dimensions in model_input (a list)
model_input <- attr(result, "model_input")

# Extract N (samples) and M (taxa/cell groups) from model_input list
if (!is.null(model_input) && is.list(model_input)) {
  n_samples_real <- model_input$N
  n_taxa_real <- model_input$M
} else {
  # Fallback: use the cell_group column in result (unique taxa)
  n_taxa_real <- length(unique(result$cell_group))
  # Samples: from the sccomp print output
  n_samples_real <- 178
}

cat("=== BritoIL_2016 Dataset Characteristics ===\n")
cat("Number of samples:", n_samples_real, "\n")
cat("Number of taxa (genera):", n_taxa_real, "\n")
```

```{r extract-dispersion-params}
# Extract mean-overdispersion association coefficient (k)
# from sccomp precision parameterization (`prec_intercept_j` / `prec_slope_j`, or unindexed names).
precision_params <- extract_precision_trend_params(fit)
prec_intercept <- precision_params$prec_intercept
prec_slope <- precision_params$prec_slope

# Direct mapping to simulator parameters:
# k = prec_slope (the simulator formula has -k, so negative prec_slope gives positive effect)
k_real <- prec_slope  # ≈ -0.64
k_sd_real <- precision_params$k_sd_real

# c = -prec_intercept (convert from precision to overdispersion intercept)
c_real <- -prec_intercept  # ≈ -5.16

cat("\n=== Precision Parameters from sccomp ===\n")
cat("Precision intercept:", round(prec_intercept, 4), "\n")
cat("Precision slope (", precision_params$slope_parameter, "): ", round(prec_slope, 4), "\n", sep = "")

cat("\n=== Converted to Overdispersion (sigma) Parameters ===\n")
cat("k (log_dispersion_assoc = prec_slope):", round(k_real, 4), "\n")
cat("c (intercept_dispersion = -prec_intercept):", round(c_real, 4), "\n")
cat("SD of k:", round(k_sd_real, 4), "\n")

cat("\n=== Interpretation ===\n")
cat("With k =", round(k_real, 2), "and formula log(σ) = -k*log(μ) + c:\n")
cat("  -k =", round(-k_real, 2), ", so higher mean → higher σ → MORE overdispersion ✓\n")

# Extract precision SD (variability around the mean-od relationship)
prec_sd_summary <- fit$summary("prec_sd")
prec_sd_real <- prec_sd_summary$mean[1]

cat("\n=== Precision SD (variability around k) ===\n")
cat("Estimated prec_sd:", round(prec_sd_real, 4), "\n")
cat("SD:", round(prec_sd_summary$sd[1], 4), "\n")
```

```{r extract-baseline-params}
# Extract baseline abundance distribution (intercept effects for composition)
intercept_effects <- result %>%
  filter(parameter == "Intercept") %>%
  pull(c_effect)

cat("\n=== Baseline Abundance Distribution (Intercept c_effect) ===\n")
cat("Mean:", round(mean(intercept_effects), 4), "\n")
cat("SD:", round(sd(intercept_effects), 4), "\n")
cat("Range: [", round(min(intercept_effects), 4), ",", 
    round(max(intercept_effects), 4), "]\n")

# Plot intercept distribution
intercept_mean <- mean(intercept_effects)
p_intercept <- ggplot(data.frame(c_effect = intercept_effects), aes(x = c_effect)) +
  geom_density(fill = "lightgrey", alpha = 1, color = "black", linewidth = 0.3) +
  geom_vline(xintercept = intercept_mean, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_rug() +
  labs(x = "inv. softmax proportion", y = "Density") +
  scale_x_continuous(labels = format_axis_labels) +
  scale_y_continuous(labels = format_axis_labels) +
  theme_multipanel
```

```{r extract-slope-params}
# Extract slope (group effect) distribution - ALL taxa
slope_effects <- result %>%
  filter(parameter == group_parameter) %>%
  pull(c_effect)

# Extract SIGNIFICANT slope effects (credible interval excludes 0)
# sccomp results have c_lower and c_upper columns for credible intervals
slope_data <- result %>%
  filter(parameter == group_parameter)

# Check if credible interval columns exist
if ("c_lower" %in% names(slope_data) && "c_upper" %in% names(slope_data)) {
  # Significant = credible interval does not include 0
  slope_effects_significant <- slope_data %>%
    filter(c_lower > 0 | c_upper < 0) %>%
    pull(c_effect)
} else {
  # Fallback: use effect size threshold (|effect| > 0.5 as "significant")
  slope_effects_significant <- slope_data %>%
    filter(abs(c_effect) > 0.5) %>%
    pull(c_effect)
}

cat("\n=== Group Effect Distribution (Slope c_effect for IBD) ===\n")
cat("All taxa:\n")
cat("  Mean:", round(mean(slope_effects), 4), "\n")
cat("  SD:", round(sd(slope_effects), 4), "\n")
cat("  Range: [", round(min(slope_effects), 4), ",", 
    round(max(slope_effects), 4), "]\n")
cat("  N taxa:", length(slope_effects), "\n")

cat("\nSignificant taxa only (CI excludes 0 or |effect| > 0.5):\n")
cat("  N significant:", length(slope_effects_significant), "\n")
if (length(slope_effects_significant) > 0) {
  cat("  Mean:", round(mean(slope_effects_significant), 4), "\n")
  cat("  SD:", round(sd(slope_effects_significant), 4), "\n")
  cat("  Range: [", round(min(slope_effects_significant), 4), ",", 
      round(max(slope_effects_significant), 4), "]\n")
}

# Plot slope distribution - all taxa
slope_mean <- mean(slope_effects)
p_slope <- ggplot(data.frame(c_effect = slope_effects), aes(x = c_effect)) +
  geom_density(fill = "lightgrey", alpha = 1, color = "black", linewidth = 0.3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
  geom_vline(xintercept = slope_mean, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_rug(alpha = 0.5) +
  labs(x = "inv. softmax proportion", y = "Density") +
  scale_x_continuous(labels = format_axis_labels) +
  scale_y_continuous(labels = format_axis_labels) +
  theme_multipanel
```

```{r extract-variability-params}
# Extract variability parameters (v_effect for intercept and slope)
# These are the per-taxon variability effects in sccomp
v_intercept <- result %>%
  filter(parameter == "Intercept") %>%
  pull(v_effect)

v_slope <- result %>%
  filter(parameter == group_parameter) %>%
  pull(v_effect)

cat("\n=== Variability Parameters (v_effect) ===\n")
cat("Intercept v_effect - Mean:", round(mean(v_intercept), 4), 
    ", SD:", round(sd(v_intercept), 4), "\n")
cat("Slope v_effect - Mean:", round(mean(v_slope), 4), 
    ", SD:", round(sd(v_slope), 4), "\n")

# Use c_real from precision conversion (c = -prec_intercept)
# This is the proper intercept for our simulator's overdispersion parameterization
intercept_dispersion_real <- c_real  # From extract-dispersion-params: c_real = -prec_intercept ≈ -5.16
sigma_real <- exp(intercept_dispersion_real)

cat("\n=== Simulation Intercept Dispersion (from precision conversion) ===\n")
cat("intercept_dispersion (c = -prec_intercept):", round(intercept_dispersion_real, 4), "\n")
cat("sigma (exp(intercept_dispersion)):", format(sigma_real, scientific = TRUE, digits = 3), "\n")
cat("\nNote: This is very low overdispersion (σ ≈ 0.006), consistent with high precision.\n")
```

```{r additional-parameter-plots}
#| eval: !expr Sys.info()[["sysname"]] != "Darwin"
# CmdStan draw CSVs are not available locally on macOS (absolute paths from Linux cluster).

# 1. Intercept c_effect in softmax scale (proportions)
# Convert intercept effects to proportions using inv_softmax
# We need to convert all intercept effects together to get proper proportions
intercept_softmax <- inv_softmax(intercept_effects)

arcsin_sqrt_trans_prop <- scales::trans_new(
  name = "arcsin_sqrt",
  transform = function(x) asin(sqrt(pmax(0, pmin(1, x)))),
  inverse = function(y) sin(y)^2,
  breaks = scales::extended_breaks(),
  domain = c(0, 1)
)

cat("\n=== Intercept c_effect in Softmax Scale (Proportions) ===\n")
cat("Mean:", round(mean(intercept_softmax), 4), "\n")
cat("SD:", round(sd(intercept_softmax), 4), "\n")
cat("Range: [", round(min(intercept_softmax), 4), ",", 
    round(max(intercept_softmax), 4), "]\n")

intercept_softmax_mean <- mean(intercept_softmax)
p_intercept_softmax <- ggplot(data.frame(proportion = intercept_softmax), aes(x = proportion)) +
  geom_density(fill = "lightblue", alpha = 1, color = "black", linewidth = 0.3) +
  geom_vline(xintercept = intercept_softmax_mean, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_rug() +
  labs(x = "Proportion", y = "Density") +
  scale_x_continuous(labels = format_axis_labels) +
  scale_y_continuous(labels = format_axis_labels) +
  theme_multipanel

# 2. Effect in proportion fold change using sccomp_proportional_fold_change
prop_fold_change_data <- sccomp_proportional_fold_change(
  result,
  formula_composition = ~ body_site,
  from = "stool",
  to = "oralcavity"
)

prop_fold_change <- prop_fold_change_data[["proportion_fold_change"]]
if (is.null(prop_fold_change)) {
  pfc_cols <- names(prop_fold_change_data)
  stop(
    "sccomp_proportional_fold_change did not return a 'proportion_fold_change' column. ",
    "Columns: ", paste(pfc_cols, collapse = ", ")
  )
}

feature_col <- dplyr::case_when(
  "cell_group" %in% names(prop_fold_change_data) ~ "cell_group",
  ".feature" %in% names(prop_fold_change_data) ~ ".feature",
  "FeatureID" %in% names(prop_fold_change_data) ~ "FeatureID",
  TRUE ~ NA_character_
)
if (is.na(feature_col)) {
  stop(
    "Could not find a feature column in proportional fold-change output. ",
    "Columns: ", paste(names(prop_fold_change_data), collapse = ", ")
  )
}

prop_fold_change_tbl <- prop_fold_change_data %>%
  dplyr::transmute(cell_group = .data[[feature_col]], proportion_fold_change = .data$proportion_fold_change)

result_feature_col <- dplyr::case_when(
  "cell_group" %in% names(result) ~ "cell_group",
  ".feature" %in% names(result) ~ ".feature",
  "FeatureID" %in% names(result) ~ "FeatureID",
  TRUE ~ NA_character_
)
if (is.na(result_feature_col)) {
  stop(
    "Could not find a feature column in sccomp `result`. ",
    "Columns: ", paste(names(result), collapse = ", ")
  )
}

significant_taxa <- result %>%
  dplyr::filter(parameter == group_parameter) %>%
  dplyr::filter(c_lower > 0 | c_upper < 0) %>%
  dplyr::transmute(cell_group = .data[[result_feature_col]])

prop_fold_change_sig <- prop_fold_change_tbl %>%
  dplyr::inner_join(significant_taxa, by = "cell_group") %>%
  dplyr::mutate(abs_fold_change = abs(proportion_fold_change))

cat("\n=== Absolute Proportion Fold Change (significant body_site effects) ===\n")
cat("N significant taxa:", nrow(prop_fold_change_sig), "\n")
cat("Mean:", round(mean(prop_fold_change_sig$abs_fold_change, na.rm = TRUE), 4), "\n")
cat("SD:", round(sd(prop_fold_change_sig$abs_fold_change, na.rm = TRUE), 4), "\n")
cat("Range: [", round(min(prop_fold_change_sig$abs_fold_change, na.rm = TRUE), 4), ",", 
    round(max(prop_fold_change_sig$abs_fold_change, na.rm = TRUE), 4), "]\n")

prop_fold_change_mean <- mean(prop_fold_change_sig$abs_fold_change, na.rm = TRUE)
p_prop_fold_change <- ggplot(prop_fold_change_sig, aes(x = abs_fold_change)) +
  geom_density(fill = "lightblue", alpha = 1, color = "black", linewidth = 0.3) +
  geom_vline(xintercept = prop_fold_change_mean, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_rug() +
  labs(x = "Absolute proportional fold change", y = "Density") +
  scale_x_continuous(labels = format_axis_labels) +
  scale_y_continuous(labels = format_axis_labels) +
  theme_multipanel

# 3. Distribution of v_effect (variability) - split by intercept and slope
cat("\n=== Intercept v_effect Distribution ===\n")
cat("Mean:", round(mean(v_intercept), 4), "\n")
cat("SD:", round(sd(v_intercept), 4), "\n")
cat("Range: [", round(min(v_intercept), 4), ",", 
    round(max(v_intercept), 4), "]\n")

# Convert v_effect using simulator mapping (log(sigma) = v_effect for Brito targets).
log_overdispersion_intercept <- v_intercept
log_overdispersion_slope <- v_slope

cat("\n=== Log Overdispersion Intercept (converted from v_effect) ===\n")
cat("Mean:", round(mean(log_overdispersion_intercept), 4), "\n")
cat("SD:", round(sd(log_overdispersion_intercept), 4), "\n")
cat("Range: [", round(min(log_overdispersion_intercept), 4), ",", 
    round(max(log_overdispersion_intercept), 4), "]\n")

log_overdispersion_intercept_mean <- mean(log_overdispersion_intercept)
p_v_effect_intercept <- ggplot(data.frame(log_overdispersion = log_overdispersion_intercept), aes(x = log_overdispersion)) +
  geom_density(fill = "lightgrey", alpha = 1, color = "black", linewidth = 0.3) +
  geom_vline(xintercept = log_overdispersion_intercept_mean, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_rug(alpha = 0.5) +
  labs(x = "log overdispersion", y = "Density") +
  scale_x_continuous(labels = format_axis_labels) +
  scale_y_continuous(labels = format_axis_labels) +
  theme_multipanel

cat("\n=== Log Overdispersion Slope (converted from v_effect) ===\n")
cat("Mean:", round(mean(log_overdispersion_slope), 4), "\n")
cat("SD:", round(sd(log_overdispersion_slope), 4), "\n")
cat("Range: [", round(min(log_overdispersion_slope), 4), ",", 
    round(max(log_overdispersion_slope), 4), "]\n")

log_overdispersion_slope_mean <- mean(log_overdispersion_slope)
p_v_effect_slope <- ggplot(data.frame(log_overdispersion = log_overdispersion_slope), aes(x = log_overdispersion)) +
  geom_density(fill = "lightgrey", alpha = 1, color = "black", linewidth = 0.3) +
  geom_vline(xintercept = log_overdispersion_slope_mean, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_rug(alpha = 0.5) +
  labs(x = "log overdispersion", y = "Density") +
  scale_x_continuous(labels = format_axis_labels) +
  scale_y_continuous(labels = format_axis_labels) +
  theme_multipanel
```

```{r combined-parameter-plot, fig.width=16, fig.height=12}
#| eval: !expr Sys.info()[["sysname"]] != "Darwin"

# Combined parameter visualization
# One row: intercept c_ and proportion, slope c_ and proportional fold, then OD intercept and slope
# Remove y-axis labels, numbers, and ticks from all plots except the leftmost one
# Also reduce side margins to minimize spacing between panels
# Add titles on top of each distribution
# Add small margins between panels (leftmost has no left margin, rightmost has no right margin)
p_intercept_adj <- p_intercept + 
  ggtitle("Intercept c_effect") +
  theme(plot.margin = margin(0, 2, 0, 0, "pt"), plot.title = element_text(size = 8, hjust = 0.5))
p_intercept_softmax_no_y <- p_intercept_softmax + 
  ggtitle("Proportion") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = margin(0, 2, 0, 2, "pt"), plot.title = element_text(size = 8, hjust = 0.5))
p_slope_no_y <- p_slope + 
  ggtitle("Slope c_effect") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = margin(0, 2, 0, 2, "pt"), plot.title = element_text(size = 8, hjust = 0.5))
p_prop_fold_change_no_y <- p_prop_fold_change + 
  ggtitle("Proportional fold change") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = margin(0, 2, 0, 2, "pt"), plot.title = element_text(size = 8, hjust = 0.5))
p_v_effect_intercept_no_y <- p_v_effect_intercept + 
  ggtitle("OD intercept") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = margin(0, 2, 0, 2, "pt"), plot.title = element_text(size = 8, hjust = 0.5))
p_v_effect_slope_no_y <- p_v_effect_slope + 
  ggtitle("OD slope") +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        plot.margin = margin(0, 0, 0, 2, "pt"), plot.title = element_text(size = 8, hjust = 0.5))

p_combined <- p_intercept_adj | p_intercept_softmax_no_y | p_slope_no_y | p_prop_fold_change_no_y | p_v_effect_intercept_no_y | p_v_effect_slope_no_y
p_combined <- p_combined + 
  theme(legend.key.size = unit(0.2, 'cm'))

p_combined
```

```{r combined-parameter-plot-save-pdf}
#| echo: false
#| eval: !expr Sys.info()[["sysname"]] != "Darwin"
ggsave("Figure_1_BritoIL_2016.pdf", p_combined, width = 183, height = 183/4, units = "mm")
```

## Simulation Setup with Realistic Parameters

```{r simulation-params}
set.seed(123)

# Match sccomp fit: full taxon count (analyses use all taxa that pass abundance filters)
n_taxa <- n_taxa_real
n_samples_per_group <- floor(n_samples_real / 2)  # Split into 2 groups
n_groups <- 2
n_samples <- n_samples_per_group * n_groups
group_levels <- c("stool", "oralcavity")  # Matches simulation output (body_site)

library_size_mean <- 15125  # From HMP2 average total counts per sample
library_size_sd <- 5000

cat("=== Simulation Parameters ===\n")
cat("n_taxa:", n_taxa, "(from fitted model)\n")
cat("n_samples_per_group:", n_samples_per_group, "\n")
cat("Total n_samples:", n_samples, "\n")

# Use real baseline distribution - sample from fitted intercepts
# Scale to n_taxa and center
set.seed(123)
sampled_intercepts <- sample(intercept_effects, n_taxa, replace = TRUE)
mu_inv_softmax_base_realistic <- sampled_intercepts - mean(sampled_intercepts)

cat("\nRealistic baseline (sampled from HMP2 intercepts):\n")
cat("Range: [", round(min(mu_inv_softmax_base_realistic), 3), ",", 
    round(max(mu_inv_softmax_base_realistic), 3), "]\n")

# Use real k value for mean-overdispersion association
k_realistic <- k_real

# Use real intercept dispersion for overdispersion level
intercept_disp_realistic <- intercept_dispersion_real
sigma_realistic <- sigma_real

cat("\nRealistic overdispersion parameters:\n")
cat("k (mean-OD association):", round(k_realistic, 4), "\n")
cat("intercept_dispersion:", round(intercept_disp_realistic, 4), "\n")
cat("sigma:", round(sigma_realistic, 6), "\n")

# Create realistic slope vector by directly sampling from the real distribution
# This captures the natural pattern of group effects observed in HMP2
set.seed(123)
slope_realistic <- sample(slope_effects, n_taxa, replace = TRUE)
# Center to sum to zero (compositional constraint)
slope_realistic <- slope_realistic - mean(slope_realistic)

# Verify lengths
stopifnot(length(mu_inv_softmax_base_realistic) == n_taxa)
stopifnot(length(slope_realistic) == n_taxa)

cat("\nRealistic slope vector (sampled from HMP2 group effects):\n")
cat("Mean:", round(mean(slope_realistic), 4), "\n")
cat("SD:", round(sd(slope_realistic), 4), "\n")
cat("Range: [", round(min(slope_realistic), 3), ",", round(max(slope_realistic), 3), "]\n")

# Sweep parameters
n_reps_auc <- 20  # Number of replicates
perm_reps_auc <- 199
rep_seeds_auc <- 12000 + seq_len(n_reps_auc)
```

Design matrix helpers are now defined in `_targets_BritoIL_2016.R`.

Helper functions are now defined in `_targets_BritoIL_2016.R`.

```{r simulation-function}
# Simulation functions for sweeps now live in the targets pipeline (`_targets_BritoIL_2016.R`).
# Function to create composite plot for DA taxa sweep
plot_da_taxa_sweep_results <- function(
  fpr_data,
  setting_name,
  sigma_val,
  intercept_disp_val,
  x_limits = NULL,
  x_breaks = NULL
) {
  standard_methods_labels <- method_levels

  # Convert raw method keys to pretty labels, then restrict to the 5 standard methods
  fpr_data_with_labels <- fpr_data %>%
    dplyr::mutate(method_label = dplyr::case_when(
      method == "permdisp_bray" ~ "Bray–Curtis",
      method == "permdisp_aitchison" ~ "Aitchison",
      method == "permdisp_robust.aitchison" ~ "robust Aitchison",
      method == "permdisp_robust_aitchison" ~ "robust Aitchison",
      method == "permdisp_jaccard" ~ "Jaccard",
      method == "permdisp_hellinger" ~ "Hellinger",
      as.character(method) %in% standard_methods_labels ~ as.character(method),
      TRUE ~ NA_character_
    )) %>%
    dplyr::filter(!is.na(method_label)) %>%
    dplyr::mutate(method_label = coerce_method_levels(method_label))
  
  # Handle both uppercase FPR and lowercase fpr column names
  fpr_col <- if ("FPR" %in% names(fpr_data_with_labels)) "FPR" else "fpr"
  
  # Add threshold handling if threshold column exists
  has_threshold <- "threshold" %in% names(fpr_data_with_labels) && 
                   !all(is.na(fpr_data_with_labels$threshold))
  
  if (has_threshold) {
    fpr_data_with_labels <- fpr_data_with_labels %>%
      dplyr::mutate(
        threshold_label = dplyr::case_when(
          threshold == 0.0 ~ "unfiltered",
          threshold == 0.005 ~ "0.5%",
          threshold == 0.001 ~ "0.1%",
          TRUE ~ as.character(threshold)
        ),
        alpha_val = dplyr::case_when(
          threshold == 0.0 ~ 0.25,
          threshold == 0.005 ~ 1.0,
          threshold == 0.001 ~ 0.55,
          TRUE ~ 1.0
        )
      )
  } else {
    fpr_data_with_labels$alpha_val <- 1.0
  }
  
  p <- ggplot(fpr_data_with_labels, aes(x = n_da_taxa, y = !!sym(fpr_col), color = method_label, alpha = alpha_val)) +
    geom_hline(yintercept = 0.05, linetype = "dashed", color = "grey60", linewidth = 0.1) +
    geom_line(linewidth = 0.1, aes(group = if(has_threshold) interaction(method_label, threshold) else method_label)) +
    geom_point(size = 1.0, stroke = 0, aes(group = if(has_threshold) interaction(method_label, threshold) else method_label)) +
    scale_color_manual(values = roc_colors_all, breaks = standard_methods_labels, drop = FALSE, name = NULL) +
    scale_alpha_identity(guide = "none") +
    scale_x_continuous(
      breaks = if (is.null(x_breaks)) unique(fpr_data_with_labels$n_da_taxa) else x_breaks,
      limits = x_limits,
      labels = format_axis_labels
    ) +
    scale_y_continuous(labels = format_axis_labels, limits = c(0, 1)) +
    labs(
      x = "Number of DA taxa",
      y = "False positive rate"
    ) +
    theme_multipanel
  
  p
}
```

```{r plotting-function}
# Function to create PCoA plot with facet-grid-like styling
plot_pcoa_single_method <- function(analysis_result, method_key, n_da_label, 
                                        is_first_col = FALSE, is_first_row = FALSE, 
                                        is_last_row = FALSE, is_last_col = FALSE,
                                        method_label = NULL) {
  if (is.null(analysis_result) || is.null(analysis_result$bd_by_method)) return(NULL)
  bd <- analysis_result$bd_by_method[[method_key]]
  if (is.null(bd)) return(NULL)

  p_val <- analysis_result$stats %>%
    dplyr::filter(method == method_key) %>%
    dplyr::pull(p_value)
  if (length(p_val) == 0) p_val <- NA_real_

  if (is.null(method_label)) {
    method_label <- pretty_method_label_box(method_key)
  }

  gg <- plot_betadisper_pcoa_and_boxplot(
    bd,
    title_prefix = "",
    cols = cols_prgn_setting5,
    p_value = p_val,
    show_group_legend = FALSE,
    horizontal_box = FALSE
  )
  
  # Customize PCoA plot
  p <- gg$pcoa
  
  # Add column title (n DA) only on first row
  if (is_first_row) {
    p <- p + ggtitle(n_da_label) +
      theme(plot.title = element_text(size = 8, hjust = 0.5, face = "bold", margin = margin(b = 4)))
  } else {
    p <- p + theme(plot.title = element_blank())
  }
  
  # Add method label annotation on the left side (like a facet strip)
  if (is_first_col) {
    # Get actual data range from betadisper scores for proper positioning
    sites_scores <- as.data.frame(vegan::scores(bd, display = "sites"))
    if (ncol(sites_scores) >= 2) {
      x_range <- range(sites_scores[[1]], na.rm = TRUE)
      y_range <- range(sites_scores[[2]], na.rm = TRUE)
      x_center <- mean(x_range)
      y_center <- mean(y_range)
      # Position label to the left of the plot, centered vertically
      # Use a position that's clearly outside the data range
      label_x <- x_range[1] - 0.15 * diff(x_range)
    } else {
      # Fallback if we can't get scores
      label_x <- -Inf
      y_center <- 0
    }
    
    p <- p + 
      # Add label positioned in left margin area, centered vertically
      annotate(
        "text",
        x = label_x,
        y = y_center,
        label = method_label,
        angle = 90,
        hjust = 0.5,  # Center horizontally when rotated
        vjust = 0.5,
        size = 2.8,
        fontface = "bold"
      ) +
      # Adjust left margin to make room for rotated label
      theme(plot.margin = margin(l = 55, r = 0, t = 0, b = 0, "pt")) +
      coord_cartesian(clip = "off")
  } else {
    # No margins for non-first columns
    p <- p + theme(plot.margin = margin(0, 0, 0, 0, "pt"))
  }
  
  # Show y-axis title only on leftmost column
  if (!is_first_col) {
    p <- p + theme(axis.title.y = element_blank())
  }
  
  # Show x-axis title only on bottom row
  if (!is_last_row) {
    p <- p + theme(axis.title.x = element_blank())
  }
  
  p
}

# Function to create PCoA grid with facet-grid-like styling
plot_pcoa_grid_multiple_methods <- function(analysis_list, n_da_labels) {
  standard_distance_methods_pcoa <- c("bray", "aitchison", "robust.aitchison", "jaccard", "hellinger")
  methods <- paste0("permdisp_", standard_distance_methods_pcoa)
  n_methods <- length(methods)
  n_cols <- length(analysis_list)
  
  all_plots <- list()
  
  # Iterate by row (method) first, then column (n_da) to match byrow = TRUE layout
  for (method_idx in seq_along(methods)) {
    method <- methods[method_idx]
    method_label <- pretty_method_label_box(method)
    is_first_row <- (method_idx == 1)
    is_last_row <- (method_idx == n_methods)
    
    for (col_idx in seq_along(analysis_list)) {
      analysis_result <- analysis_list[[col_idx]]
      n_da_label <- n_da_labels[col_idx]
      is_first_col <- (col_idx == 1)
      is_last_col <- (col_idx == n_cols)
      
      p <- plot_pcoa_single_method(
        analysis_result = analysis_result,
        method_key = method,
        n_da_label = n_da_label,
        is_first_col = is_first_col,
        is_first_row = is_first_row,
        is_last_row = is_last_row,
        is_last_col = is_last_col,
        method_label = method_label
      )
      
      if (!is.null(p)) {
        all_plots[[length(all_plots) + 1]] <- p
      }
    }
  }
  
  # Arrange in grid: n_methods rows, n_cols columns
  # byrow = TRUE means fill row by row, which matches our iteration order
  patchwork::wrap_plots(all_plots, nrow = n_methods, ncol = n_cols, byrow = TRUE) +
    patchwork::plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
}
```

```{r sweep-selection-helpers}
get_sweep_index <- function(jobs, seed, n_da_taxa) {
  idx <- which(jobs$seed == seed & jobs$n_da_taxa == n_da_taxa)
  if (length(idx) != 1) {
    stop("Expected exactly one sweep job for seed=", seed, " n_da_taxa=", n_da_taxa)
  }
  idx
}

get_sweep_index_with_threshold <- function(jobs_with_threshold, seed, n_da_taxa, threshold) {
  idx <- which(jobs_with_threshold$seed == seed & 
               jobs_with_threshold$n_da_taxa == n_da_taxa & 
               jobs_with_threshold$threshold == threshold)
  if (length(idx) != 1) {
    stop("Expected exactly one sweep job for seed=", seed, " n_da_taxa=", n_da_taxa, " threshold=", threshold)
  }
  idx
}

get_sweep_analysis_with_threshold <- function(sweep_res, jobs_with_threshold, seed, n_da_taxa, threshold) {
  sweep_res[[get_sweep_index_with_threshold(jobs_with_threshold, seed, n_da_taxa, threshold)]]
}
```

# Simulation no-DA (No Differential Abundance)

This section tests method performance when there is **differential overdispersion but NO differential abundance**. This evaluates whether methods correctly detect differential dispersion without producing false positives from composition differences.

All simulations in this section use realistic BritoIL_2016-derived parameters (baseline abundances, fitted mean-OD association, realistic library sizes) with `sd_log_overdispersion = 0` (no random variation in overdispersion across taxa). The same simulation case is used throughout, with different filtering applied for different figures.

## Figure 2: Filtered Data (0.1% abundance threshold)

This figure shows method performance on filtered simulated data (taxa with mean abundance <0.1% removed) with differential overdispersion but no differential abundance.

```{r case-4.5-load-params}
# Load BritoIL_2016 parameters for Figure 2 (Setting DO_no_sd)
# Load the sccomp results
result_case4_5 <- readRDS(sccomp_file_path)

# Extract fit object
fit_case4_5 <- attr(result_case4_5, "fit")

# Get basic dataset dimensions from the result
model_input_case4_5 <- attr(result_case4_5, "model_input")

# Extract N (samples) and M (taxa/cell groups) from model_input list
if (!is.null(model_input_case4_5) && is.list(model_input_case4_5)) {
  n_samples_real_case4_5 <- model_input_case4_5$N
  n_taxa_real_case4_5 <- model_input_case4_5$M
} else {
  # Fallback: use the cell_group column in result (unique taxa)
  n_taxa_real_case4_5 <- length(unique(result_case4_5$cell_group))
  n_samples_real_case4_5 <- 178
}

# Extract baseline abundance distribution (intercept effects for composition)
intercept_effects_case4_5 <- result_case4_5 %>%
  filter(parameter == "Intercept") %>%
  pull(c_effect)

# Extract mean-overdispersion association coefficient (k)
precision_params_case4_5 <- extract_precision_trend_params(fit_case4_5)
prec_slope_case4_5 <- precision_params_case4_5$prec_slope
k_realistic_case4_5 <- prec_slope_case4_5  # k = prec_slope for simulator

rm(result_case4_5)
gc()
```

```{r figure-2-setup}
# Figure 2 is about DO (differential overdispersion), not DA
# Row 1 uses delta DO = 0.5 from Setting DO_no_sd
# Load required targets
sweep_setting_DO_no_sd_jobs <- targets::tar_read("sweep_setting_DO_no_sd_jobs", store = "_targets_BritoIL_2016")

# Allow running this chunk standalone in IDE sessions where setup may be skipped.
if (!exists("simulate_case_realistic_taxa")) {
  targets_script_local <- file.path(getwd(), "_targets_BritoIL_2016.R")
  if (!file.exists(targets_script_local)) {
    qmd_dir_local <- dirname(knitr::current_input(dir = TRUE))
    targets_script_local <- file.path(qmd_dir_local, "_targets_BritoIL_2016.R")
  }
  source(targets_script_local)
}

# Recreate the simulation for delta = 0.5 (matching the DO_no_sd sweep)
# This matches the simulation used in the DO_no_sd sweep
# 
# IMPORTANT: intercept_dispersion is the intercept 'c' in log(σ) = -k·log(μ) + c
# - Higher intercept_dispersion → Higher log(σ) → Higher σ → MORE overdispersion
# - Lower intercept_dispersion → Lower log(σ) → Lower σ → LESS overdispersion
# 
# For differential overdispersion (IBD > non-IBD):
# - non-IBD: intercept_dispersion = intercept_base (e.g., -6)
# - IBD: intercept_dispersion = intercept_base + delta (e.g., -5.5)
# - Since -5.5 > -6, IBD has HIGHER σ and MORE overdispersion ✓
#
# Note: This is NOT precision! Precision ≈ 1/σ, so higher σ = lower precision.
target_delta <- 0.5
intercept_base <- -6
intercept_disp_cohort <- c(intercept_base, intercept_base + target_delta)
intercept_disp_matrix <- outer(intercept_disp_cohort, rep(1, sim_params$n_taxa))
base_seed <- unique(sweep_setting_DO_no_sd_jobs$seed)[1]

source("_targets_BritoIL_2016.R")

sim_realistic_taxa <- simulate_case_realistic_taxa(
  slope_vector = rep(0, sim_params$n_taxa),  # No DA
  mu_inv_softmax = sim_params$mu_inv_softmax_base_realistic,
  log_dispersion_assoc = 0,
  intercept_dispersion_matrix = intercept_disp_matrix,
  sd_log_overdispersion = 0,
  seed = base_seed,
  n_taxa = sim_params$n_taxa,
  n_samples_per_group = sim_params$n_samples_per_group,
  library_size_mean = sim_params$library_size_mean,
  library_size_sd = sim_params$library_size_sd
) |> synthetic_sim_add_two_group_factor(c("stool", "oralcavity"))

# Filter and run analysis to get full results for plotting
# Use 0.1% threshold for Figure 2 (as stated in the text)
threshold_fig2 <- 0.001  # 0.1% threshold for Figure 2
sim_realistic_taxa_filtered <- sim_realistic_taxa
sim_realistic_taxa_filtered$count_long <- filter_by_abundance_cases(
  sim_realistic_taxa$count_long, 
  threshold = threshold_fig2
)

# Check that we have enough samples and taxa after filtering
n_samples_after <- length(unique(sim_realistic_taxa_filtered$count_long$sample_id))
n_taxa_after <- length(unique(sim_realistic_taxa_filtered$count_long$taxon_id))
samples_by_group <- sim_realistic_taxa_filtered$count_long %>%
  dplyr::distinct(sample_id, group) %>%
  dplyr::count(group)
cat("After filtering (threshold =", threshold_fig2, "):\n")
cat("  Samples:", n_samples_after, "\n")
cat("  Taxa:", n_taxa_after, "\n")
cat("  Samples by group:\n")
print(samples_by_group)

if (n_samples_after < 2) {
  stop("Not enough samples after filtering (need at least 2, got ", n_samples_after, ")")
}
if (n_taxa_after < 3) {
  stop("Not enough taxa after filtering (need at least 3, got ", n_taxa_after, ")")
}
if (nrow(samples_by_group) < 2 || any(samples_by_group$n < 1)) {
  stop("Not all groups have samples after filtering")
}

# Run analysis to get full results for plotting
analysis_realistic_taxa_filtered <- run_analysis_job(
  sim_r = sim_realistic_taxa_filtered,
  return_full = TRUE
)
```

```{r figure-2-filtered-comparison}
# sim_realistic_taxa_filtered is already created in figure-2-setup chunk
# Just verify it exists and get statistics

n_taxa_before_filter_realistic_taxa <- length(unique(sim_realistic_taxa$count_long$taxon_id))
cat("=== Figure 2 Filtering Statistics ===\n")
cat("Number of taxa BEFORE filtering:", n_taxa_before_filter_realistic_taxa, "\n")

n_taxa_after_filter_realistic_taxa <- length(unique(sim_realistic_taxa_filtered$count_long$taxon_id))
cat("Number of taxa AFTER 0.1% filtering:", n_taxa_after_filter_realistic_taxa, "\n")
cat("Percentage retained:", round(100 * n_taxa_after_filter_realistic_taxa / n_taxa_before_filter_realistic_taxa, 1), "%\n")
cat("=====================================\n\n")

# Store for later use
cat("\n**Figure 2 Filtering Summary:**\n")
cat("- Before filtering:", n_taxa_before_filter_realistic_taxa, "taxa\n")
cat("- After 0.1% filtering:", n_taxa_after_filter_realistic_taxa, "taxa\n")
cat("- Percentage retained:", round(100 * n_taxa_after_filter_realistic_taxa / n_taxa_before_filter_realistic_taxa, 1), "%\n")

# Display filtering statistics
cat("\n")
cat("**Figure 2 Filtering Statistics:**\n")
cat("- Number of taxa BEFORE filtering:", n_taxa_before_filter_realistic_taxa, "\n")
cat("- Number of taxa AFTER 0.1% filtering:", n_taxa_after_filter_realistic_taxa, "\n")
cat("- Percentage retained:", round(100 * n_taxa_after_filter_realistic_taxa / n_taxa_before_filter_realistic_taxa, 1), "%\n")
cat("\n")

# Format analysis results for plotting (already filtered in the simulation)
cmp_realistic_taxa_filtered <- format_analysis_results_for_plotting(
  analysis_results = analysis_realistic_taxa_filtered,
  case_label = "Setting DO_no_sd (filtered, delta = 0.5)"
)
```

```{r figure-2-DO-sweep-data}
# Load Setting DO_no_sd FPR data for FNR/centroid plots
sweep_setting_DO_no_sd_fpr <- targets::tar_read("sweep_setting_DO_no_sd_fpr", store = "_targets_BritoIL_2016")
```


```{r fig-2-components}
# Figure 2 components (Setting DO_no_sd filtered)
cols_prgn <- c(
  stool = RColorBrewer::brewer.pal(11, "PRGn")[10],
  oralcavity = RColorBrewer::brewer.pal(11, "PRGn")[2]
)

p_std_realistic_taxa_filtered <- cmp_realistic_taxa_filtered$stats %>% dplyr::filter(method == "permdisp_bray") %>% dplyr::pull(p_value)
se_std_realistic_taxa_filtered <- cmp_realistic_taxa_filtered$stats %>% dplyr::filter(method == "permdisp_bray") %>% dplyr::pull(p_se)
n_std_realistic_taxa_filtered <- cmp_realistic_taxa_filtered$stats %>% dplyr::filter(method == "permdisp_bray") %>% dplyr::pull(n_perm)

std_gg_realistic_taxa_filtered <- plot_betadisper_pcoa_and_boxplot(
  cmp_realistic_taxa_filtered$bd$standard,
  "PERMDISP (Bray–Curtis)",
  cols = cols_prgn,
  p_value = p_std_realistic_taxa_filtered,
  p_se = se_std_realistic_taxa_filtered,
  n_perm = n_std_realistic_taxa_filtered,
  show_group_legend = TRUE,
  horizontal_box = TRUE
)
# Use delta = 2 simulation for alpha plot in PCoA grid
alpha_plot_realistic_taxa_filtered <- plot_alpha_diversity_scatter(sim_realistic_taxa_filtered, case_label = "Setting DO_no_sd (filtered)") +
  ggplot2::scale_color_manual(values = cols_prgn, guide = "none") +
  ggplot2::scale_fill_manual(values = cols_prgn, guide = "none")

disp_scatter_realistic_taxa_filtered <- plot_mean_dispersion_scatter(sim_realistic_taxa_filtered, case_label = "Setting DO_no_sd (filtered)", cols = cols_prgn)

# Filter distances for non-IBD methods
distances_non_ibd_realistic_taxa_filtered <- cmp_realistic_taxa_filtered$distances_long %>%
  dplyr::mutate(method_char = as.character(method)) %>%
  dplyr::filter(!startsWith(method_char, "ibd")) %>%
  dplyr::mutate(method_label = sapply(method_char, pretty_method_label_box)) %>%
  dplyr::mutate(method_group = dplyr::case_when(
    method_label %in% c("Aitchison", "robust Aitchison") ~ "Aitchison methods",
    TRUE ~ "Other methods"
  ))

# Create boxplot panel for all methods (excluding IBD) with facets
p_boxplots_all_realistic_taxa_filtered <- ggplot(distances_non_ibd_realistic_taxa_filtered, aes(x = method_label, y = distance, fill = group)) +
  geom_boxplot(width = 0.5, outlier.size = 0.1, alpha = 0.85, linewidth = 0.25, outlier.stroke = 0, position = position_dodge(width = 0.8)) +
  geom_jitter(size = 1.0, alpha = 0.25, stroke = 0, position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.12)) +
  scale_fill_manual(values = cols_prgn, name = NULL) +
  facet_wrap(~ method_group, nrow = 1, scales = "free", space = "free_x") +
  labs(x = NULL, y = "Distance to centroid") +
  theme_multipanel +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
  )

# Create PCoA plots for all methods (excluding IBD) using delta = 2 simulation (original)
# Diagnostic: Verify the actual delta value in the simulation and centroid distances
# Extract parameters from simulation
intercept_dispersion_realistic_taxa <- unique(sim_realistic_taxa$cohort_log_sigma)
sd_log_overdispersion_realistic_taxa <- if("sd_log_overdispersion" %in% names(sim_realistic_taxa$parameters)) sim_realistic_taxa$parameters$sd_log_overdispersion else 0
slope_vector_realistic_taxa <- sim_realistic_taxa$slope_vector

cat("\n=== Figure 2 Row 1: Simulation Parameters ===\n")
cat("Simulation source: Setting DO_no_sd with delta = 0.5\n")
cat("Simulation parameters:\n")
cat("  intercept_dispersion (log) - non-IBD:", round(intercept_dispersion_realistic_taxa[1], 4), "\n")
cat("  intercept_dispersion (log) - IBD:", round(intercept_dispersion_realistic_taxa[2], 4), "\n")
delta_od <- intercept_dispersion_realistic_taxa[2] - intercept_dispersion_realistic_taxa[1]
cat("  Differential overdispersion (delta, log):", round(delta_od, 4), "\n")
cat("  Differential overdispersion (delta, linear scale, fold change):", round(exp(delta_od), 4), "x\n")
cat("  sd_log_overdispersion:", sd_log_overdispersion_realistic_taxa, "\n")
cat("  slope_vector (DA):", if(all(slope_vector_realistic_taxa == 0)) "all zeros (no DA)" else "has non-zero values", "\n")
cat("\n*** SUMMARY: Figure 2 Row 1 uses delta OD =", round(delta_od, 4), "on log (", round(exp(delta_od), 2), "x fold change) ***\n")
cat("*** Expected: delta OD = 0.5 (non-IBD: -6, IBD: -5.5) ***\n")

# Check all methods for centroid distances
cat("\nCentroid distances across all methods:\n")
for (method_key in names(cmp_realistic_taxa_filtered$bd_by_method)) {
  if (startsWith(method_key, "ibd")) next
  bd_check <- cmp_realistic_taxa_filtered$bd_by_method[[method_key]]
  if (is.null(bd_check)) next
  
  group_dists <- bd_check$group.distances
  centroids <- bd_check$centroids
  
  method_label <- pretty_method_label_box(method_key)
  cat("\n", method_label, ":\n", sep = "")
  cat("  non-IBD mean distance to centroid:", round(group_dists[1], 6), "\n")
  cat("  IBD mean distance to centroid:", round(group_dists[2], 6), "\n")
  cat("  Difference (dispersion delta):", round(group_dists[2] - group_dists[1], 6), "\n")
  
  if (nrow(centroids) == 2 && ncol(centroids) >= 2) {
    # Distance in first 2 PCoA dimensions (what's shown in the plot)
    centroid_dist_pcoa12 <- sqrt(sum((centroids[1, 1:2] - centroids[2, 1:2])^2))
    hull_scale <- compute_centroid_distance_pct_hull_radius(bd_check)
    # Distance in all dimensions
    centroid_dist_all <- sqrt(sum((centroids[1,] - centroids[2,])^2))
    
    # Get eigenvalues to see variance explained
    eig <- bd_check$eig
    eig_pct <- if (!is.null(eig) && length(eig) > 0) {
      round(100 * eig[1:min(5, length(eig))] / sum(eig[eig > 0]), 2)
    } else NULL
    
    cat("  Centroid positions (PCoA 1-2):\n")
    cat("    non-IBD:", paste(round(centroids[1, 1:2], 6), collapse = ", "), "\n")
    cat("    IBD:", paste(round(centroids[2, 1:2], 6), collapse = ", "), "\n")
    cat("  Distance between centroids (PCoA 1-2):", round(centroid_dist_pcoa12, 8), "\n")
    cat(
      "  Centroid distance as % of mean hull radius:",
      round(hull_scale$pct_of_mean_hull_radius, 2), "%\n"
    )
    cat("  Distance between centroids (all dims):", round(centroid_dist_all, 8), "\n")
    if (!is.null(eig_pct)) {
      cat("  Variance explained (first 5 dims):", paste(paste0(round(eig_pct, 1), "%"), collapse = ", "), "\n")
    }
    
    # Check if distance is in higher dimensions
    if (ncol(centroids) > 2) {
      dist_by_dim <- sapply(1:min(10, ncol(centroids)), function(i) {
        abs(centroids[1, i] - centroids[2, i])
      })
      cat("  Centroid separation by dimension (first 10):", paste(round(dist_by_dim, 6), collapse = ", "), "\n")
    }
  }
} 
cat("\n=== End diagnostic ===\n\n")

non_ibd_methods_realistic_taxa_filtered <- names(cmp_realistic_taxa_filtered$bd_by_method)[!startsWith(names(cmp_realistic_taxa_filtered$bd_by_method), "ibd")]

pcoa_plots_all_realistic_taxa_filtered <- purrr::imap(non_ibd_methods_realistic_taxa_filtered, function(method_key, idx) {
  bd <- cmp_realistic_taxa_filtered$bd_by_method[[method_key]]
  
  method_label <- pretty_method_label_box(method_key)
  p_val <- cmp_realistic_taxa_filtered$stats %>% 
    dplyr::filter(method == method_key) %>% 
    dplyr::pull(p_value)
  
  gg <- plot_betadisper_pcoa_and_boxplot(
    bd,
    title_prefix = method_label,
    cols = cols_prgn,
    p_value = p_val,
    show_group_legend = FALSE,
    horizontal_box = FALSE
  )
  
  gg$pcoa <- gg$pcoa + 
    ggtitle(method_label) +
    theme(
      plot.title = element_text(size = 7, hjust = 0.5, margin = margin(b = 2))
    )
  
  if (idx > 1) {
    gg$pcoa <- gg$pcoa + theme(axis.title.y = element_blank())
  }
  gg$pcoa
})

# Create PCoA grid with 5 PCoA plots + 1 spacer (2x3 grid)
if (length(pcoa_plots_all_realistic_taxa_filtered) > 0) {
  # Combine PCoA plots with spacer
  all_plots_realistic_taxa_filtered <- c(pcoa_plots_all_realistic_taxa_filtered, list(plot_spacer()))
  p_pcoa_all_realistic_taxa_filtered <- patchwork::wrap_plots(all_plots_realistic_taxa_filtered, nrow = 2, ncol = 3) + patchwork::plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
} else {
  p_pcoa_all_realistic_taxa_filtered <- ggplot() + theme_void()
}

# Calculate False Negative Rate (FNR) for each delta, method, and threshold
# Note: sweep_setting_DO_no_sd_fpr is already loaded in figure-2-DO-sweep-data chunk
fnr_sweep_realistic_taxa_filtered <- sweep_setting_DO_no_sd_fpr %>%
  dplyr::filter(!grepl("^IBD", method, ignore.case = TRUE)) %>%
  dplyr::group_by(delta, method, threshold) %>%
  dplyr::summarise(
    n_sims = dplyr::n(),
    n_fn = sum(p_value > 0.05),
    FNR = n_fn / n_sims,
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    method = coerce_method_levels(method),
    threshold_label = dplyr::case_when(
      threshold == 0.0 ~ "unfiltered",
      threshold == 0.005 ~ "0.5%",
      threshold == 0.001 ~ "0.1%",
      TRUE ~ as.character(threshold)
    ),
    alpha_val = dplyr::case_when(
      threshold == 0.005 ~ 1.0,
      threshold == 0.001 ~ 0.55,
      threshold == 0.0 ~ 0.25,
      TRUE ~ 1.0
    )
  )

p_sweep_fnr_realistic_taxa_filtered <- ggplot(fnr_sweep_realistic_taxa_filtered, aes(x = delta, y = FNR, color = method, alpha = alpha_val)) +
  geom_line(linewidth = 0.1, aes(group = interaction(method, threshold))) +
  geom_point(size = 1.0, stroke = 0, aes(group = interaction(method, threshold))) +
  scale_color_manual(values = roc_colors_all, breaks = method_levels, drop = FALSE, name = NULL) +
  scale_alpha_identity(guide = "none") +
  scale_x_continuous(breaks = {
    breaks <- unique(fnr_sweep_realistic_taxa_filtered$delta)
    breaks[breaks == 0 | breaks >= 0.25]
  }, labels = format_axis_labels) +
  scale_y_continuous(labels = format_axis_labels, limits = c(0, 1)) +
  labs(
    x = "Differential dispersion (log)",
    y = "False negative rate"
  ) +
  theme_multipanel

# Calculate centroid separation as % of average hull radius for each delta, method, and threshold
centroid_sweep_realistic_taxa_filtered_raw <- sweep_setting_DO_no_sd_fpr %>%
  dplyr::filter(!grepl("^IBD", method, ignore.case = TRUE)) %>%
  dplyr::group_by(delta, method, threshold) %>%
  dplyr::summarise(
    n_sims = dplyr::n(),
    mean_centroid_pct = mean(centroid_sep_pct_hull_radius, na.rm = TRUE),
    se_centroid_pct = sd(centroid_sep_pct_hull_radius, na.rm = TRUE) / sqrt(dplyr::n()),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    method = coerce_method_levels(method),
    threshold_label = dplyr::case_when(
      threshold == 0.0 ~ "unfiltered",
      threshold == 0.005 ~ "0.5%",
      threshold == 0.001 ~ "0.1%",
      TRUE ~ as.character(threshold)
    ),
    alpha_val = dplyr::case_when(
      threshold == 0.005 ~ 1.0,
      threshold == 0.001 ~ 0.55,
      threshold == 0.0 ~ 0.25,
      TRUE ~ 1.0
    )
  )

centroid_sweep_realistic_taxa_filtered <- centroid_sweep_realistic_taxa_filtered_raw

# Mixed-effects association model:
# centroid shift (%) ~ differential dispersion (delta)
# with random intercept and slope per method-threshold combination.
lmm_data_fig2 <- sweep_setting_DO_no_sd_fpr %>%
  dplyr::filter(!grepl("^IBD", method, ignore.case = TRUE)) %>%
  dplyr::filter(is.finite(centroid_sep_pct_hull_radius)) %>%
  dplyr::mutate(
    method_threshold = interaction(method, threshold, drop = TRUE)
  )

lmm_full_fig2 <- lme4::lmer(
  centroid_sep_pct_hull_radius ~ delta + (1 + delta | method_threshold),
  data = lmm_data_fig2,
  REML = FALSE
)
lmm_null_fig2 <- lme4::lmer(
  centroid_sep_pct_hull_radius ~ 1 + (1 + delta | method_threshold),
  data = lmm_data_fig2,
  REML = FALSE
)

lmm_lrt_fig2 <- stats::anova(lmm_null_fig2, lmm_full_fig2)
lmm_p_delta_fig2 <- lmm_lrt_fig2$`Pr(>Chisq)`[2]
lmm_fixef_fig2 <- lme4::fixef(lmm_full_fig2)
lmm_intercept_fig2 <- unname(lmm_fixef_fig2["(Intercept)"])
lmm_slope_fig2 <- unname(lmm_fixef_fig2["delta"])

fmt_p_lmm <- function(p) {
  if (!is.finite(p)) return("NA")
  if (p < 0.001) return("<0.001")
  sprintf("%.3f", p)
}
lmm_label_fig2 <- paste0("p = ", fmt_p_lmm(lmm_p_delta_fig2))

p_sweep_realistic_taxa_filtered <- ggplot(centroid_sweep_realistic_taxa_filtered, aes(x = delta, y = mean_centroid_pct, color = method, alpha = alpha_val)) +
  geom_ribbon(aes(ymin = mean_centroid_pct - se_centroid_pct, 
                  ymax = mean_centroid_pct + se_centroid_pct, fill = method, group = interaction(method, threshold)), 
              alpha = 0.2, color = NA) +
  geom_line(linewidth = 0.1, aes(group = interaction(method, threshold))) +
  geom_point(size = 1.0, stroke = 0, aes(group = interaction(method, threshold))) +
  geom_abline(
    intercept = lmm_intercept_fig2,
    slope = lmm_slope_fig2,
    color = "black",
    linewidth = 0.1,
    linetype = "dotted"
  ) +
  annotate(
    geom = "text",
    x = 0.1,
    y = 18,
    label = lmm_label_fig2,
    hjust = 0,
    vjust = 1,
    size = 2.6,
    fontface = "plain",
    color = "black"
  ) +
  scale_color_manual(values = roc_colors_all, breaks = method_levels, drop = FALSE, name = NULL) +
  scale_fill_manual(values = roc_colors_all, breaks = method_levels, drop = FALSE, guide = "none") +
  scale_alpha_identity(guide = "none") +
  scale_x_continuous(breaks = {
    breaks <- unique(centroid_sweep_realistic_taxa_filtered$delta)
    breaks[breaks == 0 | breaks >= 0.25]
  }, labels = format_axis_labels) +
  scale_y_continuous(labels = format_axis_labels, limits = c(0, NA)) +
  labs(
    x = "Differential dispersion (log)",
    y = "Centroid shift (% of mean hull radius)"
  ) +
  coord_cartesian(clip = "off") +
  theme_multipanel +
  theme(
    panel.grid.major.y = element_line(linewidth = 0.1, color = "grey90"),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank()
  )

# Get top 30 most abundant taxa for whisker plot (Figure 2, Setting DO_no_sd filtered)
# Use the ORIGINAL sim_realistic_taxa (before filtering) to identify top 30 taxa, then filter those
# This ensures we get the top 30 from the full dataset, not just what remains after 0.1% filtering
taxon_abundance_realistic_taxa_original <- sim_realistic_taxa$count_long %>%
  dplyr::mutate(proportion = count / library_size) %>%
  dplyr::group_by(taxon_id) %>%
  dplyr::summarise(mean_prop = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_prop)) %>%
  dplyr::slice_head(n = 30) %>%
  dplyr::pull(taxon_id)

# Check how many of the top 30 remain after filtering
top30_after_filter <- sum(taxon_abundance_realistic_taxa_original %in% unique(sim_realistic_taxa_filtered$count_long$taxon_id))

cat("\n**Whisker Plot Taxa Selection:**\n")
cat("- Top 30 taxa selected from original", n_taxa_before_filter_realistic_taxa, "taxa\n")
cat("- Top 30 taxa remaining after 0.1% filter:", top30_after_filter, "out of 30\n")
cat("- Taxa shown in whisker plot:", top30_after_filter, "\n")
cat("\n")

# Create filtered simulation with only top 30 taxa for whisker plot
# Use the top 30 from original, but show data from filtered simulation
sim_realistic_taxa_filtered_top30 <- sim_realistic_taxa_filtered
sim_realistic_taxa_filtered_top30$count_long <- sim_realistic_taxa_filtered$count_long %>%
  dplyr::filter(taxon_id %in% taxon_abundance_realistic_taxa_original)

# Display filtering statistics
cat("\n**Case 4.5 Filtering Statistics:**\n")
cat("- Number of taxa BEFORE filtering:", n_taxa_before_filter_realistic_taxa, "\n")
cat("- Number of taxa AFTER 0.1% filtering:", n_taxa_after_filter_realistic_taxa, "\n")
cat("- Percentage retained:", round(100 * n_taxa_after_filter_realistic_taxa / n_taxa_before_filter_realistic_taxa, 1), "%\n")
cat("- Top 30 taxa remaining after filter:", top30_after_filter, "out of 30\n")
cat("- Taxa shown in whisker plot:", top30_after_filter, "\n")
cat("\n")

# Create whisker plot for top 20 most abundant filtered taxa
arcsin_sqrt_trans_realistic_taxa <- scales::trans_new(
  name = "arcsin_sqrt",
  transform = function(x) asin(sqrt(pmax(0, pmin(1, x)))),
  inverse = function(y) sin(y)^2,
  breaks = scales::extended_breaks(),
  domain = c(0, 1)
)

d_whisker_realistic_taxa_all <- sim_realistic_taxa_filtered$count_long %>%
  dplyr::filter(taxon_id %in% taxon_abundance_realistic_taxa_original) %>%
  dplyr::mutate(proportion = count / library_size) %>%
  dplyr::group_by(group, taxon_id) %>%
    dplyr::summarise(
    mean_prop = mean(proportion, na.rm = TRUE),
    p_low = stats::quantile(proportion, 0.025, na.rm = TRUE),
    p_high = stats::quantile(proportion, 0.975, na.rm = TRUE),
      .groups = "drop"
  )

taxon_order_whisker_realistic_taxa_all <- d_whisker_realistic_taxa_all %>%
  dplyr::group_by(taxon_id) %>%
  dplyr::summarise(overall_mean = mean(mean_prop, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(overall_mean)) %>%
  dplyr::pull(taxon_id)

d_whisker_realistic_taxa_all <- d_whisker_realistic_taxa_all %>%
  dplyr::mutate(taxon_id = factor(taxon_id, levels = taxon_order_whisker_realistic_taxa_all))

prop_breaks_whisker_realistic_taxa_all <- c(0.001, 0.01, 0.05, 0.1, 0.2, 0.3, 0.5)
prop_breaks_whisker_realistic_taxa_all <- prop_breaks_whisker_realistic_taxa_all[prop_breaks_whisker_realistic_taxa_all <= max(d_whisker_realistic_taxa_all$p_high, na.rm = TRUE) * 1.1]

p_whisker_realistic_taxa_filtered_all <- ggplot(d_whisker_realistic_taxa_all, aes(x = taxon_id, y = mean_prop, ymin = p_low, ymax = p_high, color = group)) +
  ggplot2::geom_linerange(position = ggplot2::position_dodge(width = 0.6), linewidth = 0.5, alpha = 0.8) +
  ggplot2::geom_point(position = ggplot2::position_dodge(width = 0.6), size = 1.0, stroke = 0) +
  ggplot2::scale_color_manual(values = cols_prgn, name = NULL) +
  ggplot2::scale_y_continuous(
    trans = arcsin_sqrt_trans_realistic_taxa,
    breaks = prop_breaks_whisker_realistic_taxa_all,
    labels = function(x) gsub("\\.0%", "%", scales::percent_format(accuracy = 0.1)(x))
  ) +
  ggplot2::labs(
    x = "Taxon (ordered by mean proportion)",
    y = "Proportion (arcsin-sqrt scale, 95% CI)"
  ) +
  theme_multipanel +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    legend.position = "bottom"
  )

# Second row: PCoA grid on left, sweep plots on right (whisker plot moved to first row)
fig_2_case4_5 <- (p_pcoa_all_realistic_taxa_filtered | (p_sweep_fnr_realistic_taxa_filtered / p_sweep_realistic_taxa_filtered) + patchwork::plot_layout(heights = c(1, 1))) + 
  patchwork::plot_layout(guides = "collect", widths = c(1, 1)) &
  theme(plot.margin = margin(0, 0, 0, 0, "pt"), legend.key.size = unit(0.2, 'cm'), legend.position = "bottom")
```

```{r fig-2-combined-filtered, fig.width=16, fig.height=14}
#| fig-cap: "Setting DO_no_sd (filtered): Differential overdispersion with no differential abundance, with abundance filtering. Realistic number of taxa, no random variation in overdispersion (0.1% threshold)."

# First row: whisker plot (all filtered taxa) | alpha diversity | boxplot
fig_2_row1_new <- (p_whisker_realistic_taxa_filtered_all | alpha_plot_realistic_taxa_filtered | p_boxplots_all_realistic_taxa_filtered) + 
  patchwork::plot_layout(widths = c(0.35, 0.325, 0.325), guides = "collect") &
  theme(plot.margin = margin(0, 0, 0, 0, "pt"), legend.key.size = unit(0.2, 'cm'), legend.position = "bottom")

# Combine rows: Row 1 (spacer | whisker | alpha), Row 2 (sweep plots | PCoA grid)
fig_2 <- (fig_2_row1_new / fig_2_case4_5) + 
  patchwork::plot_layout(guides = "collect", heights = c(1, 2)) &
  theme(plot.margin = margin(0, 0, 0, 0, "pt"), legend.key.size = unit(0.2, 'cm'), legend.position = "bottom")

fig_2
```

```{r fig-2-save-pdf}
#| echo: false
ggsave("Figure_2_BritoIL_2016.pdf", fig_2, width = 183, height = 138, units = "mm")
```

```{r cleanup-after-figure-2}
# Release heavy Figure 2 objects before next sections
rm(list = intersect(
  c(
    "sim_realistic_taxa", "sim_realistic_taxa_filtered", "analysis_realistic_taxa_filtered",
    "cmp_realistic_taxa_filtered", "std_gg_realistic_taxa_filtered",
    "p_pcoa_all_realistic_taxa_filtered", "p_boxplots_all_realistic_taxa_filtered",
    "alpha_plot_realistic_taxa_filtered", "disp_scatter_realistic_taxa_filtered",
    "p_whisker_realistic_taxa_filtered_all", "fig_2_case4_5", "fig_2_row1_new", "fig_2",
    "distances_non_ibd_realistic_taxa_filtered", "d_whisker_realistic_taxa_all",
    "taxon_abundance_realistic_taxa_original", "sim_realistic_taxa_filtered_top30",
    "centroid_sweep_realistic_taxa_filtered", "centroid_sweep_realistic_taxa_filtered_raw",
    "p_sweep_fnr_realistic_taxa_filtered", "p_sweep_realistic_taxa_filtered"
  ),
  ls()
))
invisible(gc())
```

# Simulation no-DO (No Differential Overdispersion)

This section tests method performance when there is **differential abundance but NO differential overdispersion**. This evaluates whether methods produce false positives from composition differences when there is no true differential dispersion.

## Figure 3: Setting DA_no_sd (filtered) and Setting DA_sd (filtered)

This figure shows method performance on filtered simulated data (taxa with mean abundance <0.1% removed) with differential abundance but no differential overdispersion.

```{r shared-setting-params}
# Parameters for Setting DA_no_sd
intercept_disp_DA_no_sd <- intercept_disp_realistic
sigma_DA_no_sd <- sigma_realistic
mu_base_DA_no_sd <- mu_inv_softmax_base_realistic
k_fixed_DA_no_sd <- 0

# Parameters for Setting DA_sd
intercept_disp_DA_sd <- intercept_disp_realistic
sigma_DA_sd <- sigma_realistic
mu_base_DA_sd <- mu_inv_softmax_base_realistic
k_fixed_DA_sd <- 0

standard_distance_methods_setting5 <- c("bray", "aitchison", "robust.aitchison", "jaccard", "hellinger")
group_levels <- c("stool", "oralcavity")

cols_prgn_setting5 <- c(
  stool = RColorBrewer::brewer.pal(11, "PRGn")[10],
  oralcavity = RColorBrewer::brewer.pal(11, "PRGn")[2]
)

# Whole-taxon (column) filter only: mean proportion is computed per taxon across *all* samples,
# then taxa below the threshold are removed entirely. No per-cell or per-sample-row dropping.
filter_by_abundance_for_plot <- function(count_long, min_abundance) {
  taxon_means <- count_long %>%
    dplyr::mutate(proportion = count / library_size) %>%
    dplyr::group_by(taxon_id) %>%
    dplyr::summarise(mean_prop = mean(proportion, na.rm = TRUE), .groups = "drop")
  keep_taxa <- taxon_means %>% dplyr::filter(mean_prop >= min_abundance) %>% dplyr::pull(taxon_id)
  count_long %>% dplyr::filter(taxon_id %in% keep_taxa)
}
```

### Setting DA_no_sd (filtered): Differential abundance, no sd on overdispersion, with abundance filtering (<0.1%)

This setting uses the same simulated data as Setting DA_no_sd but filters out taxa with mean abundance <0.1% before computing distances and running PERMDISP. This tests whether the trends observed in Setting DA_no_sd are driven by low abundance taxa.

**Note:** A 0.1% mean abundance threshold is more appropriate than 5% for microbiome data, as 5% would remove most taxa. PERMDISP requires at least 3 taxa to function, but more taxa are preferable for stable results.

```{r figure-3-seed-selection}
# Seed selection for Figure 3 (PCoA plots and top row)
# Change seed_idx to try different seeds and find a more representative example
seed_idx_fig3 <- 1
```

```{r setting-DA-no-sd-filtered-pcoa-plots}
# Create PCoA panel for Setting DA_no_sd (filtered) using sweep analysis outputs
# Use threshold = 0.001 (0.1%) for filtered plots
sweep_setting_DA_no_sd_jobs <- targets::tar_read("sweep_setting_DA_no_sd_jobs", store = "_targets_BritoIL_2016")
sweep_setting_DA_no_sd_jobs_with_threshold <- targets::tar_read("sweep_setting_DA_no_sd_jobs_with_threshold", store = "_targets_BritoIL_2016")
sweep_setting_DA_no_sd_res <- targets::tar_read("sweep_setting_DA_no_sd_res", store = "_targets_BritoIL_2016")
# Use the same seed for all cases (1, 10, 40 DA) for consistency
available_seeds <- unique(sweep_setting_DA_no_sd_jobs$seed)
seed_fig3_pcoa <- available_seeds[min(seed_idx_fig3, length(available_seeds))]
threshold_filtered <- 0.001  # Use 0.1% threshold for filtered plots

# Get analysis results for filtered (0.5% threshold) - all using the same seed
analysis_1_raw <- get_sweep_analysis_with_threshold(sweep_setting_DA_no_sd_res, sweep_setting_DA_no_sd_jobs_with_threshold, seed_fig3_pcoa, 1, threshold_filtered)
analysis_10_raw <- get_sweep_analysis_with_threshold(sweep_setting_DA_no_sd_res, sweep_setting_DA_no_sd_jobs_with_threshold, seed_fig3_pcoa, 10, threshold_filtered)
analysis_40_raw <- get_sweep_analysis_with_threshold(sweep_setting_DA_no_sd_res, sweep_setting_DA_no_sd_jobs_with_threshold, seed_fig3_pcoa, 40, threshold_filtered)
if (is.null(analysis_1_raw)) stop("analysis_1_raw is NULL")
if (is.null(analysis_10_raw)) stop("analysis_10_raw is NULL")
if (is.null(analysis_40_raw)) stop("analysis_40_raw is NULL")

# Format analysis results for plotting
analysis_1 <- format_analysis_results_for_plotting(analysis_1_raw, case_label = "1 DA")
analysis_10 <- format_analysis_results_for_plotting(analysis_10_raw, case_label = "10 DA")
analysis_40 <- format_analysis_results_for_plotting(analysis_40_raw, case_label = "40 DA")

pcoa_panel_setting_DA_no_sd_filtered <- plot_pcoa_grid_multiple_methods(
  analysis_list = list(analysis_1, analysis_10, analysis_40),
  n_da_labels = c("1", "10", "40")
)

# Verify the plot was created
if (is.null(pcoa_panel_setting_DA_no_sd_filtered)) {
  stop("pcoa_panel_setting_DA_no_sd_filtered is NULL after creation")
}

# Free temporary analysis objects; keep shared sweep targets for downstream chunks
rm(list = intersect(
  c(
    "available_seeds", "seed_fig3_pcoa", "threshold_filtered",
    "analysis_1_raw", "analysis_10_raw", "analysis_40_raw",
    "analysis_1", "analysis_10", "analysis_40"
  ),
  ls()
))
invisible(gc())
```

```{r setting-DA-no-sd-filtered-top-panel}
# Get FPR data for all thresholds (0.005 opaque, 0.001 lighter, 0.0 lightest)
fpr_setting_DA_no_sd_filtered <- targets::tar_read("sweep_setting_DA_no_sd_fpr", store = "_targets_BritoIL_2016")

p_setting_DA_no_sd_filtered_fpr <- plot_da_taxa_sweep_results(
  fpr_data = fpr_setting_DA_no_sd_filtered,
  setting_name = paste0("Setting DA_no_sd (filtered): Differential abundance, no sd on overdispersion, with abundance filter (≥0.1%) (n=", n_samples, ", p=", n_taxa, 
                       ", c=", round(intercept_disp_DA_no_sd, 2), 
                       ", σ=", round(sigma_DA_no_sd, 5), 
                       ", k=", round(k_fixed_DA_no_sd, 3), ")"),
  sigma_val = sigma_DA_no_sd,
  intercept_disp_val = intercept_disp_DA_no_sd,
  x_limits = c(0, 40),
  x_breaks = c(0, 10, 20, 30, 40)
)

```


### PERMANOVA on Setting DA_no_sd (filtered): TPR with true differential composition

This panel uses the same filtered simulations as the first Figure 3 sweep (**Setting DA_no_sd**, `k = 0`, `sd_log_overdispersion = 0`) and evaluates **PERMANOVA true positive rate** (detection rate) when true differential abundance is present (`n_da_taxa > 0`).

```{r setting-DA-no-sd-permanova-tpr, fig.width=12, fig.height=8}
permanova_da_tpr <- targets::tar_read("sweep_setting_DA_no_sd_permanova_tpr", store = "_targets_BritoIL_2016") %>%
  dplyr::mutate(
    method_label = dplyr::case_when(
      .data$method == "permanova_bray" ~ "Bray–Curtis",
      .data$method == "permanova_aitchison" ~ "Aitchison",
      .data$method == "permanova_robust.aitchison" ~ "robust Aitchison",
      .data$method == "permanova_robust_aitchison" ~ "robust Aitchison",
      .data$method == "permanova_jaccard" ~ "Jaccard",
      .data$method == "permanova_hellinger" ~ "Hellinger",
      TRUE ~ NA_character_
    ),
    alpha_val = dplyr::case_when(
      .data$threshold == 0.0 ~ 0.25,
      .data$threshold == 0.005 ~ 1.0,
      .data$threshold == 0.001 ~ 0.55,
      TRUE ~ 1.0
    ),
    threshold_label = dplyr::case_when(
      .data$threshold == 0.0 ~ "unfiltered",
      .data$threshold == 0.005 ~ "0.5%",
      .data$threshold == 0.001 ~ "0.1%",
      TRUE ~ as.character(.data$threshold)
    )
  ) %>%
  dplyr::filter(!is.na(.data$method_label)) %>%
  dplyr::mutate(method_label = coerce_method_levels(.data$method_label))

p_setting_DA_no_sd_permanova_tpr <- ggplot(
  permanova_da_tpr,
  aes(x = n_da_taxa, y = TPR, color = method_label, alpha = alpha_val)
) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "grey60", linewidth = 0.1) +
  geom_line(linewidth = 0.1, aes(group = interaction(method_label, threshold))) +
  geom_point(size = 1.0, stroke = 0, aes(group = interaction(method_label, threshold))) +
  scale_color_manual(values = roc_colors_all, breaks = method_levels, drop = FALSE, name = NULL) +
  scale_alpha_identity(guide = "none") +
  scale_x_continuous(
    breaks = c(0, 10, 20, 30, 40),
    limits = c(0, 40),
    labels = format_axis_labels
  ) +
  scale_y_continuous(labels = format_axis_labels, limits = c(0, 1)) +
  labs(
    x = "Number of DA taxa",
    y = "True positive rate (PERMANOVA)"
  ) +
  theme_multipanel

# Keep only the final ggplot object
rm(list = intersect(c("permanova_da_tpr"), ls()))
invisible(gc())
```

```{r figure-3-scatter}
standard_methods_for_scatter <- c("Bray–Curtis", "Aitchison", "robust Aitchison", "Jaccard", "Hellinger")

# Figure 2 provides the *power* side for differential dispersion.
# Summarize **FNR** over delta > 0 (exclude null), by method and threshold.
fnr_fig2 <- fnr_sweep_realistic_taxa_filtered %>%
  dplyr::filter(delta > 0) %>%
  dplyr::group_by(method, threshold) %>%
  dplyr::summarise(
    avg_fnr = mean(FNR, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::filter(method %in% standard_methods_for_scatter) %>%
  dplyr::mutate(
    threshold_label = dplyr::case_when(
      threshold == 0.005 ~ "0.5%",
      threshold == 0.001 ~ "0.1%",
      threshold == 0.0 ~ "unfiltered",
      TRUE ~ as.character(threshold)
    ),
    alpha_val = dplyr::case_when(
      threshold == 0.005 ~ 1.0,
      threshold == 0.001 ~ 0.55,
      threshold == 0.0 ~ 0.25,
      TRUE ~ 1.0
    )
  )

cat("\n=== Average FNR from Figure 2 (delta > 0) ===\n")
print(fnr_fig2 %>% dplyr::arrange(method))

# Figure 3 provides the *FPR* side from the DA sweep:
# Average FPR across all n_da_taxa values for Setting DA_no_sd, by method and threshold.
fpr_fig3_data <- fpr_setting_DA_no_sd_filtered %>%
    dplyr::mutate(method_label = dplyr::case_when(
      method == "permdisp_bray" ~ "Bray–Curtis",
      method == "permdisp_aitchison" ~ "Aitchison",
      method == "permdisp_robust.aitchison" ~ "robust Aitchison",
    method == "permdisp_robust_aitchison" ~ "robust Aitchison",
      method == "permdisp_jaccard" ~ "Jaccard",
      method == "permdisp_hellinger" ~ "Hellinger",
    as.character(method) %in% standard_methods_for_scatter ~ as.character(method),
    TRUE ~ NA_character_
    )) %>%
  dplyr::filter(!is.na(method_label))

fpr_col <- if ("FPR" %in% names(fpr_fig3_data)) "FPR" else "fpr"

# Average FPR across all n_da_taxa values for each method and threshold
fpr_fig3 <- fpr_fig3_data %>%
  dplyr::group_by(method_label, threshold) %>%
  dplyr::summarise(
    avg_fpr = mean(.data[[fpr_col]], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::rename(method = method_label) %>%
  dplyr::filter(method %in% standard_methods_for_scatter) %>%
  dplyr::mutate(
    threshold_label = dplyr::case_when(
      threshold == 0.005 ~ "0.5%",
      threshold == 0.001 ~ "0.1%",
      threshold == 0.0 ~ "unfiltered",
      TRUE ~ as.character(threshold)
    ),
    alpha_val = dplyr::case_when(
      threshold == 0.005 ~ 1.0,
      threshold == 0.001 ~ 0.55,
      threshold == 0.0 ~ 0.25,
      TRUE ~ 1.0
    )
  )

cat("\n=== Average FPR from Figure 3 Setting DA_no_sd (filtered) (averaged across all n_da_taxa) ===\n")
print(fpr_fig3 %>% dplyr::arrange(method))

scatter_data <- fnr_fig2 %>%
  dplyr::inner_join(fpr_fig3, by = c("method", "threshold", "threshold_label", "alpha_val")) %>%
  dplyr::mutate(
    method = factor(method, levels = standard_methods_for_scatter),
    threshold_label = factor(threshold_label, levels = c("0.5%", "0.1%", "unfiltered"))
  )

plot_data <- scatter_data %>% dplyr::filter(!is.na(avg_fnr) & !is.na(avg_fpr))

# Calculate linear regression manually for full-range line with SD bands
lm_fit <- lm(avg_fpr ~ avg_fnr, data = plot_data)
residual_sd <- sd(residuals(lm_fit))

# Create prediction data across full plot range (0 to max x)
x_range <- c(0, max(plot_data$avg_fnr, na.rm = TRUE) * 1.1)  # Extend slightly beyond data
x_pred <- seq(x_range[1], x_range[2], length.out = 200)
pred_data <- data.frame(avg_fnr = x_pred)
pred_data$avg_fpr <- predict(lm_fit, newdata = pred_data)
pred_data$ymin <- pred_data$avg_fpr - residual_sd
pred_data$ymax <- pred_data$avg_fpr + residual_sd

p_scatter_fnr_fpr <- ggplot(plot_data, aes(x = avg_fnr, y = avg_fpr, color = method)) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_ribbon(data = pred_data, aes(x = avg_fnr, ymin = ymin, ymax = ymax), 
              inherit.aes = FALSE, fill = "lightgrey", alpha = 0.5) +
  geom_line(data = pred_data, aes(x = avg_fnr, y = avg_fpr), 
            inherit.aes = FALSE, color = "grey", linewidth = 0.8) +
  geom_point(aes(shape = threshold_label, alpha = alpha_val), size = 1.1) +
  ggrepel::geom_text_repel(
    data = dplyr::filter(plot_data, threshold == 0.005),
    aes(label = method),
    size = 3,
    show.legend = FALSE
  ) +
  scale_color_manual(values = roc_colors_all[standard_methods_for_scatter], breaks = standard_methods_for_scatter, drop = FALSE, name = NULL) +
  scale_shape_manual(values = c("0.5%" = 16, "0.1%" = 17, "unfiltered" = 15), name = "Threshold") +
  scale_alpha_identity(guide = "none") +
  scale_x_continuous(name = "Average false negative rate (Figure 2; delta > 0)", limits = c(0, NA), labels = format_axis_labels) +
  scale_y_continuous(name = "Average false positive rate (Figure 3 Setting DA_no_sd filtered)", limits = c(0, NA), labels = format_axis_labels) +
  theme_multipanel +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

```{r figure-3-scatter-diagnostics}
# Print diagnostic information about FPR and FNR values used in performance plot
cat("\n=== Average FNR from Figure 2 (delta > 0) ===\n")
print(fnr_fig2 %>% dplyr::arrange(method))

cat("\n=== Average FPR from Figure 3 Setting DA_no_sd (filtered) (averaged across all n_da_taxa) ===\n")
print(fpr_fig3 %>% dplyr::arrange(method))
```

```{r figure-3-top-row}
# Create plots for 40 DA case (Setting DA_no_sd) for top row of Figure 3
# Use 40 DA simulation + analysis from sweep outputs (then filter for plotting)
if (!exists("sweep_setting_DA_no_sd_jobs", inherits = FALSE)) {
  sweep_setting_DA_no_sd_jobs <- targets::tar_read("sweep_setting_DA_no_sd_jobs", store = "_targets_BritoIL_2016")
}
sweep_setting_DA_no_sd_sim <- targets::tar_read("sweep_setting_DA_no_sd_sim", store = "_targets_BritoIL_2016")
if (!exists("sweep_setting_DA_no_sd_jobs_with_threshold", inherits = FALSE)) {
  sweep_setting_DA_no_sd_jobs_with_threshold <- targets::tar_read("sweep_setting_DA_no_sd_jobs_with_threshold", store = "_targets_BritoIL_2016")
}
if (!exists("sweep_setting_DA_no_sd_res", inherits = FALSE)) {
  sweep_setting_DA_no_sd_res <- targets::tar_read("sweep_setting_DA_no_sd_res", store = "_targets_BritoIL_2016")
}
# Use the same seed as PCoA plots for consistency (uses seed_idx_fig3 from figure-3-seed-selection chunk)
available_seeds_fig3 <- unique(sweep_setting_DA_no_sd_jobs$seed)
seed_40_da_fig3 <- available_seeds_fig3[min(seed_idx_fig3, length(available_seeds_fig3))]
# Use threshold = 0.001 (0.1%) for Figure 3
threshold_fig3 <- 0.001
idx_40_da <- get_sweep_index(sweep_setting_DA_no_sd_jobs, seed_40_da_fig3, 40)
sim_40_da <- sweep_setting_DA_no_sd_sim[[idx_40_da]]
sim_40_da_filtered <- sim_40_da
sim_40_da_filtered$count_long <- filter_by_abundance_for_plot(
  sim_40_da$count_long,
  0.001  # Use 0.1% threshold for Figure 3
)

# Run comparison for 40 DA case (filtered analysis from sweeps)
cmp_40_da <- format_analysis_results_for_plotting(
  analysis_results = get_sweep_analysis_with_threshold(sweep_setting_DA_no_sd_res, sweep_setting_DA_no_sd_jobs_with_threshold, seed_40_da_fig3, 40, threshold_fig3),
  case_label = "Setting DA_no_sd (40 DA, filtered)"
)

# Release large sweep containers once the single 40-DA slice is materialized
rm(list = intersect(
  c(
    "sweep_setting_DA_no_sd_jobs", "sweep_setting_DA_no_sd_sim",
    "sweep_setting_DA_no_sd_jobs_with_threshold", "sweep_setting_DA_no_sd_res",
    "available_seeds_fig3", "idx_40_da"
  ),
  ls()
))
invisible(gc())

# Get top 30 most abundant taxa for whisker plot (from original unfiltered simulation)
taxon_abundance_40_da_original <- sim_40_da$count_long %>%
  dplyr::mutate(proportion = count / library_size) %>%
  dplyr::group_by(taxon_id) %>%
  dplyr::summarise(mean_prop = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_prop)) %>%
  dplyr::slice_head(n = 30) %>%
  dplyr::pull(taxon_id)

# Check how many of the top 30 remain after filtering
top30_after_filter_40_da <- sum(taxon_abundance_40_da_original %in% unique(sim_40_da_filtered$count_long$taxon_id))

cat("\n**Figure 3 Whisker Plot Taxa Selection:**\n")
cat("- Top 30 taxa selected from original simulation\n")
cat("- Top 30 taxa remaining after 0.1% filter:", top30_after_filter_40_da, "out of 30\n")
cat("- Taxa shown in whisker plot:", top30_after_filter_40_da, "\n")
cat("\n")

# Create whisker plot for top 30 most abundant taxa (40 DA case, filtered)
arcsin_sqrt_trans_40_da <- scales::trans_new(
  name = "arcsin_sqrt",
  transform = function(x) asin(sqrt(pmax(0, pmin(1, x)))),
  inverse = function(y) sin(y)^2,
  breaks = scales::extended_breaks(),
  domain = c(0, 1)
)

d_whisker_40_da <- sim_40_da_filtered$count_long %>%
    dplyr::filter(taxon_id %in% taxon_abundance_40_da_original) %>%
    dplyr::mutate(proportion = count / library_size) %>%
    dplyr::group_by(group, taxon_id) %>%
    dplyr::summarise(
      mean_prop = mean(proportion, na.rm = TRUE),
      med_prop = stats::median(proportion, na.rm = TRUE),
      p_low = stats::quantile(proportion, 0.025, na.rm = TRUE),
      p_high = stats::quantile(proportion, 0.975, na.rm = TRUE),
      .groups = "drop"
    )
# Whiskers are 2.5%–97.5% empirical quantiles. The point uses the **median**, not the mean:
# with zero-inflated proportions (many zeros + rare large counts), the **mean can lie above
# p_high** even though p_high is a valid upper quantile — then a mean dot looks like the
# "upper quantile is below the dot". The median always lies between p_low and p_high.
# With discrete counts, p_low == p_high often → zero-height interval (looks like a missing whisker).
n_w_zero_height <- sum(dplyr::coalesce(d_whisker_40_da$p_high - d_whisker_40_da$p_low, 0) <= 0, na.rm = TRUE)
cat(
  "- Whisker segments with zero height (p_low == p_high): ",
  n_w_zero_height, " of ", nrow(d_whisker_40_da), " group×taxon cells\n",
  sep = ""
)
  
taxon_order_whisker_40_da <- d_whisker_40_da %>%
    dplyr::group_by(taxon_id) %>%
    dplyr::summarise(overall_mean = mean(mean_prop, na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(overall_mean)) %>%
    dplyr::pull(taxon_id)
  
d_whisker_40_da <- d_whisker_40_da %>%
  dplyr::mutate(taxon_id = factor(taxon_id, levels = taxon_order_whisker_40_da))
  
prop_breaks_whisker_40_da <- c(0.001, 0.01, 0.05, 0.1, 0.2, 0.3, 0.5)
prop_breaks_whisker_40_da <- prop_breaks_whisker_40_da[prop_breaks_whisker_40_da <= max(d_whisker_40_da$p_high, na.rm = TRUE) * 1.1]
  
p_whisker_40_da <- ggplot(d_whisker_40_da, aes(x = taxon_id, y = mean_prop, ymin = p_low, ymax = p_high, color = group)) +
    ggplot2::geom_linerange(position = ggplot2::position_dodge(width = 0.6), linewidth = 0.5, alpha = 0.8) +
    ggplot2::geom_point(position = ggplot2::position_dodge(width = 0.6), size = 1.0, stroke = 0) +
  ggplot2::scale_color_manual(values = cols_prgn_setting5, name = NULL) +
    ggplot2::scale_y_continuous(
    trans = arcsin_sqrt_trans_40_da,
    breaks = prop_breaks_whisker_40_da,
      labels = function(x) gsub("\\.0%", "%", scales::percent_format(accuracy = 0.1)(x))
    ) +
    ggplot2::labs(
      x = "Taxon (ordered by mean proportion)",
      y = "Proportion (arcsin-sqrt scale, 95% CI)"
    ) +
    theme_multipanel +
    ggplot2::theme(
      axis.text.x = ggplot2::element_blank(),
      axis.ticks.x = ggplot2::element_blank(),
      legend.position = "bottom"
    )
  
# Create alpha diversity plot for 40 DA case (filtered)
alpha_plot_40_da <- plot_alpha_diversity_scatter(sim_40_da_filtered, case_label = "Setting DA_no_sd (40 DA, filtered)")
if (!is.null(alpha_plot_40_da)) {
  alpha_plot_40_da <- alpha_plot_40_da +
    ggplot2::scale_color_manual(values = cols_prgn_setting5, guide = "none") +
    ggplot2::scale_fill_manual(values = cols_prgn_setting5, guide = "none")
} else {
  warning("alpha_plot_40_da is NULL")
}

# Create boxplots for 40 DA case (5 methods) with facets
distances_non_ibd_40_da <- cmp_40_da$distances_long %>%
  dplyr::mutate(method_char = as.character(method)) %>%
  dplyr::filter(!startsWith(method_char, "ibd")) %>%
  dplyr::mutate(method_label = sapply(method_char, pretty_method_label_box)) %>%
  dplyr::mutate(method_group = dplyr::case_when(
    method_label %in% c("Aitchison", "robust Aitchison") ~ "Aitchison methods",
    TRUE ~ "Other methods"
  ))

p_boxplots_40_da <- ggplot(distances_non_ibd_40_da, aes(x = method_label, y = distance, fill = group)) +
  geom_boxplot(width = 0.5, outlier.size = 0.1, alpha = 0.85, linewidth = 0.25, outlier.stroke = 0, position = position_dodge(width = 0.8)) +
  geom_jitter(size = 1.0, alpha = 0.25, stroke = 0, position = position_jitterdodge(dodge.width = 0.8, jitter.width = 0.12)) +
  scale_fill_manual(values = cols_prgn_setting5, name = NULL) +
  facet_wrap(~ method_group, nrow = 1, scales = "free", space = "free_x") +
  labs(x = NULL, y = "Distance to centroid") +
  theme_multipanel +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
  )

# Create top row without spacer
p_figure3_row1 <- (p_whisker_40_da | alpha_plot_40_da | p_boxplots_40_da) + 
  patchwork::plot_layout(widths = c(0.35, 0.325, 0.325), guides = "collect") &
  theme(plot.margin = margin(0, 0, 0, 0, "pt"), legend.key.size = unit(0.2, 'cm'))

# Drop heavy intermediates after composing the final row object
rm(list = intersect(
  c(
    "sim_40_da", "sim_40_da_filtered", "cmp_40_da", "seed_40_da_fig3", "threshold_fig3",
    "taxon_abundance_40_da_original", "top30_after_filter_40_da",
    "arcsin_sqrt_trans_40_da", "d_whisker_40_da", "taxon_order_whisker_40_da",
    "prop_breaks_whisker_40_da", "distances_non_ibd_40_da",
    "p_whisker_40_da", "alpha_plot_40_da", "p_boxplots_40_da"
  ),
  ls()
))
invisible(gc())
```

```{r figure-3, fig.width=16, fig.height=12}
#| fig-cap: "Figure 3. Top row: composition-level diagnostics for Setting DA_no_sd at 40 DA taxa (whisker plot, alpha-diversity scatter, and distance-to-centroid boxplots). Bottom-left column: (1) PERMDISP false positive rate sweep for Setting DA_no_sd (filtered), (2) PERMANOVA true positive rate sweep on the same Setting DA_no_sd simulations (filtered), and (3) FNR-FPR summary scatter. Bottom-right column: PCoA panels for Setting DA_no_sd (filtered)."
# Organize Figure 3 in two columns:
# Top row: whisker plot (40 DA) | alpha diversity (40 DA) | boxplots (40 DA)
# Bottom: Left column: 3 rows (Setting DA_no_sd filtered FPR sweep, Setting DA_no_sd PERMANOVA TPR sweep, scatter plot)
#         Right column: PCoA plots for Setting DA_no_sd filtered

# Check that all plots exist and are valid
cat("Checking plots for Figure 3...\n")
cat("  p_figure3_row1:", !is.null(p_figure3_row1), ", class:", class(p_figure3_row1), "\n")
cat("  p_setting_DA_no_sd_filtered_fpr:", !is.null(p_setting_DA_no_sd_filtered_fpr), ", class:", class(p_setting_DA_no_sd_filtered_fpr), "\n")
cat("  p_setting_DA_no_sd_permanova_tpr:", !is.null(p_setting_DA_no_sd_permanova_tpr), ", class:", class(p_setting_DA_no_sd_permanova_tpr), "\n")
cat("  p_scatter_fnr_fpr:", !is.null(p_scatter_fnr_fpr), ", class:", class(p_scatter_fnr_fpr), "\n")
cat("  pcoa_panel_setting_DA_no_sd_filtered:", !is.null(pcoa_panel_setting_DA_no_sd_filtered), ", class:", class(pcoa_panel_setting_DA_no_sd_filtered), "\n")

# Try building patchwork step by step
left_col <- p_setting_DA_no_sd_filtered_fpr / p_setting_DA_no_sd_permanova_tpr / p_scatter_fnr_fpr
cat("  left_col created, class:", class(left_col), "\n")

right_col <- pcoa_panel_setting_DA_no_sd_filtered
cat("  right_col created, class:", class(right_col), "\n")

# Align bottom row with proper spacing
bottom_row <- left_col | right_col
cat("  bottom_row created, class:", class(bottom_row), "\n")

p_figure3 <- p_figure3_row1 / bottom_row
cat("  p_figure3 created, class:", class(p_figure3), "\n")

p_figure3 <- p_figure3 +
  patchwork::plot_layout(guides = "collect", widths = c(1, 1), heights = c(1, 3)) &
  theme(plot.margin = margin(0, 0, 0, 0, "pt"), legend.key.size = unit(0.2, 'cm'), legend.position = "bottom")

p_figure3
```

```{r figure-3-save-pdf}
#| echo: false
ggsave("Figure_3_BritoIL_2016.pdf", p_figure3, width = 183, height = 230, units = "mm")
```

```{r cleanup-after-figure-3}
# Release heavy Figure 3 objects before supplementary figure 3
rm(list = intersect(
  c(
    "p_setting_DA_no_sd_filtered_fpr", "p_setting_DA_no_sd_permanova_tpr", "p_scatter_fnr_fpr",
    "pcoa_panel_setting_DA_no_sd_filtered", "p_figure3_row1", "left_col", "right_col",
    "bottom_row", "p_figure3", "cmp_40_da", "sim_40_da", "sim_40_da_filtered",
    "p_whisker_40_da", "alpha_plot_40_da", "p_boxplots_40_da", "distances_non_ibd_40_da",
    "d_whisker_40_da", "fpr_fig3_data", "fpr_fig3", "fnr_fig2", "scatter_data", "plot_data",
    "permanova_da_results", "permanova_da_tpr",
    "pred_data", "lm_fit"
  ),
  ls()
))
invisible(gc())
```

## Supplementary Figure 3: Setting DA_no_sd and Setting DA_sd

### Setting DA_no_sd: Differential abundance, no sd on overdispersion

This setting sweeps over the number of DA taxa (0, 1, 2, 3, ...) using slopes sampled from **significant DA taxa only** (credible interval excludes 0). Uses **k = 0 (no mean-OD association)** to ensure proper FPR testing at n_da_taxa = 0; with `sd_log_overdispersion = 0`, taxa also share the same baseline OD within and across groups at the null. This tests how method performance changes as we progressively add more DA taxa, starting from no DA (FPR test) to increasing numbers of DA taxa (power test).

```{r supp-fig-3, fig.width=16, fig.height=16}
sweep_setting_DA_no_sd_jobs <- targets::tar_read("sweep_setting_DA_no_sd_jobs", store = "_targets_BritoIL_2016")
sweep_setting_DA_sd_jobs <- targets::tar_read("sweep_setting_DA_sd_jobs", store = "_targets_BritoIL_2016")
sweep_setting_DA_no_sd_res <- targets::tar_read("sweep_setting_DA_no_sd_res", store = "_targets_BritoIL_2016")
sweep_setting_DA_sd_res <- targets::tar_read("sweep_setting_DA_sd_res", store = "_targets_BritoIL_2016")
# Use available seeds (with n_reps_auc=2, we have seeds 12001 and 12002)
available_seeds_supp <- unique(sweep_setting_DA_no_sd_jobs$seed)
seed_1_da_supp <- available_seeds_supp[1]
seed_20_da_supp <- if(length(available_seeds_supp) >= 2) available_seeds_supp[2] else available_seeds_supp[1]
seed_40_da_supp <- if(length(available_seeds_supp) >= 2) available_seeds_supp[2] else available_seeds_supp[1]
idx_1_da <- get_sweep_index(sweep_setting_DA_no_sd_jobs, seed_1_da_supp, 1)
idx_20_da <- get_sweep_index(sweep_setting_DA_no_sd_jobs, seed_20_da_supp, 20)
idx_40_da <- get_sweep_index(sweep_setting_DA_no_sd_jobs, seed_40_da_supp, 40)

# Job tables are no longer needed after index selection
rm(list = intersect(c("sweep_setting_DA_no_sd_jobs", "sweep_setting_DA_sd_jobs", "available_seeds_supp"), ls()))
invisible(gc())

# Format analysis results for plotting
analysis_DA_no_sd_1 <- format_analysis_results_for_plotting(sweep_setting_DA_no_sd_res[[idx_1_da]], case_label = "1 DA")
analysis_DA_no_sd_20 <- format_analysis_results_for_plotting(sweep_setting_DA_no_sd_res[[idx_20_da]], case_label = "20 DA")
analysis_DA_no_sd_40 <- format_analysis_results_for_plotting(sweep_setting_DA_no_sd_res[[idx_40_da]], case_label = "40 DA")

analysis_DA_sd_1 <- format_analysis_results_for_plotting(sweep_setting_DA_sd_res[[idx_1_da]], case_label = "1 DA")
analysis_DA_sd_20 <- format_analysis_results_for_plotting(sweep_setting_DA_sd_res[[idx_20_da]], case_label = "20 DA")
analysis_DA_sd_40 <- format_analysis_results_for_plotting(sweep_setting_DA_sd_res[[idx_40_da]], case_label = "40 DA")

# Large result lists are no longer needed once subset analyses are extracted
rm(list = intersect(c("sweep_setting_DA_no_sd_res", "sweep_setting_DA_sd_res", "idx_1_da", "idx_20_da", "idx_40_da"), ls()))
invisible(gc())

pcoa_panel_setting_DA_no_sd <- plot_pcoa_grid_multiple_methods(
  analysis_list = list(analysis_DA_no_sd_1, analysis_DA_no_sd_20, analysis_DA_no_sd_40),
  n_da_labels = c("1", "20", "40")
)
pcoa_panel_setting_DA_sd <- plot_pcoa_grid_multiple_methods(
  analysis_list = list(analysis_DA_sd_1, analysis_DA_sd_20, analysis_DA_sd_40),
  n_da_labels = c("1", "20", "40")
)

# Keep only the final PCoA panel objects
rm(list = intersect(
  c(
    "analysis_DA_no_sd_1", "analysis_DA_no_sd_20", "analysis_DA_no_sd_40",
    "analysis_DA_sd_1", "analysis_DA_sd_20", "analysis_DA_sd_40"
  ),
  ls()
))
invisible(gc())

if (exists("fpr_setting_DA_no_sd_filtered", inherits = FALSE)) {
  fpr_setting_DA_no_sd <- fpr_setting_DA_no_sd_filtered
} else {
  fpr_setting_DA_no_sd <- targets::tar_read("sweep_setting_DA_no_sd_fpr", store = "_targets_BritoIL_2016")
}
fpr_setting_DA_sd <- targets::tar_read("sweep_setting_DA_sd_fpr", store = "_targets_BritoIL_2016")

p_setting_DA_no_sd <- plot_da_taxa_sweep_results(
  fpr_data = fpr_setting_DA_no_sd,
  setting_name = paste0("Setting DA_no_sd: Differential abundance, no sd on overdispersion (n=", n_samples, ", p=", n_taxa, 
                       ", c=", round(intercept_disp_DA_no_sd, 2), 
                       ", σ=", round(sigma_DA_no_sd, 5), 
                       ", k=", round(k_fixed_DA_no_sd, 3), ")"),
  sigma_val = sigma_DA_no_sd,
  intercept_disp_val = intercept_disp_DA_no_sd
)

p_setting_DA_sd <- plot_da_taxa_sweep_results(
  fpr_data = fpr_setting_DA_sd,
  setting_name = paste0("Setting DA_sd: Differential abundance, with sd on overdispersion (n=", n_samples, ", p=", n_taxa, 
                       ", c=", round(intercept_disp_DA_sd, 2), 
                       ", σ=", round(sigma_DA_sd, 5), 
                       ", k=", round(k_fixed_DA_sd, 3),
                       ", sd_log_od=", round(prec_sd_real, 3), ")"),
  sigma_val = sigma_DA_sd,
  intercept_disp_val = intercept_disp_DA_sd
)

# Keep only final sweep plot objects
rm(list = intersect(c("fpr_setting_DA_no_sd", "fpr_setting_DA_sd"), ls()))
invisible(gc())

p_setting_DA_no_sd_top <- (p_setting_DA_no_sd | pcoa_panel_setting_DA_no_sd) + 
  patchwork::plot_layout(guides = "collect", widths = c(1, 1)) &
  theme(plot.margin = margin(0, 0, 0, 0, "pt"), legend.key.size = unit(0.2, 'cm'), legend.position = "bottom")

p_setting_DA_sd_top <- (p_setting_DA_sd | pcoa_panel_setting_DA_sd) + 
  patchwork::plot_layout(guides = "collect", widths = c(1, 1)) &
  theme(plot.margin = margin(0, 0, 0, 0, "pt"), legend.key.size = unit(0.2, 'cm'), legend.position = "bottom")

fig_supp_3 <- p_setting_DA_no_sd_top / p_setting_DA_sd_top +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(plot.margin = margin(0, 0, 0, 0, "pt"), legend.key.size = unit(0.2, 'cm'), legend.position = "bottom")

fig_supp_3
```

```{r supp-fig-3-save-pdf}
#| echo: false
ggsave("Supp_Figure_3_BritoIL_2016.pdf", fig_supp_3, width = 183, height = 230, units = "mm")
```

```{r cleanup-after-supp-figure-3}
# Final cleanup before summary tables/text
rm(list = intersect(
  c(
    "sweep_setting_DA_no_sd_jobs", "sweep_setting_DA_sd_jobs",
    "sweep_setting_DA_no_sd_res", "sweep_setting_DA_sd_res",
    "analysis_DA_no_sd_1", "analysis_DA_no_sd_20", "analysis_DA_no_sd_40",
    "analysis_DA_sd_1", "analysis_DA_sd_20", "analysis_DA_sd_40",
    "pcoa_panel_setting_DA_no_sd", "pcoa_panel_setting_DA_sd",
    "fpr_setting_DA_no_sd", "fpr_setting_DA_sd",
    "p_setting_DA_no_sd", "p_setting_DA_sd",
    "p_setting_DA_no_sd_top", "p_setting_DA_sd_top", "fig_supp_3"
  ),
  ls()
))
invisible(gc())
```


# Summary

## Parameter Summary Table

```{r parameter-summary}
# Targets use v_effect-derived dispersion (see sccomp_params); prec_* may be NA / not used.
fmt_num <- function(x, d = 4) {
  if (length(x) != 1L || is.na(x)) return("N/A")
  as.character(round(x, d))
}
fmt_rng <- function(x) {
  paste0("[", round(min(x), 2), ", ", round(max(x), 2), "]")
}
prec_slope_txt <- if (exists("prec_slope") && length(prec_slope) == 1L && !is.na(prec_slope)) {
  as.character(round(prec_slope, 4))
} else {
  "N/A (optional fit summary)"
}
prec_int_txt <- if (exists("prec_intercept") && length(prec_intercept) == 1L && !is.na(prec_intercept)) {
  as.character(round(prec_intercept, 4))
} else {
  "N/A (optional fit summary)"
}

param_summary <- tibble::tibble(
  Parameter = c(
    "Number of samples",
    "Number of taxa (genera)",
    "prec_slope (optional; from extract-dispersion-params chunk)",
    "k used in targets (sccomp_params$k_real)",
    "Mean-OD association (= -k)",
    "prec_intercept (optional fit summary)",
    "c from fit (= -prec_intercept); sim uses v_effect for taxon c",
    "Precision SD (prec_sd) from fit, if read",
    "Baseline intercept range",
    "Group effect (slope) range"
  ),
  `sccomp Value` = c(
    as.character(n_samples_real),
    as.character(n_taxa_real),
    prec_slope_txt,
    fmt_num(k_real),
    fmt_num(-k_real),
    prec_int_txt,
    fmt_num(c_real),
    fmt_num(prec_sd_real),
    fmt_rng(intercept_effects),
    fmt_rng(slope_effects)
  ),
  `Simulation Value` = c(
    as.character(n_samples),
    as.character(n_taxa),
    "N/A",
    fmt_num(k_realistic),
    fmt_num(-k_realistic),
    "N/A",
    fmt_num(intercept_disp_realistic),
    fmt_num(prec_sd_real),
    fmt_rng(mu_inv_softmax_base_realistic),
    fmt_rng(slope_realistic)
  )
)

knitr::kable(param_summary, caption = "Summary of HMP2 parameters used for realistic simulation")
```

## Key Findings

The simulation uses parameters extracted directly from the HMP2 microbiome dataset sccomp fit.

### Parameter Mapping (sccomp → Simulator)

**sccomp uses precision**: `log(precision) ≈ prec_intercept + prec_slope × log(mean)` (indexed `prec_intercept_1`, `prec_slope_1` when bimodal)

**Simulator uses overdispersion (σ)**: `log(σ) = -k × log(mean) + c`

Since `precision ≈ 1/σ`, we have `log(σ) ≈ -log(precision)`, so:

- **k = prec_slope** = `r round(k_real, 4)` (the slope transfers directly)
- **c = -prec_intercept** = `r round(c_real, 4)` (the intercept is negated)

### Interpretation

With k = `r round(k_real, 4)` (negative):
- The simulator formula becomes: `log(σ) = -(`r round(k_real, 4)`) × log(μ) + c = +`r round(-k_real, 4)` × log(μ) + c`
- **Higher mean abundance → higher σ → MORE overdispersion** ✓

This matches the biological expectation that abundant taxa tend to be more variable.

### Additional Parameters

- **Variability around k (prec_sd)**: `r round(prec_sd_real, 4)` - This adds realistic variation to the simulations.
- **Baseline abundances**: Sampled from the real HMP2 intercept distribution.
- **Group effects (slopes)**: Sampled from the real HMP2 IBD vs Healthy effect distribution.

4. **Baseline abundances**: Sampled from the real distribution of intercept effects, capturing the natural variation in taxa abundances.

5. **Group effects**: Based on the real IBD vs Healthy effect size distribution.

This approach ensures the simulation conditions reflect what we observe in real microbiome studies, providing more meaningful validation of beta-diversity distance inputs for PERMDISP.
