---
title: "Zeller et al. 2014 (CRC) validation analysis"
subtitle: "Observed beta dispersion versus sccomp simulations with differential abundance switched off"
author: "Mangiola Laboratory"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    fig-width: 10
    fig-height: 8
    theme: cosmo
execute:
  warning: false
  message: false
  fig.path: "zellerg_2014_validation_figures/"
params:
  group_variable: "study_condition"
  group_levels: ["control", "CRC"]
  group_labels: ["Control", "CRC"]
  body_site_levels: ["stool"]
  n_reps_null: 40
  n_permutations_observed: 999
  n_permutations_permanova: 199
  n_permutations_null: 999
  abundance_threshold: 0.001
---

## Overview

The Anna Karenina Principle (AKP) interpretation, popularised by
Zaneveld et al. (2017), claims that diseased or perturbed states are
more variable in community composition than healthy reference states,
typically supported by PERMDISP. This report extends our central
critique to a canonical disease cohort: the Zeller et al. (2014)
metagenomic colorectal cancer (CRC) study, accessed via
`curatedMetagenomicData`. CRC versus healthy stool is one of the
most-cited perturbation contrasts where increased beta dispersion has
been claimed.

This report:

1. Reproduces an observed PERMDISP comparison in the Zeller et al. CRC
   stool data (CRC vs control).
2. Fits a generative `sccomp` model to the same contrast with
   non-interest covariates (`age_category`, `gender`).
3. Simulates matched datasets under two null scenarios:
   - **No DA, No DD**: differential abundance and group differential
     dispersion both switched off (the formal null for PERMDISP).
   - **No DA, Yes DD**: differential abundance switched off but
     fitted taxon-level group differential dispersion preserved.
4. Re-runs the same PERMDISP workflow on each set of simulations.
5. Compares the observed beta-dispersion effect with the DA-off null
   distribution.

```{r setup}
#| include: false

library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(ggplot2)
library(vegan)
library(parallelly)
library(patchwork)
library(stringr)
library(sccomp)

source("../functions.R")

knitr::opts_chunk$set(
  fig.path = "zellerg_2014_validation_figures/",
  dpi = 220
)

theme_validation <-
  theme_bw() +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(linewidth = 0.25),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.background = element_blank()
  )

# Compatibility shim for tidybulk::scale_y_log10_reverse() on ggplot2/scales
# versions that look up a named `log10_reverse` transform at render time.
log10_reverse_trans <- function() {
  scales::trans_new(
    name = "log10_reverse",
    transform = function(x) -log10(pmax(x, .Machine$double.xmin)),
    inverse = function(x) 10^(-x)
  )
}
transform_log10_reverse <- log10_reverse_trans

standard_distance_methods <- c("bray", "aitchison", "robust.aitchison", "jaccard", "hellinger")
group_levels <- unlist(params$group_levels)
group_labels <- unlist(params$group_labels)
if (length(group_levels) != 2L || length(group_labels) != 2L) {
  stop("params$group_levels and params$group_labels must both have length 2.")
}
names(group_labels) <- group_levels

sccomp_file_path <- "ZellerG_2014_sccomp_results.rds"

# Robust Stan bytecode cache under this quarto folder. The package default cache
# can live on flaky cloud-sync volumes and occasionally fails `gzfile()` when
# writing compiled models during knit.
sccomp_stan_cache_dir <- file.path(normalizePath("."), "_sccomp_stan_model_cache")
dir.create(sccomp_stan_cache_dir, showWarnings = FALSE, recursive = TRUE)
```

## Data Preparation

```{r data-preparation}
#| cache: false

suppressPackageStartupMessages({
  library(curatedMetagenomicData)
  library(TreeSummarizedExperiment)
  library(SummarizedExperiment)
  library(S4Vectors)
  library(mia)
  library(magrittr)
})

pick_existing_column <- function(dat, candidates, required = TRUE, label = "column") {
  hit <- intersect(candidates, colnames(dat))
  if (length(hit) > 0L) {
    return(hit[[1L]])
  }
  hit_lower <- match(tolower(candidates), tolower(colnames(dat)), nomatch = 0L)
  hit_lower <- hit_lower[hit_lower > 0L]
  if (length(hit_lower) > 0L) {
    return(colnames(dat)[[hit_lower[[1L]]]])
  }
  if (required) {
    stop(
      "Could not find ", label, ". Tried: ",
      paste(candidates, collapse = ", "),
      "\nAvailable columns: ",
      paste(colnames(dat), collapse = ", ")
    )
  }
  NULL
}

normalise_text <- function(x) {
  tolower(gsub("[^A-Za-z0-9]+", "_", trimws(as.character(x))))
}

as_count_long <- function(tse, assay_name) {
  count_matrix <- SummarizedExperiment::assay(tse, assay_name)
  count_matrix <- as.matrix(count_matrix)
  storage.mode(count_matrix) <- "numeric"
  if (is.null(rownames(count_matrix))) {
    rownames(count_matrix) <- paste0("taxon_", seq_len(nrow(count_matrix)))
  }
  if (is.null(colnames(count_matrix))) {
    colnames(count_matrix) <- paste0("sample_", seq_len(ncol(count_matrix)))
  }

  as.data.frame(count_matrix, check.names = FALSE) |>
    tibble::rownames_to_column("taxon_id") |>
    tidyr::pivot_longer(
      cols = -taxon_id,
      names_to = "sample_id",
      values_to = "count"
    ) |>
  dplyr::mutate(
    count = if (all(abs(.data$count - round(.data$count)) < .Machine$double.eps^0.5, na.rm = TRUE)) {
      as.integer(round(.data$count))
    } else {
      as.numeric(.data$count)
    }
  )
}

# curatedMetagenomicData returns counts when counts = TRUE, even though the
# assay slot is named "relative_abundance".
tse_raw <- curatedMetagenomicData::curatedMetagenomicData(
  "ZellerG_2014.relative_abundance",
  dryrun = FALSE,
  counts = TRUE
)[[1L]]

assay_name <- SummarizedExperiment::assayNames(tse_raw)[[1L]]
assay_preview <- as.matrix(SummarizedExperiment::assay(tse_raw, assay_name))
is_relative_abundance_assay <- max(assay_preview, na.rm = TRUE) <= 1
detection_threshold <- if (is_relative_abundance_assay) 1e-4 else 10L

metadata_raw <- as.data.frame(SummarizedExperiment::colData(tse_raw)) |>
  tibble::rownames_to_column("sample_id") |>
  dplyr::mutate(across(where(is.character), trimws))

body_site_col <- pick_existing_column(
  metadata_raw,
  c("body_site", "Sample_Type", "body_site_name", "sample_type", "SampleType", "site"),
  required = FALSE,
  label = "body-site column"
)

group_variable <- params$group_variable
if (!group_variable %in% colnames(metadata_raw)) {
  stop(
    "Requested params$group_variable = '", group_variable, "' was not found in colData.\n",
    "Available columns: ", paste(colnames(metadata_raw), collapse = ", ")
  )
}

metadata_prepared <- metadata_raw |>
  dplyr::mutate(
    group_raw = normalise_text(.data[[group_variable]]),
    group = dplyr::case_when(
      group_raw == normalise_text(group_levels[[1L]]) ~ group_levels[[1L]],
      group_raw == normalise_text(group_levels[[2L]]) ~ group_levels[[2L]],
      TRUE ~ NA_character_
    ),
    group = factor(.data$group, levels = group_levels)
  )

if (!is.null(body_site_col)) {
  body_site_keep <- normalise_text(unlist(params$body_site_levels))
  metadata_prepared <- metadata_prepared |>
    dplyr::filter(normalise_text(.data[[body_site_col]]) %in% body_site_keep)
}

metadata_prepared <- metadata_prepared |>
  dplyr::filter(!is.na(.data$group)) |>
  dplyr::mutate(group_label = factor(group_labels[as.character(.data$group)], levels = group_labels))

if (nrow(metadata_prepared) == 0L) {
  stop(
    "No ZellerG_2014 samples remained after applying group/body-site filters. ",
    "Check params$group_variable, params$group_levels, and params$body_site_levels."
  )
}

sample_keep <- metadata_prepared$sample_id
tse <- tse_raw[, sample_keep]

n_taxa_before <- nrow(tse)
tse <- mia::subsetByPrevalent(
  tse,
  detection = detection_threshold,
  prevalence = 0.10,
  assay.type = assay_name
)
n_taxa_after <- nrow(tse)

count_long <- as_count_long(tse, assay_name) |>
  dplyr::filter(.data$sample_id %in% metadata_prepared$sample_id) |>
  dplyr::group_by(.data$sample_id) |>
  dplyr::mutate(library_size = sum(.data$count, na.rm = TRUE)) |>
  dplyr::ungroup()

sample_metadata <- metadata_prepared |>
  dplyr::filter(.data$sample_id %in% unique(count_long$sample_id)) |>
  dplyr::select(sample_id, group, group_label, dplyr::everything())

count_long <- count_long |>
  dplyr::inner_join(sample_metadata |> dplyr::select(sample_id, group), by = "sample_id")

count_long_analysis <- filter_by_abundance_cases(
  count_long,
  threshold = params$abundance_threshold
)

cat("ZellerG_2014 input prepared.\n")
cat("Assay:", assay_name, "\n")
cat("Prevalence-filter detection threshold:", detection_threshold, "\n")
cat("Samples:", dplyr::n_distinct(sample_metadata$sample_id), "\n")
cat("Taxa before/after prevalence filter:", n_taxa_before, "/", n_taxa_after, "\n")
cat("Taxa after abundance filter:", dplyr::n_distinct(count_long_analysis$taxon_id), "\n")
print(sample_metadata |> dplyr::count(group_label, name = "n_samples"))
```

## Observed PERMDISP

```{r observed-permdisp}

run_betadisper_table <- function(count_long, sample_metadata, group_levels, permutations) {
  inputs <- purrr::set_names(
    lapply(standard_distance_methods, function(dm) {
      build_standard_betadisper_inputs_paper(
        count_long = count_long,
        sample_metadata = sample_metadata,
        distance_method = dm,
        group_levels = group_levels
      )
    }),
    nm = paste0("permdisp_", standard_distance_methods)
  )

  purrr::imap_dfr(inputs, function(inp, method_key) {
    bd <- vegan::betadisper(inp$dist, inp$group)
    pt <- permutest_betadisper_fixed(bd, permutations = permutations)
    dist_df <- tibble::tibble(
      group = bd$group,
      distance_to_centroid = unname(bd$distances)
    )
    group_means <- dist_df |>
      dplyr::group_by(.data$group) |>
      dplyr::summarise(mean_distance = mean(.data$distance_to_centroid), .groups = "drop")

    tibble::tibble(
      method_key = method_key,
      method = pretty_method_label(method_key),
      reference_group = group_levels[[1L]],
      focal_group = group_levels[[2L]],
      reference_mean_distance = group_means$mean_distance[group_means$group == group_levels[[1L]]],
      focal_mean_distance = group_means$mean_distance[group_means$group == group_levels[[2L]]],
      dispersion_delta = focal_mean_distance - reference_mean_distance,
      p_value = pt$p_value,
      n_perm = pt$n_perm,
      bd = list(bd),
      distances = list(dist_df)
    )
  })
}

observed_permdisp <- run_betadisper_table(
  count_long = count_long_analysis,
  sample_metadata = sample_metadata,
  group_levels = group_levels,
  permutations = params$n_permutations_observed
)

observed_permdisp |>
  dplyr::select(method, reference_mean_distance, focal_mean_distance, dispersion_delta, p_value, n_perm) |>
  dplyr::arrange(p_value) |>
  knitr::kable(digits = 4)
```

```{r observed-permdisp-plot}
#| fig-cap: "Observed ZellerG_2014 distances to group centroid. Positive deltas indicate greater beta dispersion in CRC than in controls."

observed_distances <- observed_permdisp |>
  dplyr::select(method, distances) |>
  tidyr::unnest(distances) |>
  dplyr::mutate(group_label = factor(group_labels[as.character(.data$group)], levels = group_labels))

observed_permdisp_boxplot <- ggplot(observed_distances, aes(x = group_label, y = distance_to_centroid, fill = group_label)) +
  geom_boxplot(outlier.alpha = 0.25, width = 0.65) +
  facet_wrap(~ method, scales = "free_y") +
  labs(x = NULL, y = "Distance to group centroid", fill = NULL) +
  theme_validation

observed_permdisp_boxplot

ggplot2::ggsave(
  filename = "observed_permdisp_boxplot.pdf",
  plot = observed_permdisp_boxplot,
  width = 12,
  height = 7,
  units = "in",
  device = grDevices::cairo_pdf,
  dpi = 320
)
```

## PERMANOVA Context

```{r observed-permanova}

run_permanova_table <- function(count_long, sample_metadata, group_levels, permutations) {
  purrr::map_dfr(c("bray"), function(dm) {
    inp <- build_standard_betadisper_inputs_paper(
      count_long = count_long,
      sample_metadata = sample_metadata,
      distance_method = dm,
      group_levels = group_levels
    )
    group <- inp$group
    ad <- vegan::adonis2(inp$dist ~ group, permutations = permutations)
    tibble::tibble(
      method = pretty_method_label(dm),
      r2 = ad$R2[[1L]],
      p_value = ad$`Pr(>F)`[[1L]]
    )
  })
}

observed_permanova <- run_permanova_table(
  count_long = count_long_analysis,
  sample_metadata = sample_metadata,
  group_levels = group_levels,
  permutations = params$n_permutations_permanova
)

observed_permanova |>
  dplyr::arrange(p_value) |>
  knitr::kable(digits = 4)
```

## sccomp Model

```{r fit-sccomp-zeller-if-needed}
#| cache: false

metadata_for_fit <- sample_metadata |>
  dplyr::mutate(
    sccomp_sample_id = .data$sample_id,
    group = factor(.data$group, levels = group_levels)
  )

count_wide <- count_long |>
  dplyr::select(sample_id, taxon_id, count) |>
  tidyr::pivot_wider(names_from = taxon_id, values_from = count, values_fill = 0)

sccomp_input <- count_wide |>
  tidyr::pivot_longer(
    cols = -sample_id,
    names_to = "taxon_id",
    values_to = "count"
  ) |>
  dplyr::left_join(metadata_for_fit, by = "sample_id")

candidate_covariates <- c("age_category", "gender")
covariates <- unlist(
  lapply(candidate_covariates, function(column) {
    pick_existing_column(sccomp_input, column, required = FALSE)
  }),
  use.names = FALSE
)
covariates <- unique(covariates[!is.na(covariates)])
covariates <- covariates[vapply(sccomp_input[covariates], function(x) {
  dplyr::n_distinct(x[!is.na(x)]) > 1L && dplyr::n_distinct(x[!is.na(x)]) < nrow(metadata_for_fit)
}, logical(1))]

composition_terms <- c("group", covariates)
formula_composition <- stats::as.formula(paste("~", paste(composition_terms, collapse = " + ")))
formula_variability <- ~ group

message("sccomp composition formula: ", deparse(formula_composition))
message("sccomp variability formula: ", deparse(formula_variability))

if (!file.exists(sccomp_file_path)) {
  fit <- sccomp_input |>
    sccomp::sccomp_estimate(
      formula_composition = formula_composition,
      formula_variability = formula_variability,
      sample = "sample_id",
      cell_group = "taxon_id",
      abundance = "count",
      inference_method = "hmc",
      cores = max(1, parallelly::availableCores() - 1),
      bimodal_mean_variability_association = TRUE,
      verbose = TRUE,
      max_sampling_iterations = 3000,
      sig_figs = 6,
      portable = FALSE,
      cache_stan_model = sccomp_stan_cache_dir,
      exclude_mean_variability_association = TRUE
    ) |>
    sccomp::sccomp_test(contrasts = c("Intercept" = "`(Intercept)`", paste0("group", group_levels[[2L]])))

  fit <- fit |>
    sccomp:::incorporate_parameters_into_sccomp_object(
      parameters_to_load = c(
        "prec_intercept_1",
        "prec_slope_1",
        "prec_intercept_2",
        "prec_slope_2",
        "prec_sd",
        "mix_p",
        "beta",
        "alpha"
      )
    )

  saveRDS(fit, sccomp_file_path)
} else {
  message("Using existing sccomp fit: ", sccomp_file_path)
}

fit <- readRDS(sccomp_file_path)
```

```{r sccomp-qc}
#| fig-cap: "Per-taxon (c_effect, v_effect) posterior for the original Zeller fit. Left panel: Intercept (baseline taxon-level log-odds abundance vs baseline log-precision). Right panel: groupCRC slope (CRC vs control effect on abundance vs on overdispersion). Horizontal bars show 95% CI on c_effect; vertical bars show 95% CI on v_effect. Points are coloured by per-taxon FDR (red: c_FDR < 0.05; blue: v_FDR < 0.05; purple: both; grey: neither). Note that for the Intercept panel `c_FDR < 0.05` simply flags taxa whose baseline log-abundance differs from the across-taxa geometric mean, so most rare/abundant taxa qualify; for the groupCRC panel the FDR colours flag genuine CRC-vs-control differential signals."
#| fig-width: 14
#| fig-height: 6

# Shared colour scheme + helpers, reused below for the DA-off comparison plot.
sig_colors <- c(
  "neither"          = "grey75",
  "abundance only"   = "#E41A1C",
  "variability only" = "#377EB8",
  "both"             = "#984EA3"
)

extract_taxon_data <- function(this_fit, case_label, param_name) {
  taxon_col <- attr(this_fit, ".cell_group")
  this_fit |>
    dplyr::filter(.data$parameter == param_name) |>
    dplyr::transmute(
      taxon_id = .data[[taxon_col]],
      case = case_label,
      parameter = param_name,
      c_effect, c_lower, c_upper, c_FDR,
      v_effect, v_lower, v_upper, v_FDR,
      sig_c = .data$c_FDR < 0.05,
      sig_v = .data$v_FDR < 0.05,
      significance = dplyr::case_when(
        sig_c & sig_v ~ "both",
        sig_c        ~ "abundance only",
        sig_v        ~ "variability only",
        TRUE         ~ "neither"
      )
    )
}

make_2d_plot <- function(plot_data, title_label, x_lab, y_lab,
                          n_labels_per_facet = 5L,
                          omit_ci = TRUE) {
  label_data <- plot_data |>
    dplyr::filter(.data$significance != "neither") |>
    dplyr::group_by(.data$case) |>
    dplyr::arrange(dplyr::desc(pmax(
      -log10(pmax(.data$c_FDR, 1e-300)),
      -log10(pmax(.data$v_FDR, 1e-300))
    ))) |>
    dplyr::slice_head(n = n_labels_per_facet) |>
    dplyr::ungroup()

  plot_data <- plot_data |>
    dplyr::mutate(
      significance = factor(.data$significance, levels = names(sig_colors))
    )

  p <- ggplot(plot_data, aes(x = c_effect, y = v_effect, color = significance)) +
    geom_vline(xintercept = 0, linetype = "dotted", color = "grey50") +
    geom_hline(yintercept = 0, linetype = "dotted", color = "grey50")

  if (!omit_ci) {
    p <- p +
      geom_errorbarh(aes(xmin = c_lower, xmax = c_upper),
                      height = 0, alpha = 0.5, linewidth = 0.25) +
      geom_errorbar(aes(ymin = v_lower, ymax = v_upper),
                      width = 0, alpha = 0.5, linewidth = 0.25)
  }

  p +
    geom_point(size = 0.45, alpha = 0.7) +
    ggrepel::geom_text_repel(
      data = label_data,
      aes(label = stringr::str_trunc(taxon_id, 40)),
      size = 2.4, max.overlaps = 8, segment.size = 0.2, show.legend = FALSE
    ) +
    facet_wrap(~ case, ncol = 1) +
    scale_color_manual(values = sig_colors, drop = FALSE) +
    labs(title = title_label, x = x_lab, y = y_lab, color = "FDR < 0.05") +
    theme_validation +
    theme(strip.text = element_text(face = "bold"),
          plot.title = element_text(face = "bold", hjust = 0.5))
}

slope_param <- paste0("group", group_levels[[2L]])

qc_intercept <- extract_taxon_data(fit, "Original Zeller cohort", "Intercept")
qc_slope     <- extract_taxon_data(fit, "Original Zeller cohort", slope_param)

p_qc_int <- make_2d_plot(
  qc_intercept,
  title_label = "Intercept",
  x_lab = "c_effect (baseline abundance, log-odds vs geometric mean)",
  y_lab = "v_effect (baseline overdispersion)"
)
p_qc_slope <- make_2d_plot(
  qc_slope,
  title_label = slope_param,
  x_lab = paste0("c_effect (abundance: ", group_labels[[2L]], " vs ", group_labels[[1L]], ")"),
  y_lab = paste0("v_effect (variability: ", group_labels[[2L]], " vs ", group_labels[[1L]], ")")
)

qc_2d_plot <- (p_qc_int + theme(legend.position = "none")) |
  (p_qc_slope + theme(legend.position = "bottom"))

qc_2d_plot

ggplot2::ggsave(
  filename = "sccomp_qc_2d_plot.pdf",
  plot = qc_2d_plot,
  width = 14,
  height = 6,
  units = "in",
  device = grDevices::cairo_pdf,
  dpi = 320
)
```

```{r sccomp-parameters}

sccomp_params <- extract_sccomp_params_brito(fit)
# Tell the simulator to use the observed per-class sample counts from the
# filtered cohort (instead of the default balanced floor(N/2)).
observed_n_per_group <- as.integer(table(factor(sample_metadata$group, levels = group_levels)))
stopifnot(length(observed_n_per_group) == 2L, all(observed_n_per_group > 0L))
sim_params <- build_simulation_params_brito(
  sccomp_params,
  n_reps_auc = params$n_reps_null,
  group_levels = group_levels,
  n_samples_per_group = observed_n_per_group
)
message(
  "Simulating with observed per-class sample sizes: ",
  paste(group_levels, observed_n_per_group, sep = "=", collapse = ", ")
)

parameter_summary <- tibble::tibble(
  quantity = c(
    "Samples used for simulation",
    "Taxa used for simulation",
    "Significant DA taxa available to simulator",
    "Mean library size",
    "Library-size SD",
    "Taxon-level log-sigma SD"
  ),
  value = c(
    sim_params$n_samples,
    sim_params$n_taxa,
    length(sccomp_params$slope_effects_significant),
    round(sim_params$library_size_mean, 1),
    round(sim_params$library_size_sd, 1),
    round(sim_params$sd_log_overdispersion_realistic, 4)
  )
)

parameter_summary |>
  knitr::kable()
```

## DA-Off Simulations

```{r simulate-da-off-null}

simulate_da_off_once <- function(seed, preserve_group_variability = FALSE) {
  if (preserve_group_variability) {
    set.seed(seed + 20000)
    paired_idx <- sample.int(
      length(sccomp_params$intercept_effects),
      size = sim_params$n_taxa,
      replace = TRUE
    )
    composition_intercepts <- sccomp_params$intercept_effects[paired_idx]
    composition_intercepts <- composition_intercepts - mean(composition_intercepts)
    dispersion_group_0 <- sccomp_params$alpha_intercept_effects[paired_idx]
    dispersion_group_1 <- dispersion_group_0 + sccomp_params$v_slope[paired_idx]
    intercept_dispersion_matrix <- rbind(dispersion_group_0, dispersion_group_1)

    sim_r <- simulate_case_realistic_taxa(
      slope_vector = rep(0, sim_params$n_taxa),
      mu_inv_softmax = composition_intercepts,
      log_dispersion_assoc = 0,
      intercept_dispersion_matrix = intercept_dispersion_matrix,
      sd_log_overdispersion = 0,
      seed = seed,
      n_taxa = sim_params$n_taxa,
      n_samples_per_group = sim_params$n_samples_per_group,
      library_size_mean = sim_params$library_size_mean,
      library_size_sd = sim_params$library_size_sd
    )
    sim_r <- synthetic_sim_add_two_group_factor(sim_r, group_levels)
    sim_r$seed <- seed
    sim_r$n_da_taxa <- 0L
    return(sim_r)
  }

  run_simulation_job_brito(
    job_row = tibble::tibble(seed = seed, n_da_taxa = 0L),
    composition_intercept_by_taxon = sim_params$mu_inv_softmax_base_realistic,
    dispersion_intercept_by_taxon = sim_params$intercept_disp_realistic_taxon,
    mean_dispersion_assoc_slope = 0,
    slope_effects_distribution = sccomp_params$slope_effects,
    sd_log_overdispersion = sim_params$sd_log_overdispersion_realistic,
    n_taxa = sim_params$n_taxa,
    n_samples = sim_params$n_samples,
    n_samples_per_group = sim_params$n_samples_per_group,
    library_size_mean = sim_params$library_size_mean,
    library_size_sd = sim_params$library_size_sd,
    group_levels = group_levels
  )
}

analyse_da_off_once <- function(seed, preserve_group_variability = FALSE) {
  sim_r <- simulate_da_off_once(seed, preserve_group_variability = preserve_group_variability)
  sim_r$count_long <- filter_by_abundance_cases(
    sim_r$count_long,
    threshold = params$abundance_threshold
  )
  result <- run_analysis_job(
    sim_r = sim_r,
    max_perm = params$n_permutations_null,
    batch = params$n_permutations_null,
    target_se = 1e-9,
    return_full = TRUE
  )

  purrr::imap_dfr(result$method_results, function(mr, method_key) {
    tibble::tibble(
      seed = seed,
      method_key = method_key,
      method = pretty_method_label(method_key),
      reference_mean_distance = mr$group1,
      focal_mean_distance = mr$group2,
      dispersion_delta = mr$delta,
      p_value = mr$p_value,
      n_perm = mr$n_perm,
      simulation = if (preserve_group_variability) {
        "No DA, Yes DD"
      } else {
        "No DA, No DD"
      }
    )
  })
}

null_seeds <- sim_params$rep_seeds_auc
null_results <- dplyr::bind_rows(
  lapply(null_seeds, analyse_da_off_once, preserve_group_variability = FALSE),
  lapply(null_seeds, analyse_da_off_once, preserve_group_variability = TRUE)
)

null_results |>
  dplyr::count(simulation, method, significant = .data$p_value < 0.05) |>
  tidyr::complete(
    simulation,
    method,
    significant = c(FALSE, TRUE),
    fill = list(n = 0L)
  ) |>
  tidyr::pivot_wider(names_from = significant, values_from = n, values_fill = 0) |>
  dplyr::rename(n_not_significant = `FALSE`, n_significant = `TRUE`) |>
  dplyr::mutate(false_positive_rate = n_significant / (n_significant + n_not_significant)) |>
  knitr::kable(digits = 3)
```

## PCoA: Original vs No DA, Yes DD

```{r pcoa-original-vs-no-da-yes-dd}
#| fig-cap: "Bray-Curtis PCoA for the observed Zeller cohort (left) and a representative DA-off simulation with group variability retained (No DA, Yes DD; right). Each point is a sample; large points mark group centroids; faint segments connect samples to their group centroid."
#| fig-width: 14
#| fig-height: 6

build_pcoa_plot_case <- function(count_long_case, sample_metadata_case, case_label) {
  inp <- build_standard_betadisper_inputs_paper(
    count_long = count_long_case,
    sample_metadata = sample_metadata_case,
    distance_method = "bray",
    group_levels = group_levels
  )

  pcoa <- stats::cmdscale(inp$dist, k = 2, eig = TRUE)
  sites <- as.data.frame(pcoa$points)
  colnames(sites) <- c("Axis1", "Axis2")
  sites$sample_id <- rownames(sites)

  md <- sample_metadata_case |>
    dplyr::mutate(sample_id = as.character(.data$sample_id)) |>
    dplyr::distinct(.data$sample_id, .data$group)

  sites <- sites |>
    dplyr::mutate(sample_id = as.character(.data$sample_id)) |>
    dplyr::left_join(md, by = "sample_id") |>
    dplyr::mutate(
      group = factor(.data$group, levels = group_levels),
      group_label = factor(group_labels[as.character(.data$group)], levels = group_labels)
    )

  centroids <- sites |>
    dplyr::group_by(.data$group_label) |>
    dplyr::summarise(
      cx = mean(.data$Axis1, na.rm = TRUE),
      cy = mean(.data$Axis2, na.rm = TRUE),
      .groups = "drop"
    )

  seg_df <- sites |>
    dplyr::left_join(centroids, by = "group_label")

  hull_df <- sites |>
    dplyr::group_by(.data$group_label) |>
    dplyr::group_modify(\(d, ...) {
      if (nrow(d) < 3) return(d[0, , drop = FALSE])
      idx <- grDevices::chull(d$Axis1, d$Axis2)
      d[idx, , drop = FALSE] |>
        dplyr::slice(c(seq_len(n()), 1))
    }) |>
    dplyr::ungroup()

  eigvals <- pcoa$eig
  pct <- 100 * eigvals / sum(pmax(eigvals, 0), na.rm = TRUE)
  pc1_lab <- sprintf("PCoA 1 (%.1f%%)", pct[[1L]])
  pc2_lab <- sprintf("PCoA 2 (%.1f%%)", pct[[2L]])

  # Publication palette (purple/orange) for high contrast on white background.
  pcoa_group_cols <- stats::setNames(
    c("#5E3C99", "#D17A22"),
    group_labels
  )

  ggplot(sites, aes(x = Axis1, y = Axis2)) +
    geom_polygon(
      data = hull_df,
      aes(x = Axis1, y = Axis2, fill = group_label, group = group_label),
      alpha = 0.14,
      color = NA,
      inherit.aes = FALSE
    ) +
    geom_segment(
      data = seg_df,
      aes(x = cx, y = cy, xend = Axis1, yend = Axis2, color = group_label),
      alpha = 0.22, linewidth = 0.2
    ) +
    scale_color_manual(values = pcoa_group_cols, drop = FALSE) +
    scale_fill_manual(values = pcoa_group_cols, drop = FALSE, guide = "none") +
    labs(
      title = case_label,
      x = pc1_lab,
      y = pc2_lab,
      color = NULL
    ) +
    theme_validation +
    theme(plot.title = element_text(face = "bold", hjust = 0.5))
}

pcoa_original <- build_pcoa_plot_case(
  count_long_case = count_long_analysis,
  sample_metadata_case = sample_metadata,
  case_label = "Original"
)

representative_seed <- sim_params$rep_seeds_auc[[1L]]
sim_retained_pcoa <- simulate_da_off_once(
  seed = representative_seed,
  preserve_group_variability = TRUE
)
sim_retained_pcoa$count_long <- filter_by_abundance_cases(
  sim_retained_pcoa$count_long,
  threshold = params$abundance_threshold
)

pcoa_no_da_yes_dd <- build_pcoa_plot_case(
  count_long_case = sim_retained_pcoa$count_long,
  sample_metadata_case = sim_retained_pcoa$sample_metadata,
  case_label = "Simulated: No DA, Yes DD"
)

pcoa_pair_plot <- pcoa_original | pcoa_no_da_yes_dd
pcoa_pair_plot

ggplot2::ggsave(
  filename = "pcoa_original_vs_no_da_yes_dd.pdf",
  plot = pcoa_pair_plot,
  width = 14,
  height = 6,
  units = "in",
  device = grDevices::cairo_pdf,
  dpi = 320
)
```

## UMAP: Sample Composition (Filtered Taxa)

```{r umap-original-vs-no-da-yes-dd}
#| fig-cap: "UMAP of sample composition for filtered taxa in Original and representative No DA, Yes DD cohorts. Rows show preprocessing variants: raw proportions, arcsin-sqrt transformed proportions, and arcsin-sqrt transformed proportions with per-taxon z-scaling to reduce the dominance of highly abundant taxa."
#| fig-width: 14
#| fig-height: 10

build_umap_case <- function(count_long_case, sample_metadata_case, case_label, transform_label) {
  md <- sample_metadata_case |>
    dplyr::mutate(sample_id = as.character(.data$sample_id)) |>
    dplyr::distinct(.data$sample_id, .data$group)

  count_matrix <- count_long_case |>
    dplyr::mutate(
      sample_id = as.character(.data$sample_id),
      taxon_id = as.character(.data$taxon_id),
      count = as.numeric(.data$count)
    ) |>
    dplyr::group_by(.data$sample_id, .data$taxon_id) |>
    dplyr::summarise(count = sum(.data$count, na.rm = TRUE), .groups = "drop") |>
    tidyr::pivot_wider(
      names_from = .data$taxon_id,
      values_from = .data$count,
      values_fill = 0
    ) |>
    tibble::column_to_rownames("sample_id") |>
    as.matrix()

  sample_totals <- rowSums(count_matrix)
  feature_mat <- sweep(count_matrix, 1, pmax(sample_totals, 1), "/")
  sample_ids <- rownames(feature_mat)

  # Enforce strict numeric matrix format for uwot.
  if (!is.matrix(feature_mat)) {
    feature_mat <- matrix(feature_mat, ncol = 1L)
  }
  storage.mode(feature_mat) <- "double"
  feature_mat[!is.finite(feature_mat)] <- 0

  # UMAP needs at least 2 feature columns to be informative/stable.
  if (ncol(feature_mat) < 2L) {
    feature_mat <- cbind(feature_mat, rep(0, nrow(feature_mat)))
  }

  clip01 <- function(x) {
    x[x < 0] <- 0
    x[x > 1] <- 1
    x
  }

  feature_mat <- switch(
    transform_label,
    "Raw proportions" = feature_mat,
    "Arcsin-sqrt proportions" = {
      x <- clip01(feature_mat)
      asin(sqrt(x))
    },
    "Arcsin-sqrt + per-taxon scaling" = {
      x <- clip01(feature_mat)
      x <- asin(sqrt(x))
      x <- scale(x, center = TRUE, scale = TRUE)
      x[!is.finite(x)] <- 0
      x
    },
    stop("Unknown transform label: ", transform_label)
  )

  feature_mat <- as.matrix(feature_mat)
  feature_mat[!is.finite(feature_mat)] <- 0

  # Drop zero-variance taxa columns to stabilise UMAP initialization.
  keep_var <- apply(feature_mat, 2, stats::sd, na.rm = TRUE) > 0
  if (any(keep_var)) {
    feature_mat <- feature_mat[, keep_var, drop = FALSE]
  }
  if (ncol(feature_mat) < 2L) {
    feature_mat <- cbind(feature_mat, stats::rnorm(nrow(feature_mat), sd = 1e-6))
  }

  set.seed(
    100 + match(
      transform_label,
      c(
        "Raw proportions",
        "Arcsin-sqrt proportions",
        "Arcsin-sqrt + per-taxon scaling"
      )
    ) +
      ifelse(case_label == "Original", 0, 10)
  )

  um <- try(
    uwot::umap(
      feature_mat,
      n_neighbors = 15,
      min_dist = 0.2,
      metric = "euclidean",
      n_components = 2,
      init = "random",
      ret_model = FALSE,
      n_threads = max(1, parallelly::availableCores() - 1),
      verbose = FALSE
    ),
    silent = TRUE
  )
  if (inherits(um, "try-error")) {
    um_cfg <- umap::umap.defaults
    um_cfg$n_neighbors <- 15
    um_cfg$min_dist <- 0.2
    um_cfg$metric <- "euclidean"
    um_cfg$n_components <- 2
    um_cfg$random_state <- 101
    um_cfg$verbose <- FALSE
    um <- umap::umap(feature_mat, config = um_cfg)$layout
  }
  um_vec <- suppressWarnings(as.numeric(unlist(um)))
  if (length(um_vec) %% length(sample_ids) == 0) {
    um <- matrix(um_vec, nrow = length(sample_ids), byrow = FALSE)
  } else {
    um <- as.matrix(um)
  }
  storage.mode(um) <- "double"
  if (!is.matrix(um) || nrow(um) != length(sample_ids) || ncol(um) != 2) {
    um_cfg <- umap::umap.defaults
    um_cfg$n_neighbors <- 15
    um_cfg$min_dist <- 0.2
    um_cfg$metric <- "euclidean"
    um_cfg$n_components <- 2
    um_cfg$random_state <- 101
    um_cfg$verbose <- FALSE
    um <- umap::umap(feature_mat, config = um_cfg)$layout |>
      as.matrix()
  }
  um_vec <- suppressWarnings(as.numeric(unlist(um)))
  if (length(um_vec) %% length(sample_ids) == 0) {
    um <- matrix(um_vec, nrow = length(sample_ids), byrow = FALSE)
  } else {
    um <- as.matrix(um)
  }
  storage.mode(um) <- "double"
  if (!is.matrix(um) || nrow(um) != length(sample_ids) || ncol(um) != 2) {
    # Rare backend failure mode: fallback to a stable 2D projection so the
    # report remains renderable.
    pc <- stats::prcomp(feature_mat, center = TRUE, scale. = TRUE)
    um <- pc$x[, 1:2, drop = FALSE]
  }

  if (length(um[, 1]) != length(sample_ids) || length(um[, 2]) != length(sample_ids)) {
    stop(
      "UMAP embedding size mismatch [", case_label, " / ", transform_label, "]: sample_n=",
      length(sample_ids),
      "; nrow=", nrow(um),
      "; ncol=", ncol(um),
      "; len_col1=", length(um[, 1]),
      "; class_col1=", class(um[, 1])[1]
    )
  }

  out <- tibble::tibble(
    sample_id = sample_ids,
    UMAP1 = um[, 1],
    UMAP2 = um[, 2]
  ) |>
    dplyr::left_join(md, by = "sample_id") |>
    dplyr::mutate(
      case = case_label,
      transform = transform_label,
      group_label = factor(group_labels[as.character(.data$group)], levels = group_labels)
    )

  hull <- out |>
    dplyr::group_by(.data$group_label) |>
    dplyr::group_modify(\(d, ...) {
      if (nrow(d) < 3) return(d[0, , drop = FALSE])
      idx <- grDevices::chull(d$UMAP1, d$UMAP2)
      d[idx, , drop = FALSE] |>
        dplyr::slice(c(seq_len(n()), 1))
    }) |>
    dplyr::ungroup() |>
    dplyr::mutate(case = case_label, transform = transform_label)

  list(points = out, hull = hull)
}

umap_transforms <- c(
  "Raw proportions",
  "Arcsin-sqrt proportions",
  "Arcsin-sqrt + per-taxon scaling"
)

umap_case_results <- c(
  lapply(umap_transforms, function(tr) {
    build_umap_case(
      count_long_case = count_long_analysis,
      sample_metadata_case = sample_metadata,
      case_label = "Original",
      transform_label = tr
    )
  }),
  lapply(umap_transforms, function(tr) {
    build_umap_case(
      count_long_case = sim_retained_pcoa$count_long,
      sample_metadata_case = sim_retained_pcoa$sample_metadata,
      case_label = "Simulated: No DA, Yes DD",
      transform_label = tr
    )
  })
)

umap_points <- dplyr::bind_rows(lapply(umap_case_results, `[[`, "points")) |>
  dplyr::mutate(
    case = factor(.data$case, levels = c("Original", "Simulated: No DA, Yes DD")),
    transform = factor(.data$transform, levels = umap_transforms)
  )
umap_hulls <- dplyr::bind_rows(lapply(umap_case_results, `[[`, "hull")) |>
  dplyr::mutate(
    case = factor(.data$case, levels = c("Original", "Simulated: No DA, Yes DD")),
    transform = factor(.data$transform, levels = umap_transforms)
  )

umap_group_cols <- stats::setNames(
  c("#5E3C99", "#D17A22"),
  group_labels
)

umap_comparison_plot <- ggplot(umap_points, aes(x = UMAP1, y = UMAP2, color = group_label)) +
  geom_polygon(
    data = umap_hulls,
    aes(
      x = UMAP1,
      y = UMAP2,
      fill = group_label,
      group = interaction(group_label, case, transform)
    ),
    alpha = 0.14,
    color = NA,
    inherit.aes = FALSE
  ) +
  geom_point(size = 1.6, alpha = 0.75) +
  facet_grid(transform ~ case, scales = "free") +
  scale_color_manual(values = umap_group_cols, drop = FALSE) +
  scale_fill_manual(values = umap_group_cols, drop = FALSE, guide = "none") +
  labs(x = "UMAP 1", y = "UMAP 2", color = NULL) +
  theme_validation

umap_comparison_plot

ggplot2::ggsave(
  filename = "umap_sample_composition_comparison.pdf",
  plot = umap_comparison_plot,
  width = 14,
  height = 10,
  units = "in",
  device = grDevices::cairo_pdf,
  dpi = 320
)
```

## sccomp 2D Diagnostics: Original vs DA-Off Refits

For each taxon, sccomp returns the posterior of the **composition**
effect (`c_effect`, mean abundance) and the **variability** effect
(`v_effect`, overdispersion) for the `groupCRC` slope. We plot one
panel per case (Original / No DA, No DD / No DA, Yes DD), each showing
`c_effect` (x) vs `v_effect` (y) with their 95% credible intervals.

We deliberately do **not** use `sccomp_plot_intervals_2D` here. That
function unconditionally splits each parameter into `raw` and
`adjusted` facets ; for readability we plot the posterior summaries
`(c_effect, v_effect)` in a single panel per coefficient.

Every `sccomp_estimate` fit in this report passes
`exclude_mean_variability_association = TRUE`. This replaces the older
misleading `exclude_priors` knob and excludes the inferred
mean-variability coupling during fitting (distinct from shrinking the rest
of the prior stack). For this benchmarking figure we want the plotted
effects read directly off the posterior table, without an extra facet
split.

```{r refit-sccomp-on-simulated}
#| cache: false

sccomp_simulated_fits_path <- "ZellerG_2014_sccomp_simulated_fits.rds"

fit_sccomp_on_sim <- function(sim_r, sim_label) {
  sim_input <- sim_r$count_long |>
    dplyr::transmute(
      sample_id = as.character(.data$sample_id),
      taxon_id  = as.character(.data$taxon_id),
      count     = as.integer(round(.data$count)),
      group     = factor(.data$group, levels = group_levels)
    )

  message("Refitting sccomp on simulated cohort: ", sim_label)

  sim_input |>
    sccomp::sccomp_estimate(
      formula_composition = ~ group,
      formula_variability = ~ group,
      sample = "sample_id",
      cell_group = "taxon_id",
      abundance = "count",
      inference_method = "hmc",
      cores = max(1, parallelly::availableCores() - 1),
      bimodal_mean_variability_association = TRUE,
      verbose = FALSE,
      max_sampling_iterations = 3000,
      sig_figs = 6,
      portable = FALSE,
      cache_stan_model = sccomp_stan_cache_dir,
      exclude_mean_variability_association = TRUE
    ) |>
    sccomp::sccomp_test(contrasts = c(
      "Intercept" = "`(Intercept)`",
      paste0("group", group_levels[[2L]])
    ))
}

if (!file.exists(sccomp_simulated_fits_path)) {
  representative_seed <- sim_params$rep_seeds_auc[[1L]]
  sim_no_dd  <- simulate_da_off_once(representative_seed, preserve_group_variability = FALSE)
  sim_yes_dd <- simulate_da_off_once(representative_seed, preserve_group_variability = TRUE)

  fit_no_dd  <- fit_sccomp_on_sim(sim_no_dd,  "No DA, No DD")
  fit_yes_dd <- fit_sccomp_on_sim(sim_yes_dd, "No DA, Yes DD")

  saveRDS(
    list(seed = representative_seed, fit_no_dd = fit_no_dd, fit_yes_dd = fit_yes_dd),
    sccomp_simulated_fits_path
  )
} else {
  message("Using existing sccomp fits on simulated cohorts: ",
          sccomp_simulated_fits_path)
}

sccomp_simulated_fits <- readRDS(sccomp_simulated_fits_path)
```

```{r sccomp-2d-comparison}
#| fig-cap: "Per-taxon (c_effect, v_effect) posterior for the original Zeller cohort and for sccomp refits on a representative DA-off simulated cohort. Left column: Intercept (baseline taxon log-odds abundance vs baseline log-precision). Right column: groupCRC slope (CRC vs control effect on abundance vs on overdispersion). Horizontal bars: 95% CI on c_effect; vertical bars: 95% CI on v_effect. Colours: c_FDR < 0.05 = red, v_FDR < 0.05 = blue, both = purple, neither = grey. Reading guide: the Intercept column should look essentially identical across rows because the simulator re-uses the empirical intercept distribution; the groupCRC column should show DA collapse (c_effect near zero) in both DA-off cases, with vertical (v_effect) spread retained in the No DA, Yes DD case relative to No DA, No DD."
#| fig-width: 14
#| fig-height: 16

fits_with_labels <- list(
  list(fit = fit,                              label = "Original Zeller cohort"),
  list(fit = sccomp_simulated_fits$fit_no_dd,  label = "Simulated: No DA, No DD"),
  list(fit = sccomp_simulated_fits$fit_yes_dd, label = "Simulated: No DA, Yes DD")
)

case_levels <- vapply(fits_with_labels, `[[`, character(1), "label")

bind_param_data <- function(param_name) {
  purrr::map_dfr(fits_with_labels, function(x) {
    extract_taxon_data(x$fit, x$label, param_name)
  }) |>
    dplyr::mutate(case = factor(.data$case, levels = case_levels))
}

intercept_data <- bind_param_data("Intercept")
slope_data     <- bind_param_data(slope_param)

p_int <- make_2d_plot(
  intercept_data,
  title_label = "Intercept",
  x_lab = "c_effect (baseline abundance, log-odds vs geometric mean)",
  y_lab = "v_effect (baseline overdispersion)"
)
p_slope <- make_2d_plot(
  slope_data,
  title_label = slope_param,
  x_lab = paste0("c_effect (abundance: ", group_labels[[2L]], " vs ", group_labels[[1L]], ")"),
  y_lab = paste0("v_effect (variability: ", group_labels[[2L]], " vs ", group_labels[[1L]], ")")
)

sccomp_2d_comparison_plot <- (p_int + theme(legend.position = "none")) |
  (p_slope + theme(legend.position = "bottom"))

sccomp_2d_comparison_plot

ggplot2::ggsave(
  filename = "sccomp_2d_comparison_plot.pdf",
  plot = sccomp_2d_comparison_plot,
  width = 14,
  height = 16,
  units = "in",
  device = grDevices::cairo_pdf,
  dpi = 320
)
```

## Observed Versus DA-Off

```{r observed-vs-da-off-table}

comparison_table <- observed_permdisp |>
  dplyr::select(method, observed_delta = dispersion_delta, observed_p = p_value) |>
  dplyr::left_join(
    null_results |>
      dplyr::group_by(simulation, method) |>
      dplyr::summarise(
        null_median_delta = median(.data$dispersion_delta, na.rm = TRUE),
        null_q025_delta = quantile(.data$dispersion_delta, 0.025, na.rm = TRUE),
        null_q975_delta = quantile(.data$dispersion_delta, 0.975, na.rm = TRUE),
        null_p_lt_0_05 = mean(.data$p_value < 0.05, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "method"
  ) |>
  dplyr::mutate(
    observed_outside_null_interval = .data$observed_delta < .data$null_q025_delta |
      .data$observed_delta > .data$null_q975_delta
  ) |>
  dplyr::arrange(simulation, method)

comparison_table |>
  knitr::kable(digits = 4)
```

```{r observed-vs-da-off-plot}
#| fig-cap: "Original ZellerG_2014 PERMDISP delta plotted with two DA-off sccomp simulation controls. The 'No DA, No DD' null removes both group differential abundance and group differential dispersion; the 'No DA, Yes DD' control removes group differential abundance but preserves taxon-specific group differential dispersion."

delta_plot_levels <- c(
  "Original",
  "No DA, No DD",
  "No DA, Yes DD"
)
delta_plot_colors <- c(
  "Original" = "#D55E00",
  "No DA, No DD" = "#0072B2",
  "No DA, Yes DD" = "#009E73"
)

observed_delta_results <- observed_permdisp |>
  dplyr::transmute(
    method,
    simulation = "Original",
    dispersion_delta,
    p_value
  )

delta_plot_data <- dplyr::bind_rows(
  observed_delta_results,
  null_results |>
    dplyr::select(method, simulation, dispersion_delta, p_value)
) |>
  dplyr::mutate(simulation = factor(.data$simulation, levels = delta_plot_levels))

observed_vs_da_off_boxplot <- ggplot(delta_plot_data, aes(x = simulation, y = dispersion_delta, color = simulation)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey45") +
  geom_boxplot(
    data = delta_plot_data |> dplyr::filter(.data$simulation != "Original"),
    aes(fill = simulation),
    alpha = 0.18,
    outlier.shape = NA,
    width = 0.55
  ) +
  geom_jitter(
    data = delta_plot_data |> dplyr::filter(.data$simulation != "Original"),
    width = 0.10,
    height = 0,
    alpha = 0.55,
    size = 1.6
  ) +
  geom_point(
    data = delta_plot_data |> dplyr::filter(.data$simulation == "Original"),
    size = 3.2
  ) +
  facet_wrap(~ method, scales = "free_y") +
  scale_color_manual(values = delta_plot_colors, drop = FALSE) +
  scale_fill_manual(values = delta_plot_colors, drop = FALSE) +
  labs(
    x = "Case",
    y = paste0("PERMDISP delta: ", group_labels[[2L]], " - ", group_labels[[1L]]),
    color = NULL,
    fill = NULL
  ) +
  theme_validation

observed_vs_da_off_boxplot

ggplot2::ggsave(
  filename = "observed_vs_da_off_boxplot.pdf",
  plot = observed_vs_da_off_boxplot,
  width = 12,
  height = 7,
  units = "in",
  device = grDevices::cairo_pdf,
  dpi = 320
)
```

```{r pvalue-comparison-plot}
#| fig-cap: "Original ZellerG_2014 PERMDISP p-values plotted with DA-off simulation p-values. The dotted line marks the nominal 0.05 threshold."

pvalue_plot_data <- delta_plot_data |>
  dplyr::select(method, simulation, p_value)

pvalue_comparison_boxplot <- ggplot(pvalue_plot_data, aes(x = simulation, y = p_value, color = simulation)) +
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "grey45") +
  geom_boxplot(
    data = pvalue_plot_data |> dplyr::filter(.data$simulation != "Original"),
    aes(fill = simulation),
    alpha = 0.18,
    outlier.shape = NA,
    width = 0.55
  ) +
  geom_jitter(
    data = pvalue_plot_data |> dplyr::filter(.data$simulation != "Original"),
    width = 0.10,
    height = 0,
    alpha = 0.55,
    size = 1.6
  ) +
  geom_point(
    data = pvalue_plot_data |> dplyr::filter(.data$simulation == "Original"),
    size = 3.2
  ) +
  facet_wrap(~ method) +
  tidybulk::scale_y_log10_reverse() +
  scale_color_manual(values = delta_plot_colors, drop = FALSE) +
  scale_fill_manual(values = delta_plot_colors, drop = FALSE) +
  labs(x = "Case", y = "PERMDISP p-value", color = NULL, fill = NULL) +
  theme_validation

pvalue_comparison_boxplot

ggplot2::ggsave(
  filename = "pvalue_comparison_boxplot.pdf",
  plot = pvalue_comparison_boxplot,
  width = 12,
  height = 7,
  units = "in",
  device = grDevices::cairo_pdf,
  dpi = 320
)
```

```{r proportion-whisker-three-cases}
#| fig-cap: "Taxon proportion distributions for the Original ZellerG_2014 data and representative DA-off sccomp simulations (No DA, No DD; No DA, Yes DD). Points show mean proportions; whiskers show 95% sample intervals. The y-axis uses an arcsin-square-root transform."

arcsin_sqrt_trans_proportion <- scales::trans_new(
  name = "arcsin_sqrt",
  transform = function(x) asin(sqrt(pmax(0, pmin(1, x)))),
  inverse = function(x) sin(x)^2
)

representative_seed <- sim_params$rep_seeds_auc[[1L]]

sim_matched_rep <- simulate_da_off_once(
  seed = representative_seed,
  preserve_group_variability = FALSE
)
sim_matched_rep$count_long <- filter_by_abundance_cases(
  sim_matched_rep$count_long,
  threshold = params$abundance_threshold
)

sim_retained_rep <- simulate_da_off_once(
  seed = representative_seed,
  preserve_group_variability = TRUE
)
sim_retained_rep$count_long <- filter_by_abundance_cases(
  sim_retained_rep$count_long,
  threshold = params$abundance_threshold
)

build_whisker_case <- function(count_long_case, case_label, n_top_taxa = 20L) {
  count_long_case <- count_long_case |>
    dplyr::mutate(
      group_label = factor(group_labels[as.character(.data$group)], levels = group_labels),
      proportion = .data$count / .data$library_size
    )

  taxon_order <- count_long_case |>
    dplyr::group_by(.data$taxon_id) |>
    dplyr::summarise(mean_prop = mean(.data$proportion, na.rm = TRUE), .groups = "drop") |>
    dplyr::arrange(dplyr::desc(.data$mean_prop)) |>
    dplyr::slice_head(n = n_top_taxa) |>
    dplyr::mutate(
      taxon_rank = dplyr::row_number(),
      taxon_display = paste0("T", .data$taxon_rank)
    ) |>
    dplyr::select(.data$taxon_id, .data$taxon_rank, .data$taxon_display)

  count_long_case |>
    dplyr::inner_join(taxon_order, by = "taxon_id") |>
    dplyr::mutate(case = case_label) |>
    dplyr::group_by(.data$case, .data$group_label, .data$taxon_display, .data$taxon_rank) |>
    dplyr::summarise(
      mean_prop = mean(.data$proportion, na.rm = TRUE),
      p_low = stats::quantile(.data$proportion, 0.025, na.rm = TRUE),
      p_high = stats::quantile(.data$proportion, 0.975, na.rm = TRUE),
      .groups = "drop"
    )
}

proportion_whisker_data <- dplyr::bind_rows(
  build_whisker_case(count_long_analysis, "Original"),
  build_whisker_case(sim_matched_rep$count_long, "No DA, No DD"),
  build_whisker_case(sim_retained_rep$count_long, "No DA, Yes DD")
) |>
  dplyr::mutate(
    case = factor(.data$case, levels = delta_plot_levels),
    taxon_display = factor(.data$taxon_display, levels = paste0("T", 1:20))
  )

prop_breaks_whisker <- c(0.001, 0.005, 0.01, 0.05, 0.1, 0.2, 0.5)
prop_breaks_whisker <- prop_breaks_whisker[
  prop_breaks_whisker <= max(proportion_whisker_data$p_high, na.rm = TRUE) * 1.1
]

ggplot(
  proportion_whisker_data,
  aes(x = taxon_display, y = mean_prop, ymin = p_low, ymax = p_high, color = group_label)
) +
  geom_pointrange(
    position = position_dodge(width = 0.55),
    linewidth = 0.35,
    fatten = 1.5
  ) +
  facet_wrap(~ case, ncol = 1) +
  scale_y_continuous(
    trans = arcsin_sqrt_trans_proportion,
    breaks = prop_breaks_whisker,
    labels = scales::label_percent(accuracy = 0.1)
  ) +
  labs(
    x = "Taxon rank within case (top 20 mean proportion)",
    y = "Proportion (arcsin-sqrt scale, 95% interval)",
    color = NULL
  ) +
  theme_validation +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

## Interpretation Template

```{r interpretation-template}
#| results: asis

primary_simulation <- "No DA, No DD"
primary_summary <- comparison_table |>
  dplyr::filter(.data$simulation == primary_simulation) |>
  dplyr::arrange(.data$observed_p)

cat(
  "In the observed ZellerG_2014 data, the strongest PERMDISP signal was observed for ",
  primary_summary$method[[1L]],
  " (delta = ",
  round(primary_summary$observed_delta[[1L]], 4),
  ", p = ",
  signif(primary_summary$observed_p[[1L]], 3),
  "). After refitting the cohort with sccomp and simulating matched datasets with group differential abundance switched off, the corresponding DA-off null median delta was ",
  round(primary_summary$null_median_delta[[1L]], 4),
  " with a 95% simulation interval of [",
  round(primary_summary$null_q025_delta[[1L]], 4),
  ", ",
  round(primary_summary$null_q975_delta[[1L]], 4),
  "].\n\n",
  sep = ""
)

cat(
  "This is the manuscript-ready contrast: PERMDISP detects differential beta dispersion in the actual ZellerG_2014 CRC vs control comparison, but the same analysis applied to sccomp simulations where DA is neutralized provides the null scale expected from fitted abundance and overdispersion alone."
)
```

## Final Composite Patchwork Figure

```{r final-composite-patchwork}
#| fig-cap: "Final composite panel combining sccomp 2D summaries, PCoA hull plots, and Bray-Curtis DA-off null comparisons."
#| fig-width: 12
#| fig-height: 7

compact_case_levels <- c("Original Zeller cohort", "Simulated: No DA, Yes DD")
theme_patchwork_compact <- theme_validation +
  theme(
    axis.title.y = element_text(size = 7, margin = margin(0, 0, 0, 0)),
    axis.title.x = element_text(size = 7, margin = margin(0, 0, 0, 0)),
    axis.text.x = element_text(size = 6),
    axis.text.y = element_text(size = 6),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 6),
    plot.title = element_text(size = 8, face = "bold", hjust = 0.5),
    panel.spacing.x = grid::unit(0.1, "lines"),
    panel.spacing.y = grid::unit(0.1, "lines")
  )

intercept_data_compact <- intercept_data |>
  dplyr::filter(.data$case %in% compact_case_levels) |>
  dplyr::mutate(case = factor(.data$case, levels = compact_case_levels))
slope_data_compact <- slope_data |>
  dplyr::filter(.data$case %in% compact_case_levels) |>
  dplyr::mutate(case = factor(.data$case, levels = compact_case_levels))

p_int_compact <- make_2d_plot(
  intercept_data_compact,
  title_label = "Intercept",
  x_lab = "c_effect",
  y_lab = "v_effect",
  n_labels_per_facet = 0L
)
p_slope_compact <- make_2d_plot(
  slope_data_compact,
  title_label = slope_param,
  x_lab = "c_effect",
  y_lab = "v_effect",
  n_labels_per_facet = 0L
)

sccomp_compact_plot <- (p_int_compact + theme(legend.position = "none")) |
  (p_slope_compact + theme(legend.position = "bottom"))

bray_method_label <- delta_plot_data |>
  dplyr::distinct(.data$method) |>
  dplyr::filter(stringr::str_detect(stringr::str_to_lower(.data$method), "bray")) |>
  dplyr::slice_head(n = 1) |>
  dplyr::pull(.data$method)

if (length(bray_method_label) == 0) {
  stop("Could not find a Bray-Curtis method label in delta_plot_data.")
}

bray_delta_plot <- delta_plot_data |>
  dplyr::filter(.data$method == bray_method_label) |>
  dplyr::mutate(
    simulation = factor(
      as.character(.data$simulation),
      levels = c("No DA, No DD", "No DA, Yes DD", "Original")
    )
  )
bray_delta_non_original <- bray_delta_plot |>
  dplyr::filter(.data$simulation != "Original")
bray_delta_original <- bray_delta_plot |>
  dplyr::filter(.data$simulation == "Original")

bray_delta_plot <- bray_delta_plot |>
  ggplot(aes(x = simulation, y = dispersion_delta)) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "grey45") +
  geom_boxplot(
    data = bray_delta_non_original,
    aes(group = simulation),
    alpha = 0.18,
    fill = "grey92",
    color = "grey35",
    outlier.shape = NA,
    width = 0.55
  ) +
  geom_jitter(
    data = bray_delta_non_original,
    color = "grey35",
    width = 0.10,
    height = 0,
    alpha = 0.45,
    size = 0.8
  ) +
  geom_point(
    data = bray_delta_original,
    color = "#D55E00",
    size = 1.6
  ) +
  labs(
    title = bray_method_label,
    x = "Case",
    y = "PERMDISP delta"
  ) +
  theme_patchwork_compact +
  theme(
    legend.position = "bottom",
    axis.title.y = element_text(size = 7)
  )

bray_pvalue_plot <- pvalue_plot_data |>
  dplyr::filter(.data$method == bray_method_label) |>
  dplyr::mutate(
    simulation = factor(
      as.character(.data$simulation),
      levels = c("No DA, No DD", "No DA, Yes DD", "Original")
    )
  )
bray_pvalue_non_original <- bray_pvalue_plot |>
  dplyr::filter(.data$simulation != "Original")
bray_pvalue_original <- bray_pvalue_plot |>
  dplyr::filter(.data$simulation == "Original")

bray_pvalue_plot <- bray_pvalue_plot |>
  ggplot(aes(x = simulation, y = p_value)) +
  geom_hline(yintercept = 0.05, linetype = "dotted", color = "grey45") +
  geom_boxplot(
    data = bray_pvalue_non_original,
    aes(group = simulation),
    alpha = 0.18,
    fill = "grey92",
    color = "grey35",
    outlier.shape = NA,
    width = 0.55
  ) +
  geom_jitter(
    data = bray_pvalue_non_original,
    color = "grey35",
    width = 0.10,
    height = 0,
    alpha = 0.45,
    size = 0.8
  ) +
  geom_point(
    data = bray_pvalue_original,
    color = "#D55E00",
    size = 1.6
  ) +
  tidybulk::scale_y_log10_reverse() +
  labs(
    title = bray_method_label,
    x = "Case",
    y = "PERMDISP p-value"
  ) +
  theme_patchwork_compact +
  theme(
    legend.position = "bottom",
    axis.title.y = element_text(size = 7)
  )

pcoa_pair_compact_plot <- pcoa_pair_plot &
  theme_patchwork_compact &
  theme(
    legend.position = "bottom",
    axis.title.y = element_text(size = 7),
    plot.margin = margin(0, 2, 0, 0, "pt")
  )

top_row_plot <- patchwork::wrap_plots(
  patchwork::plot_spacer(),
  sccomp_compact_plot,
  ncol = 2,
  widths = c(1.15, 1)
)
bottom_right_plot <- bray_delta_plot / bray_pvalue_plot +
  patchwork::plot_layout(heights = c(1, 1))
bottom_row_plot <- patchwork::wrap_plots(
  pcoa_pair_compact_plot,
  bottom_right_plot,
  ncol = 2,
  widths = c(1.15, 1)
)

final_patchwork_plot <- top_row_plot / bottom_row_plot +
  patchwork::plot_layout(heights = c(1.08, 1), guides = "collect") &
  theme_patchwork_compact &
  theme(
    plot.margin = margin(0, 0, 0, 0, "pt"),
    legend.position = "bottom",
    legend.key.size = grid::unit(0.2, "cm")
  )

final_patchwork_plot

target_width_mm <- 183
target_height_mm <- target_width_mm / (1024 / 596) * 1.2

ggplot2::ggsave(
  filename = "final_patchwork_summary.pdf",
  plot = final_patchwork_plot,
  width = target_width_mm,
  height = target_height_mm,
  units = "mm",
  device = grDevices::cairo_pdf,
  dpi = 320
)
```
